### 2016-03-30 manuscript.tex
first commit

\documentclass[12pt]{article}

%==============Packages & Commands==============
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{tikz}
%%%<
\usepackage{verbatim}
%\usepackage[active,tightpage]{preview}
%\PreviewEnvironment{tikzpicture}
%\setlength\PreviewBorder{5pt}%

\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                   		% ... or a4paper or a5paper or ...
%\geometry{landscape}                		% Activat\usetikzlibrary{arrows}e for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex
\usepackage{amssymb}

\usepackage[ruled,vlined]{algorithm2e}
\usetikzlibrary{arrows}
\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{indentfirst}
\usepackage[longnamesfirst]{natbib} % For references
\bibpunct{(}{)}{;}{a}{}{,} % Reference punctuation
\usepackage{changepage}
\usepackage{setspace}
\usepackage{booktabs} % For tables
\usepackage{rotating} % For sideways tables/figures
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{color}
\usepackage{dcolumn}
\usepackage{comment}
%\usepackage{fullwidth}
\newcolumntype{d}[1]{D{.}{\cdot}{#1}}
\newcolumntype{.}{D{.}{.}{-1}}
\newcolumntype{3}{D{.}{.}{3}}
\newcolumntype{4}{D{.}{.}{4}}
\newcolumntype{5}{D{.}{.}{5}}
\usepackage{float}
\usepackage[hyphens]{url}
%\usepackage[margin = 1.25in]{geometry}
%\usepackage[nolists,figuresfirst]{endfloat} % Figures and tables at the end
\usepackage{subfig}
\captionsetup[subfloat]{position = top, font = normalsize} % For sub-figure captions
\usepackage{fancyhdr}
%\makeatletter
%\def\url@leostyle{%
%  \@ifundefined{selectfont}{\def\UrlFont{\sf}}{\def\UrlFont{\small\ttfamily}}}
%\makeatother
%% Now actually use the newly defined style.
\urlstyle{same}
\usepackage{times}
% \usepackage{mathptmx}
%\usepackage[colorlinks = true,
%						bookmarksopen = true,
%						pagebackref = true,
%						linkcolor = black,
%						citecolor = black,
% 					urlcolor = black]{hyperref}
%\usepackage[all]{hypcap}
%\urlstyle{same}
\newcommand{\fnote}[1]{\footnote{\normalsize{#1}}} % 12 pt, double spaced footnotes
\def\citeapos#1{\citeauthor{#1}'s (\citeyear{#1})}
\def\citeaposs#1{\citeauthor{#1}' (\citeyear{#1})}
\newcommand{\bm}[1]{\boldsymbol{#1}} %makes bold math symbols easier
\newcommand{\R}{\textsf{R}\space} %R in textsf font
\newcommand{\netinf}{\texttt{NetInf}\space} %R in textsf font
\newcommand{\iid}{i.i.d} %shorthand for iid
\newcommand{\cites}{{\bf \textcolor{red}{CITES}}} %shorthand for iid
%\usepackage[compact]{titlesec}
%\titlespacing{\section}{0pt}{*0}{*0}
%\titlespacing{\subsection}{0pt}{*0}{*0}
%\titlespacing{\subsubsection}{0pt}{*0}{*0}
%\setlength{\parskip}{0pt}
%\setlength{\parsep}{0pt}
%\setlength{\bibsep}{2pt}
%\renewcommand{\headrulewidth}{0pt}

%\renewcommand{\figureplace}{ % This places [Insert Table X here] and [Insert Figure Y here] in the text
%\begin{center}
%[Insert \figurename~\thepostfig\ here]
%\end{center}}
%\renewcommand{\tableplace}{%
%\begin{center}
%[Insert \tablename~\theposttbl\ here]
%\end{center}}

\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\Y}{\bm{\mathcal{Y}}}
\newcommand{\bZ}{\bm{Z}}

\usepackage[colorlinks = TRUE, urlcolor = black, linkcolor = black, citecolor = black, pdfstartview = FitV]{hyperref}

%============Article Title, Authors==================
\title{\vspace{-2cm} Inference on the Effects of Observed Features 
\\ in Latent Space Models for Networks }

\author{ Zachary Jones \and Matthew Denny \and Bruce Desmarais \and Hanna Wallach} \date{\today}

%===================Startup=======================
\begin{document}
\maketitle

%=============Abstract & Keywords==================

\begin{abstract}

\noindent The latent space model (LSM) for network data is a popular model that combine a generalized linear model with a latent spatial embedding of the network to model ties. It has been used to decrease error in the estimation of and inference regarding the effect of observed covariates. In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounders that have an effect on the value of edges in a network. We investigate the latent space model's performance on this task via a Monte Carlo simulation wherein we show that in the presence of an unmeasured covariate which can be appropriately modeled using a latent space, estimation and inferential error does decrease. However, when no such unmeasured covariate is present, a generalized linear model substantially outperforms the latent space model. The LSM is often biased downwards by nearly half the magnitude of the coefficient of the observed covariate. This pattern is repeated with generalization error and with type 1 and type 2 error rates. We investigate solutions to this issue and conclude with recommendations as to the appropriate use of the latent space model for observational studies.

\end{abstract}
\thispagestyle{empty}
% \doublespacing
% Description of the possible challenges
\section{Introduction}

\subsection{Problem}

Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models. The latent space model (LSM) \cite{hoff2002latent} for network data is regularly used to estimate the effect of covariates in the presence of latent network structure. However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured are correlated with measured covariates. There are two reasons that using the LSM may lead to increased error. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spurriously correlated with the observed covariates, leading to reduced efficiency due to multicollinearity. Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable, while the latent space parameters are used to model other sources of variation.

In what follows we identify the conditions under which using the LSM reduces the estimation and inferential error with respect to the effects of observed covariates. We compare the performance of LSM to the generalized linear model (GLM) as a baseline. We develop a simulation design in which we control the degree of collinearity between omitted network structure and observed covariates. We investigate whether the prior can be tuned to avoid efficiency losses in using the LSM. We conclude with recommendations regarding the ideal use of the LSM.

\section{Background}

\subsection{Applications and Development of the Latent Space Model}

The LSM has seen use in a variety of fields in which network data is common, particularly the social sciences. This appeal appears to have been driven primarily by the LSM's usefulness in modeling transitivity and homophily which appear to be ubiquitous in many (social) networks. In political science the LSM and variations on the form developed in \cite{hoff2002latent} have been used to estimate the effect of democracy on the probability of a militarized interstate dispute \citep{ward2007disputes}, the amount of portfolio investment between states \citep{cao2013democracies}, and the effect of multimember districts on the probability of collaboration between state legislators in the United States \citep{kirkland2012multimember}. Variations of the LSM developed for networks measured over time have been applied to the study of international trade, wherein the effects of various features of trading partners are estimated \citep{ward2013gravity}. In ecology the LSM has been used to study the sociality of elephants \citep{vance2009social} and orcas \citep{fearnbach2014spatial}, to discover ecological communities \citep{fletcher2011social, fletcher2013network}, and to study food webs \citep{chiu2011unifying}. In epidemeology the LSM has been used to identify clusters of infected persons for later isolation \citep{zhang2015cluster} and to study patterns of interaction amongst physicians \citep{paul2014results}. In marketing and business research the LSM has been used to study inter-group trust \citep{dass2011impact}, optimal bundling and pricing of goods and brands for retailers \citep{dass2012assessing}. In neuroscience the LSM has been proposed as a method for modelling fMRI data \citep{simpson2013analyzing}.

% need to add marketing stuff and probably should look again for applications
% also need to read and fit in the newer ward student stuff

Although the latent space model is arguably most useful as an exploratory or predictive model \citep{hoff2002latent, shmueli2010explain} it has been applied in some cases in a manner that suggests that users of the model assume that the latent space parameters reduce estimation and/or inferential error with respect to the effects of measured explanatory variables. Though it is not the case that all of the aforementioned articles explicitly or implicitly make such a claim (not all of the above applications of the LSM even include covariates), if the LSM does reduce inferential/estimation error under certain conditions, the LSM might find wider use due to the pervasiveness of network data where there remains unmeasured structure which either confounds relationships between measured covariates and edges, in which case the LSM might be applied to reuduce bias, or in which unmeasured structure is related to the strength/presence of edges and can be useful in decreasing estimator variance. Additionally, the predictive performance of the LSM has only seen limited evaluation \citep{hoff2002latent}, despite having been used for model selection \citep{ward2013gravity, fletcher2011social, fletcher2013network, chiu2011unifying}. Hence, finding the condtions under which the LSM reduces prediction error also may affect future use of the LSM.

The LSM has also seen a substantial amount of subsequent development and extension. The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors \citep{hoff2002latent, hoff2009multiplicative}. Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}. Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process. \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models. \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, 
but not vice-versa.

\subsection{Collinearity in Generalized Linear Models}

\citep{mackinnon1989collinearity} and \citep{weissfeld1991multicollinearity} investigate multicollinearity in generalized linear models and propose diagnostic criteria. Collinearity of the covariates causes the observed information matrix to be ill-conditioned, which produces nmerical instability in the iterative esimation of $\hat{\mathbf{\beta}}$. \citep{weissfeld1991multicollinearity}, extending the work of \citep{davis1986example}, propose the use of the condition index and the variance decomposition proportion of each condition index, to diagnose the presence and degree of multicollinearity in generalized linear models. The observed information matrix at the maximum likelihood estimate, $\mathcal{I}(\hat{\mathbf{\beta}})$ is scaled to have unit length and singular value decomposition is applied. The resulting right singular values are the eigenvectors and the associated singular values are the square root of the eigenvalues of the scaled observed information matrix since it is real Hermitian. The condition index is the ratio of the largest eigenvalue to each of the eigenvalues, and is at least 1. The square root of the condition number gives the potential magnification of the numerical error in inverting the observed information matrix. Hence our expectation is that as collinearity increases the variance of estimators of the effect of observed covariates will increase regardless of whether this is ommitted confounding structure in the data.

\section{Research Design}

\subsection{Priors for the Latent Space Model}

\subsubsection{Current Implementations}

What priors are currently used and why? \cite{hoff2002latent} Propose the use of diffuse independent normal priors for the latent positions and regression parameters. The also recognize the fact that isolates' positions are weakly identified in the LSM, and in one application they actually just exclude the isolate.\footnote{Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space. However, such deletion would introduce bias in estimating covariate effects.} \cite{hoff2002latent} experiment with the use of an exponential prior for the intercept, which they argue sets a lower bound on the probability of a tie between nodes. Of course, the latent space positions could simply drift further apart in order to compensate for the high intercept. The following paragraph on priors appears on page 1097.

\begin{quote}
We have not discussed in detail the choice of a prior distribution for latent positions in this article. Although simple, the diffuse independent normal priors presented in the examples may not accurately represent prior beliefs about the structure of social networks. More appropriate might be clustered point processes or mixtures of normals with an unknown number of components. Such priors could allow one to incorporate prior information on tendencies for clustering, without specifying cluster membership. This would add another level of hierarchy to the analysis, although the resulting model would be more flexible and perhaps more accurately represent any tendencies of populations to form segregating groups.
\end{quote}

\citep{Krivitsky2009} implement a hierarchical model in which each node's latent space position is drawn from a mixture of $G$ normals. The covariate priors are single Gaussians.

\subsubsection{Alternatives to consider}

Would regularization/sparsity priors help with identifying the covariate effects? \cite{mitchell1988} propose a class of ``spike and slab'' prior as a class of priors for the coefficients in linear regression.  These priors are well-suited to the problem of variable selection for linear regression. The prior is formed as a mixture between a diffuse uniform prior and a point mass at zero. Such a prior would not be appropriate for latent space coordinates since weakly identified coordinates of isolates or separate components would likely exhibit strange behavior at the boundary of the diffuse uniform. We might consider instead mixing, e.g., a point mass at zero with a distribution such as the Cauchy, which is fairly flat in the tails, but is defined on the entire real line and still exhibits a surface that descends from the center of the distribution. This prior structure would build a preference into the model for using the observed covariates, rather than the latent coordinate parameters, to explain tie formation.

What priors could be used to avoid collinearity between observed covariates and latent space? Look to Bayesian Latent Trait models and specifically differential item functioning (DIF) in the IRT literature.

Should priors be calibrated against observed covariates? Using informative priors assures that the latent space coordinates are weakly identified, meaning that at some distance from the origin the increase in the likelihood from moving isolates away from the other nodes, or components away from each other, is offset by decreases in the prior from moving coordinates away from the origin. However, we may also use the priors to penalize the model for inferring latent coordinates that result in distances that are comparable in scale to the observed covariates. This would embed a tendency for the model to use the observed covariates, rather than the latent space, to model the outcome. Consider the following example of a one-dimensional spatial model with an additional dyadic covariate.

$$ pr(y_{ij} = 1) = \text{logit}^{-1}(\beta_0 + \beta_1x_{ij} -  |z_i - z_j|).$$

If we assume that the $z$ are independent and distributed $N(0,\sigma^2)$, it is straightforward to derive the distribution of $ |z_i - z_j|$.\footnote{The assumption that $z \sim N(0,\sigma^2$) is consistent with $z$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim N(0,2\sigma^2)$ \citep{devore2012}. The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}. Following from this result, we know that the variance of  $|z_i - z_j|$ is $$\sigma^2(2-4/\pi). $$ Let $\alpha^2$ be the empirical variance of $x$. Then the variance in the linear predictor is $\beta_1^2\alpha^2$. Given an estimate $\hat{\beta_1}$, we can set the variance of the latent space prior to approximately $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$ in order to tune the prior to avoid replicating $x$ via the distances between latent coordinates. This result is limited in that it requires the assumption of normally distributed latent coordinates and applies only to a single latent dimension. However, it may provide a reasonable approximation in the case of non-normal coordinates and/or multiple dimensions.

\cite{thirey2015} derive the distribution of Euclidean Distances between two $k$-dimensional points where each coordinate is drawn from a standard normal distribution.

\subsection{Tasks}
\begin{itemize}
\item Review multicollinearity issues in GLM (ZJ)
\item Describe how multicollinearity will manifest in LSM (ZJ)
\item Delve into LSM priors (BD)
\begin{itemize}
\item Scaling variance to observed covariates
\item Spike and slab
\item Covariance
\end{itemize}
\end{itemize}

\section{Analysis}

\subsection{simulation design}

To study how the LSM behaves in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space. For simplicity we consider a unidimensional latent space. To generate an observed covariate which has a controllable collinearity with the latent network structure we follow a three-step process. First, we simulate $k$-dimensional positions for each node, drawing each from a normal distribution, and calculate dyadic distances $d$ between each pair of positions. Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle X,d \rangle$, we derive the conditional mean vector and covariance, assuming that $X$ has a joint normal distribution given $d$ (see \cite[pp. 116--117]{eaton1983} for the conditional normal derivation). Third, we simulate $X$ as a multivariate normal random variable with the respective conditional means and covariance. Finally we standardize each $X$ to have 0 mean and unit variance. To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $X$ we utilize the C-vine method of \cite{lewandowski2009generating}. % fill in a basic description of the method here

We consider three exponential family distributions for the edges: Gaussian, Binomial, and Poisson. Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space $k$ at 1 for simplicity, and consider 5 values of the collinearity parameter $\eta$, equally spaced between $.75$ and $100$, where $.75$ corresponds to near perfect collinearity and $100$ to independence.

We standardize $X$ to have a variance of $1$ and a mean of $0$, and the linear predictor, $X \beta - d$

\begin{figure}
\includegraphics[width=\textwidth]{max_r_vine.png}
\caption{The distribution of the absolute value of the maximum off-diagonal correlation.}
\end{figure}

Each step in the simulation involves the estimation of a GLM, the LSM, and the LSM with a prior on the latent space which is scaled by $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$, where $\alpha^2$ is the empirical variance of $X$. At each step in the simulation, the LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}. An initial run of 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler. Every 100th iteration is saved. Convergence in the log probability of the model is assessed using the Geweke diagnostic. If the convergence criteria is satisfied the simulation continues to the next set of arguments, otherwise the number of iterations is doubled. If the convergence criteria is still not satisfied, then the aforementioned step in the simulation is flagged for review. At each point in the simulation's parameter space, we execute 500 Monte Carlo iterations.

We first compute the MLE estimated by iteratively reweighted least squares in \texttt{R}. For the LSM we use the posterior mode. In the cases where edges are Binomial or Poisson distributed, we scale the estiamtes % figure out how to do this.

To evaluate estimation error we compute the bias: the difference between the true coefficient $\beta_1$ and the estimate. Figure \ref{fig:estimation} shows the results. When a latent space is present (i.e., $\beta_2 = -1$) the LSM outperforms the GLM in terms of bias across the board. However, when latent network structure is not present ($\beta_2 = 0$) the LSM is sharply biased downwards, indicating that the latent space is representing variability in the edges that is attributable to the observed covariate $X$.

\begin{figure}
\includegraphics[width=\textwidth]{estimation.png}
\caption{The bias of estimates of the effect of the observed covariate $X$. The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The $y$-axis gives a Monte Carlo estimate of the bias, and the grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate. The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels. Each panel represents 3 values of $\eta$ with 500 Monte Carlo iterations executed at each point.
\label{fig:estimation}}
\end{figure}

We consider networks of varying size (i.e. number of nodes), density (i.e., the number/strength of edges between nodes), and collinearity structure between the ommitted distance between nodes and the measured covariates. For each point in this space we evaluate the mean square prediction error of the model on new data drawn from the same distribution (i.e., the generalization error), the bias of the estimated coefficients for the measured covariate, the type 2 error rate (i.e., the probability that the null hypothesis that each regression parameter is $0$ is incorrectly accepted), and the type 1 error rate, that is, the probability of rejecting the null hypothesis that the effect of the measured covariate is 0 when it is in fact 0. Inference regarding $\boldsymbol\beta$ is computed using the maximum a posteriori estimate (i.e., the mode of the posterior), and credible sets for $\boldsymbol\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution for each regression parameter.

\begin{figure}
\includegraphics[width=\textwidth]{inference.png}
\caption{Monte Carlo estimates of the Type-1 and Type-2 error regarding the effect of the observed covariate $X$ are shown on the $y-axis$. When the right gray panel indicates ``Type-1'' $\beta_1 = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta_1 = 0$ is falsely rejected. When the right gray panel indicates ``Type-2'' then $\beta_1 = 1$ and the error rate shown in each panel in that row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis. The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate. The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels. Each panel represents 5 values of $\eta$ with 500 Monte Carlo iterations executed at each point. \label{fig:inference}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{generalization.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the generative model using a model trained on independent data. The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate. The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels. Each panel represents 5 values of $\eta$ with 500 Monte Carlo iterations executed at each point. \label{fig:generalization}}
\end{figure}

One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM. We utilize the \texttt{BatchJobs} and \texttt{BatchExperiments} \R packages to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.

\newpage

\bibliographystyle{apsr}
\bibliography{ref}

\end{document}

### 2016-03-30 .BatchJobs.R

@@ -1,0 +1,5 @@
+## cluster.functions = makeClusterFunctionsTorque("template.tmpl")
+## max.concurrent.jobs = 80L
+## debug = TRUE
+cluster.functions = makeClusterFunctionsMulticore()
+## cluster.functions = makeClusterFunctionsInteractive(write.logs = TRUE)

### 2016-03-30 analysis.R

@@ -1,0 +1,34 @@
+pkgs <- c("dplyr", "reshape2", "ggplot2", "stringr", "coda")
+invisible(lapply(pkgs, library, character.only = TRUE))
+
+results <- read.csv("results.csv", stringsAsFactors = FALSE) %>%
+  select(-one_of("X", "id", "prob", "repl")) %>%
+  rename(model = algo)
+
+results$model <- ifelse(!is.na(results$scale) & results$scale, "lsm_scaled", results$model)
+
+results$bias <- results$estimate - results$beta
+results$error <- ifelse(results$beta == 0, 1 - results$coverage, results$coverage)
+results$type <- ifelse(results$beta == 0, 1, 2)
+
+ggplot(results[results$beta == 1, ], aes(eta, loss, color = model, linetype = model)) +
+  scale_x_log10() +
+  stat_smooth(method = "loess", se = TRUE) +
+  facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
+  labs(x = expression(eta), y = "Generalization Error (MSE)")
+ggsave("generalization.png", width = 10, height = 8)
+
+ggplot(results[results$beta == 1, ], aes(eta, bias, color = model, linetype = model)) +
+  scale_x_log10() +
+  stat_smooth(method = "loess", se = TRUE) +
+  facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
+  geom_hline(aes(yintercept = 0), linetype = "dashed") +
+  labs(x = expression(eta), y = "Bias")
+ggsave("estimation.png", width = 10, height = 8)
+
+ggplot(results, aes(eta, error, color = model, linetype = model)) +
+  stat_smooth(method = "loess", se = TRUE) +
+  facet_grid(family + type ~ nodes + latent_space, scales = "free", labeller = label_context) +
+  labs(x = expression(eta), y = expression(paste("Error rate at ", alpha, " = .05"))) +
+  scale_x_log10()
+ggsave("inference.png", width = 10, height = 8)

### 2016-03-30 glm_sim.R

@@ -1,0 +1,95 @@
+set.seed(1987)
+
+library(clusterGeneration)
+library(far)
+library(foreach)
+library(iterators)
+library(doParallel)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(ggplot2)
+
+cl <- makePSOCKcluster(6)
+registerDoParallel(cl)
+
+loss <- mse <- function(a, b) mean((a - b)^2)
+
+diagnostic <- function(fit) {
+  stopifnot(any(class(fit) %in% "glm"))
+  SVD <- svd(scale(solve(vcov(fit)), center = FALSE))
+  max_d <- max(SVD$d)
+  CI <- sapply(SVD$d, function(x) max_d / x)
+  phi <- t((SVD$v %*% diag(1 / SVD$d))^2)
+  pi <- prop.table(phi, 2)
+  out <- data.frame(CI, pi)
+  colnames(out) <- c("ci", names(coef(fit)))
+  round(out, 3)
+}
+
+families <- c("gaussian", "binomial", "poisson")
+M <- 5
+mc <- 100
+n <- 100
+p <- 2
+sigma <- 1
+res_eta <- 20
+res_beta <- 20
+r <- sort(runif(res_eta, 1, M))
+beta <- lapply(seq_len(res_beta), function(x) runif(p, -1, 1))
+
+sim <- foreach(f = families) %:% foreach(b = beta) %:% foreach(r = r, .packages = "clusterGeneration") %:%
+  foreach(icount(mc)) %dopar% {
+    m = cov2cor(genPositiveDefMat(p, "c-vine", eta = r)$Sigma)
+    X <- t(t(chol(m)) %*% matrix(rnorm(n * p), p, n))
+    lp <- X %*% b
+    y <- switch(f,
+                gaussian = rnorm(n, lp, sigma),
+                binomial = rbinom(n, 1, 1 / (1 + exp(-(lp)))),
+                poisson = rpois(n, exp(lp)))
+    fit <- glm(y ~ -1 + X, family = f)
+    list("beta" = coef(fit), "se" = sqrt(diag(vcov(fit))), "d" = diagnostic(fit), "max_r" = max(cor(X) - diag(p)),
+         "mse" = mse(b, coef(fit)))
+  }
+
+est_r <- sapply(sim, function(f) sapply(f, function(b) sapply(b, function(m) do.call(c, lapply(m, function(x) x$max_r)))))
[…]

### 2016-03-30 simulation.R

@@ -1,0 +1,238 @@
+library("BatchExperiments")
+library("latentnet")
+
+create_network <- function(y, X, name = "eweight") {
+  library("network")
+  
+  net <- network(y, directed = FALSE)
+  el <- as.matrix(net, "edgelist")
+  w <- y[el]
+  set.edge.attribute(net, name, w)
+  for (i in 1:length(X))
+    net %e% names(X) <- X[[i]]
+  net
+}
+
+rmvnorm_d <- function(sigma, d) {
+  library("mvtnorm")
+  
+  var_d <- var(d)
+  mu_d <- mean(d)
+  n_var <- ncol(sigma)
+  d_inv <- solve(as.matrix(var_d))
+  sigma_x <- sigma[1:(n_var - 1), 1:(n_var - 1)] - sigma[1:(n_var - 1), n_var] %*%
+    d_inv %*% sigma[n_var, 1:(n_var - 1)]
+  mu_x <- sapply(d, function(z) sigma[1:(n_var - 1), n_var] %*% d_inv %*% z - mu_d)
+  if (!is.matrix(mu_x))
+    as.matrix(sapply(mu_x, rnorm, n = 1, sd = sqrt(sigma_x)))
+  else
+    t(apply(mu_x, 2, function(mu) rmvnorm(1, mu, sigma_x)))
+}
+
+ms_error <- function(x, y, each_column = FALSE) {
+  if (each_column)
+    sapply((x - y)^2, mean)
+  else
+    mean((x - y)^2)
+}
+
+sign_error <- function(x, y, each_column = TRUE) {
+  if (each_column)
+    sapply(sign(x) != sign(y), mean)
+  else
+    mean(sign(x) != sign(y))
+}
+
+unstack_vector <- function(x, n) {
+  idx <- t(combn(1:n, 2))
+  mat <- diag(n)
+  mat[idx] <- x
+  mat[idx[, c(2, 1)]] <- x
+  stopifnot(isSymmetric(mat))
+  mat
+}
+
+diagnostic <- function(fit) {
+  stopifnot(any(class(fit) %in% "glm"))
+  SVD <- svd(scale(solve(vcov(fit)), center = FALSE))
+  max_d <- max(SVD$d)
+  CI <- sapply(SVD$d, function(x) max_d / x)
+  phi <- t((SVD$v %*% diag(1 / SVD$d))^2)
+  pi <- prop.table(phi, 2)
+  out <- data.frame(CI, pi)
[…]

### 2016-03-31 simulation.R
fixes environments issue and adds prior variance arg

@@ -3,13 +3,11 @@ library("latentnet")
 
 create_network <- function(y, X, name = "eweight") {
   library("network")
-  
   net <- network(y, directed = FALSE)
   el <- as.matrix(net, "edgelist")
   w <- y[el]
   set.edge.attribute(net, name, w)
-  for (i in 1:length(X))
-    net %e% names(X) <- X[[i]]
+  net %e% "X" <- X
   net
 }
 
@@ -34,13 +32,6 @@ ms_error <- function(x, y, each_column = FALSE) {
     sapply((x - y)^2, mean)
   else
     mean((x - y)^2)
-}
-
-sign_error <- function(x, y, each_column = TRUE) {
-  if (each_column)
-    sapply(sign(x) != sign(y), mean)
-  else
-    mean(sign(x) != sign(y))
 }
 
 unstack_vector <- function(x, n) {
@@ -80,7 +71,7 @@ create_data <- function(static, family, eta, nodes, beta, latent_space) {
                     poisson = rpois(length(lp), exp(lp)))
   y <- unstack_vector(y_stack, nodes)
   X_stack <- data.frame(X)
-  X <- lapply(as.list(X_stack), function(x) unstack_vector(x, nodes))
+  X <- unstack_vector(X, nodes)
   y_stack_new <- switch(family,
                         gaussian = rnorm(length(lp), lp, static$sigma),
                         binomial = rbinom(length(lp), 1, 1 / (1 + exp(-(lp)))),
@@ -108,24 +99,28 @@ lsm.wrapper <- function(static, dynamic, ...) {
   library("coda")
 
   dots <- list(...)
-  
+
   net <- create_network(dynamic$y, dynamic$X)
+  .GlobalEnv$net <- net ## fuck you latentet
+
   lf <- c("gaussian" = "Gaussian", "binomial" = "Bernoulli", "poisson" = "Poisson")
 
   if (dots$scale) {
     glm_fit <- glm(y ~ X, dynamic$family, data.frame(dynamic$X_stack, "y" = dynamic$y_stack))
     prior_variance <- (abs(coef(glm_fit)[2]) * var(dynamic$X_stack)) / (2 - 4 / pi)
     prior_variance <- unname(as.numeric(prior_variance))
-  } 
+    prior <- ergmm.prior("Z.var" = prior_variance, ...)
+  } else {
+    prior <- ergmm.prior(...)
+  }
 
[…]

### 2016-04-06 .BatchJobs.R
modifies viz code for addition of beta.var param

@@ -1,5 +1,5 @@
-## cluster.functions = makeClusterFunctionsTorque("template.tmpl")
-## max.concurrent.jobs = 80L
+cluster.functions = makeClusterFunctionsTorque("template.tmpl")
+max.concurrent.jobs = 80L
 ## debug = TRUE
-cluster.functions = makeClusterFunctionsMulticore()
+## cluster.functions = makeClusterFunctionsMulticore()
 ## cluster.functions = makeClusterFunctionsInteractive(write.logs = TRUE)

### 2016-04-06 analysis.R

@@ -6,7 +6,8 @@ results <- read.csv("results.csv", stringsAsFactors = FALSE) %>%
   rename(model = algo)
 
 results$model <- ifelse(!is.na(results$scale) & results$scale, "lsm_scaled", results$model)
-
+results$model <- ifelse(!is.na(results$beta.var), paste0(results$model, " (beta.var = ", results$beta.var, ")"),
+                        results$model)
 results$bias <- results$estimate - results$beta
 results$error <- ifelse(results$beta == 0, 1 - results$coverage, results$coverage)
 results$type <- ifelse(results$beta == 0, 1, 2)
@@ -27,8 +28,8 @@ ggplot(results[results$beta == 1, ], aes(eta, bias, color = model, linetype = mo
 ggsave("estimation.png", width = 10, height = 8)
 
 ggplot(results, aes(eta, error, color = model, linetype = model)) +
+  scale_x_log10() + 
   stat_smooth(method = "loess", se = TRUE) +
   facet_grid(family + type ~ nodes + latent_space, scales = "free", labeller = label_context) +
-  labs(x = expression(eta), y = expression(paste("Error rate at ", alpha, " = .05"))) +
-  scale_x_log10()
+  labs(x = expression(eta), y = expression(paste("Error rate at ", alpha, " = .05")))
 ggsave("inference.png", width = 10, height = 8)

### 2016-04-06 simulation.R

@@ -109,7 +109,7 @@ lsm.wrapper <- function(static, dynamic, ...) {
     glm_fit <- glm(y ~ X, dynamic$family, data.frame(dynamic$X_stack, "y" = dynamic$y_stack))
     prior_variance <- (abs(coef(glm_fit)[2]) * var(dynamic$X_stack)) / (2 - 4 / pi)
     prior_variance <- unname(as.numeric(prior_variance))
-    prior <- ergmm.prior("Z.var" = prior_variance, ...)
+    prior <- ergmm.prior("Z.var" = prior_variance)
   } else {
     prior <- ergmm.prior(...)
   }
@@ -205,7 +205,7 @@ while(!done) {
   ## find how many chunked jobs are running
   r <- as.integer(system("qstat -u zmj102 | wc -l", intern = TRUE))
   submitJobs(reg, chunk(ids, chunk.size = 10)[seq_len(max(80 - r, 0))], resources)
-  Sys.sleep(300)
+  Sys.sleep(60)
 
   if (getJobNr(reg) == length(findDone(reg)))
     done <- TRUE

### 2016-04-06 manuscript.tex
add text on scaling binomial estimates for glm

@@
 
-We standardize $X$ to have a variance of $1$ and a mean of $0$, and the linear predictor, $X \beta - d$
+We standardize $X$ to have a variance of $1$ and a mean of $0$.
 
@@
 For the LSM we use the posterior mode.
-In the cases where edges are Binomial or Poisson distributed, we scale the estiamtes % figure out how to do this.
+In the cases where edges are Binomial distributed, we scale the estiamtes using the reciprocal of the bias of $\beta_1$ when $X$ and $d$ are independent: $\\frac{\sqrt{3.28 + \beta_2^2 \text{Var(d)}}}{\sqrt{3.29}}$, where $3.29$ is the variance of a standard logistic distribution \cite{mood2010logistic}.
 

### 2016-04-14 manuscript.tex
remove se in graphs, up mc iter, reduce param for beta.var

--- 2016-03-30 manuscript.tex
@@
 For the LSM we use the posterior mode.
-In the cases where edges are Binomial or Poisson distributed, we scale the estiamtes % figure out how to do this.
+In the cases where edges are Binomial distributed, we scale the estiamtes using the reciprocal of the bias of $\beta_1$ when $X$ and $d$ are independent: $\frac{\sqrt{3.28 + \beta_2^2 \text{Var(d)}}}{\sqrt{3.29}}$, where $3.29$ is the variance of a standard logistic distribution \cite{mood2010logistic}.
 

### 2016-04-14 analysis.R

@@ -14,14 +14,14 @@ results$type <- ifelse(results$beta == 0, 1, 2)
 
 ggplot(results[results$beta == 1, ], aes(eta, loss, color = model, linetype = model)) +
   scale_x_log10() +
-  stat_smooth(method = "loess", se = TRUE) +
+  stat_smooth(method = "loess", se = FALSE) +
   facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = "Generalization Error (MSE)")
 ggsave("generalization.png", width = 10, height = 8)
 
 ggplot(results[results$beta == 1, ], aes(eta, bias, color = model, linetype = model)) +
   scale_x_log10() +
-  stat_smooth(method = "loess", se = TRUE) +
+  stat_smooth(method = "loess", se = FALSE) +
   facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
   geom_hline(aes(yintercept = 0), linetype = "dashed") +
   labs(x = expression(eta), y = "Bias")
@@ -29,7 +29,7 @@ ggsave("estimation.png", width = 10, height = 8)
 
 ggplot(results, aes(eta, error, color = model, linetype = model)) +
   scale_x_log10() + 
-  stat_smooth(method = "loess", se = TRUE) +
+  stat_smooth(method = "loess", se = FALSE) +
   facet_grid(family + type ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = expression(paste("Error rate at ", alpha, " = .05")))
 ggsave("inference.png", width = 10, height = 8)

### 2016-04-14 simulation.R

@@ -180,7 +180,7 @@ problem.pars <- list(
 problem.design <- makeDesign("test", exhaustive = problem.pars)
 
 lsm.pars <- list("scale" = c(TRUE, FALSE),
-                 "beta.var" = c(1, 10, 100))
+                 "beta.var" = c(1, 10))
 glm.pars <- list()
 
 lsm.design <- makeDesign("lsm", exhaustive = lsm.pars)
@@ -190,7 +190,7 @@ addAlgorithm(reg, "lsm", lsm.wrapper, overwrite = TRUE)
 addAlgorithm(reg, "glm", glm.wrapper, overwrite = TRUE)
 
 addExperiments(reg, problem.design, list("lsm" = lsm.design, "glm" = glm.design),
-               repls = 10, skip.defined = TRUE)
+               repls = 100, skip.defined = TRUE)
 batchExport(reg, create_network = create_network, ms_error = ms_error,
             rmvnorm_d = rmvnorm_d, unstack_vector = unstack_vector, diagnostic = diagnostic,
             overwrite = TRUE)

### 2016-04-20 manuscript.tex
updates ahead of lab presentation

--- 2016-03-30 manuscript.tex
@@
 In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounders that have an effect on the value of edges in a network.
-We investigate the latent space model's performance on this task via a Monte Carlo simulation wherein we show that in the presence of an unmeasured covariate which can be appropriately modeled using a latent space, estimation and inferential error does decrease.
-However, when no such unmeasured covariate is present, a generalized linear model substantially outperforms the latent space model.
-The LSM is often biased downwards by nearly half the magnitude of the coefficient of the observed covariate.
-This pattern is repeated with generalization error and with type 1 and type 2 error rates.
+We investigate the latent space model's performance on this task via a Monte Carlo simulation wherein we show that in the presence of an unmeasured covariate which can be appropriately modeled using a latent space, estimation and inferential error decreases, however, the difference is not large, even at high levels of multicollinearity between the unmeasured and observed covariates.
+However, the prediction error of the LSM when an unmeasure covariate is present is substantially lower in most cases.
+When an omitted covariate is not present, the LSM can introduce bias and increased inferential error above that of a generalized linear model.
 We investigate solutions to this issue and conclude with recommendations as to the appropriate use of the latent space model for observational studies.
@@
 
-Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models.
+Latent variable inference, generally conceived, present the possibility of representing unmeasured data in statistical models.
 The latent space model (LSM) \cite{hoff2002latent} for network data is regularly used to estimate the effect of covariates in the presence of latent network structure.
@@
 \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models.
-\cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models,
-but not vice-versa.
+\cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
 
@@
 \citep{mackinnon1989collinearity} and \citep{weissfeld1991multicollinearity} investigate multicollinearity in generalized linear models and propose diagnostic criteria.
-Collinearity of the covariates causes the observed information matrix to be ill-conditioned, which produces nmerical instability in the iterative esimation of $\hat{\mathbf{\beta}}$.
+Collinearity of the covariates causes the observed information matrix to be ill-conditioned, which produces numerical instability in the iterative estimation of $\hat{\mathbf{\beta}}$, the ``effects'' of covariates included in a regression.
 \citep{weissfeld1991multicollinearity}, extending the work of \citep{davis1986example}, propose the use of the condition index and the variance decomposition proportion of each condition index, to diagnose the presence and degree of multicollinearity in generalized linear models.
@@
 However, it may provide a reasonable approximation in the case of non-normal coordinates and/or multiple dimensions.
-
 \cite{thirey2015} derive the distribution of Euclidean Distances between two $k$-dimensional points where each coordinate is drawn from a standard normal distribution.
-
-\subsection{Tasks}
-\begin{itemize}
-\item Review multicollinearity issues in GLM (ZJ)
-\item Describe how multicollinearity will manifest in LSM (ZJ)
-\item Delve into LSM priors (BD)
-\begin{itemize}
-\item Scaling variance to observed covariates
-\item Spike and slab
-\item Covariance
-\end{itemize}
-\end{itemize}
 
@@
 Third, we simulate $X$ as a multivariate normal random variable with the respective conditional means and covariance.
-Finally we standardize each $X$ to have 0 mean and unit variance.
+Finally we standardize each $X$ to have zero mean and unit variance.
 To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $X$ we utilize the C-vine method of \cite{lewandowski2009generating}.
@@
 We consider three exponential family distributions for the edges: Gaussian, Binomial, and Poisson.
-Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space $k$ at 1 for simplicity, and consider 5 values of the collinearity parameter $\eta$, equally spaced between $.75$ and $100$, where $.75$ corresponds to near perfect collinearity and $100$ to independence.
-
-We standardize $X$ to have a variance of $1$ and a mean of $0$, and the linear predictor, $X \beta - d$
+Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space $k$ at 1 for simplicity, and consider 4 values of the collinearity parameter $\eta$, $.1$, $1$, $100$, and $1000000$, where $.1$ a corresponds to near perfect collinearity and $1000000$ to independence.
+See Figure \ref{fig:vine} for the distribution of the absolute value of the maximal off-diagnoal correlation generated at each value of $\eta$.
 
@@
 \includegraphics[width=\textwidth]{max_r_vine.png}
-\caption{The distribution of the absolute value of the maximum off-diagonal correlation.}
+\caption{The distribution of the absolute value of the maximum off-diagonal correlation. \label{fig:vine}}
 \end{figure}
 
-Each step in the simulation involves the estimation of a GLM, the LSM, and the LSM with a prior on the latent space which is scaled by $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$, where $\alpha^2$ is the empirical variance of $X$.
+Each step in the simulation involves the estimation of a GLM, and the LSM.
+We consider the LSM with several different priors on the coefficient for the observed covariate and the latent space.
+We set the prior variance of $\beta_1$ to be either $1$ or $10$ and alternatively use a diffuse normal prior on the latent space or scale it by $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$, where $\alpha^2$ is the empirical variance of $X$.
+
 At each step in the simulation, the LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}.
@@
 
+One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM.
+We utilize the \texttt{BatchJobs} and \texttt{BatchExperiments} \R packages to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.
+
 We first compute the MLE estimated by iteratively reweighted least squares in \texttt{R}.
 For the LSM we use the posterior mode.
-In the cases where edges are Binomial or Poisson distributed, we scale the estiamtes % figure out how to do this.
+In the cases where edges are Binomial distributed, we scale the estiamtes using the reciprocal of the bias of $\beta_1$ when $X$ and $d$ are independent: $\frac{\sqrt{3.28 + \beta_2^2 \text{Var(d)}}}{\sqrt{3.29}}$, where $3.29$ is the variance of a standard logistic distribution (See the derivation of the bias under an independent but ommitted covariante in \cite{mood2010logistic}).
+
+We consider networks of varying size (i.e.
+number of nodes) and collinearity structure between the ommitted distance between nodes and the measured covariates.
+For each point in this space we evaluate the mean square prediction error of the model on new data drawn from the same distribution (i.e., the generalization error), the bias of the estimated coefficients for the measured covariate, the type 2 error rate (i.e., the probability that the null hypothesis that each regression parameter is $0$ is incorrectly accepted), and the type 1 error rate, that is, the probability of rejecting the null hypothesis that the effect of the measured covariate is 0 when it is in fact 0.
+Inference regarding $\beta_1$ is computed using the maximum a posteriori estimate (i.e., the mode of the posterior), and credible intervals for $\beta_1$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution for each regression parameter.
 
 To evaluate estimation error we compute the bias: the difference between the true coefficient $\beta_1$ and the estimate.
-Figure \ref{fig:estimation} shows the results.
-When a latent space is present (i.e., $\beta_2 = -1$) the LSM outperforms the GLM in terms of bias across the board.
-However, when latent network structure is not present ($\beta_2 = 0$) the LSM is sharply biased downwards, indicating that the latent space is representing variability in the edges that is attributable to the observed covariate $X$.
+Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results.
+When a latent space is present (i.e., $\beta_2 = -1$, shown in Figure \ref{fig:estimation_ls}) the LSM outperforms the GLM in terms of bias across the board.
+However, when latent network structure is not present ($\beta_2 = 0$, shown in Figure \ref{fig:estimation_nls}) the LSM is biased upwards in the Binomial case and appears to have higher variance (perhaps due to efficiency losses) for Poisson and Gaussian edges (preliminary interpretation).
 
 \begin{figure}
-\includegraphics[width=\textwidth]{estimation.png}
-\caption{The bias of estimates of the effect of the observed covariate $X$.
-The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate.
+\includegraphics[width=\textwidth]{estimation_ls.png}
+\caption{The bias of estimates of the effect of the observed covariate $X$ when there is an ommitted variable.
+The $x$-axis gives the value of the parameter $\eta$ which controls the degree of dependence between $X$ and ommitted covariate.
 Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate.
-The $y$-axis gives a Monte Carlo estimate of the bias, and the grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate.
-The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels.
-Each panel represents 3 values of $\eta$ with 500 Monte Carlo iterations executed at each point.
-\label{fig:estimation}}
-\end{figure}
-
-We consider networks of varying size (i.e.
-number of nodes), density (i.e., the number/strength of edges between nodes), and collinearity structure between the ommitted distance between nodes and the measured covariates.
-For each point in this space we evaluate the mean square prediction error of the model on new data drawn from the same distribution (i.e., the generalization error), the bias of the estimated coefficients for the measured covariate, the type 2 error rate (i.e., the probability that the null hypothesis that each regression parameter is $0$ is incorrectly accepted), and the type 1 error rate, that is, the probability of rejecting the null hypothesis that the effect of the measured covariate is 0 when it is in fact 0.
-Inference regarding $\boldsymbol\beta$ is computed using the maximum a posteriori estimate (i.e., the mode of the posterior), and credible sets for $\boldsymbol\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution for each regression parameter.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{inference.png}
-\caption{Monte Carlo estimates of the Type-1 and Type-2 error regarding the effect of the observed covariate $X$ are shown on the $y-axis$.
-When the right gray panel indicates ``Type-1'' $\beta_1 = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta_1 = 0$ is falsely rejected.
-When the right gray panel indicates ``Type-2'' then $\beta_1 = 1$ and the error rate shown in each panel in that row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis.
-The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate.
-Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate.
-The grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate.
-The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels.
-Each panel represents 5 values of $\eta$ with 500 Monte Carlo iterations executed at each point.
-\label{fig:inference}}
+The $y$-axis gives a Monte Carlo estimate of the bias.
+The number of nodes are indicated in the top panels, while the distributional family of the edges is shown on the right panel.
+Each panel represents 4 values of $\eta$ with 500 (partially complete) Monte Carlo iterations executed at each point.
+\label{fig:estimation_ls}}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{generalization.png}
-\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the generative model using a model trained on independent data.
-The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate.
-Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate.
-The grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate.
-The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels.
-Each panel represents 5 values of $\eta$ with 500 Monte Carlo iterations executed at each point.
-\label{fig:generalization}}
+\includegraphics[width=\textwidth]{estimation_nls.png}
+\caption{The bias of estimates of the effect of the observed covariate $X$ when there is no ommitted variable.
+\label{fig:estimation_nls}}
 \end{figure}
 
-One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM.
-We utilize the \texttt{BatchJobs} and \texttt{BatchExperiments} \R packages to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.
+Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta_1$ and the latent space compared to a GLM.
+Type 1 error rates for the LSM are in general lower than those of a GLM when an unmeasured covariate exists, however, when one does not exist the GLM outperforms the LSM.
+However, scaling the prior of the latent space to match that of the observed covariates appears to make the LSM's error rate comparable to that of the GLM.
+Type 2 error rates are very similar for the GLM and the LSM with all evaluated prior settings; to the point that they are indistinguishable.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{inference_type_1.png}
+\caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $X$ are shown on the $y$$-axis.
+Here, $\beta_1 = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta_1 = 0$ is falsely rejected.
+\label{fig:inference_type_1}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{inference_type_2.png}
+\caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $X$ are shown on the $y-axis$.
+Here, $\beta_1 = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis.
+\label{fig:inference_type_2}}
+\end{figure}
+
+Generalization error, in this case defined as the expected error on new edges generated using a fixed set of latent distances between the nodes and a node-level covariate is shown in figure \ref{fig:generalization}.
+In nearly all cases the LSM substantially outperforms the GLM, especially when the ommitted covariate is not highly collinear with the observed covariate.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{generalization.png}
+\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the generative model using a model trained on independent data. \label{fig:generalization}}
+\end{figure}
+
+\section{Conclusion}
+
+Although the results are (very) preliminary (and have been plagued by my incompetence!), it seems increasingly clear that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space to uncover patterns of interest or to predict edges in networks where there is likely to be omitted structure that is not adequately modelled using measured covariates.
+However, inferential and estimation error are often smaller than that of a GLM when an omitted variable that corresponds to a form that can be modelled using a latent space is present, and can lead to an up to $25\%$ decrease in bias relative to a GLM.
+Whether or not there is a reduction in estimation and/or inferential error depends on the strength of the ommitted confounder, sample size (the LSM having many more parameters than an equivalent GLM), and relies on the ommitted confounder being representable as a latent space.
+Lastly, there is some (preliminary) evidence to suggest that scaling the prior of the latent space to match that of the measured covariates can reduce Type-1 error.
 

### 2016-04-20 analysis.R

@@ -1,35 +1,52 @@
-pkgs <- c("dplyr", "reshape2", "ggplot2", "stringr", "coda")
+pkgs <- c("dplyr", "reshape2", "ggplot2", "stringr", "coda", "scales")
 invisible(lapply(pkgs, library, character.only = TRUE))
 
 results <- read.csv("results.csv", stringsAsFactors = FALSE) %>%
   select(-one_of("X", "id", "prob", "repl")) %>%
-  rename(model = algo)
-
-results$model <- ifelse(!is.na(results$scale) & results$scale, "lsm_scaled", results$model)
-results$model <- ifelse(!is.na(results$beta.var), paste0(results$model, " (beta.var = ", results$beta.var, ")"),
-                        results$model)
-results$bias <- results$estimate - results$beta
-results$error <- ifelse(results$beta == 0, 1 - results$coverage, results$coverage)
-results$type <- ifelse(results$beta == 0, 1, 2)
+  rename(model = algo) %>%
+  mutate(model = ifelse(!is.na(scale) & scale, "lsm_scaled", model)) %>%
+  mutate(model = ifelse(!is.na(beta.var), paste0(model, " (beta.var = ", beta.var, ")"), model)) %>%
+  mutate(bias = estimate - beta, error = ifelse(beta == 0, 1 - coverage, coverage),
+         type = ifelse(beta == 0, 1, 2)) %>%
+  group_by(model, family, eta, beta, type, latent_space, nodes) %>%
+  summarise(estimate = mean(estimate), loss = mean(loss), coverage = mean(coverage), bias = mean(bias),
+            error = mean(error))
 
 ggplot(results[results$beta == 1, ], aes(eta, loss, color = model, linetype = model)) +
-  scale_x_log10() +
-  stat_smooth(method = "loess", se = FALSE) +
+  scale_x_log10(label = scientific_format()) +
+  geom_point() + geom_line() +
   facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = "Generalization Error (MSE)")
-ggsave("generalization.png", width = 10, height = 8)
+ggsave("generalization.png", width = 12, height = 8)
 
-ggplot(results[results$beta == 1, ], aes(eta, bias, color = model, linetype = model)) +
-  scale_x_log10() +
-  stat_smooth(method = "loess", se = FALSE) +
-  facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
[…]

### 2016-04-20 simulation.R

@@ -1,13 +1,12 @@
 library("BatchExperiments")
 library("latentnet")
 
-create_network <- function(y, X, name = "eweight") {
+create_network <- function(y, name = "eweight") {
   library("network")
   net <- network(y, directed = FALSE)
   el <- as.matrix(net, "edgelist")
   w <- y[el]
   set.edge.attribute(net, name, w)
-  net %e% "X" <- X
   net
 }
 
@@ -100,8 +99,8 @@ lsm.wrapper <- function(static, dynamic, ...) {
 
   dots <- list(...)
 
-  net <- create_network(dynamic$y, dynamic$X)
-  .GlobalEnv$net <- net ## fuck you latentet
+  net <- create_network(dynamic$y)
+  .GlobalEnv$X <- dynamic$X ## fuck you latentet
 
   lf <- c("gaussian" = "Gaussian", "binomial" = "Bernoulli", "poisson" = "Poisson")
 
@@ -120,7 +119,7 @@ lsm.wrapper <- function(static, dynamic, ...) {
   flag <- TRUE
   iter <- 1
   while (flag) {
-    fit <- ergmm(net ~ edgecov(net, "X") + euclidean(d = 1),
+    fit <- ergmm(net ~ edgecov(X) + euclidean(d = 1),
                  "eweight", unname(lf[names(lf) %in% dynamic$family]),
                  fam.par = fam.par, prior = prior,
                  tofit = c("mcmc", "pmode"), seed = static$seed,
@@ -190,25 +189,20 @@ addAlgorithm(reg, "lsm", lsm.wrapper, overwrite = TRUE)
 addAlgorithm(reg, "glm", glm.wrapper, overwrite = TRUE)
 
 addExperiments(reg, problem.design, list("lsm" = lsm.design, "glm" = glm.design),
-               repls = 100, skip.defined = TRUE)
+               repls = 500, skip.defined = TRUE)
 batchExport(reg, create_network = create_network, ms_error = ms_error,
             rmvnorm_d = rmvnorm_d, unstack_vector = unstack_vector, diagnostic = diagnostic,
             overwrite = TRUE)
 
-resources <- list(walltime = 86400L, nodes = 1L, memory = "2gb")
+resources <- list(walltime = 86400L, nodes = 1L, memory = "8gb")
 
-done <- FALSE
-ids <- findNotStarted(reg)
-
-while(!done) {
+while (getJobNr(reg) > length(findDone(reg))) {
[…]

### 2016-04-21 simulation.R
remove old one_iter code

@@ -221,19 +221,6 @@ reduce <- function(job, res) {
   )
 }
 
-## reduce_nothing <- function(job, res) {
-##   list("fit" = res$fit,
-##        "data" = res$data)
-## }
-
-## id <- findExperiments(reg, algo.pattern = "lsm",
-##                       algo.pars = (scale == FALSE & beta.var == 100),
-##                       prob.pars = (nodes == 100 & eta == 1000000 & family == "binomial" &
-##                                      beta == 1 & latent_space == -1),
-##                       repls = 1)
-## one_iter <- reduceResultsList(reg, id, fun = reduce_nothing)[[1]]
-## save(one_iter, file = "one_iter.RData")
-
 done <- summarizeExperiments(reg, findErrors(reg),
                              show = c("algo", "scale", "beta.var", "family", "eta", "nodes", "beta", "latent_space"))
 write.csv(done, "done.csv")

### 2016-04-25 analysis.R
small bugfixes and changes to submit/reduce code

@@ -47,6 +47,6 @@ ggplot(results[results$type == 2, ], aes(eta, error, color = model, linetype = m
   scale_x_log10(label = scientific_format()) + 
   geom_point() + geom_line() +
   facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
-  labs(x = expression(eta), y = expression(paste("Type 1 Error rate at ", alpha, " = .05")),
+  labs(x = expression(eta), y = expression(paste("Type 2 Error rate at ", alpha, " = .05")),
        title = "Type-2 Inferential Error")
 ggsave("inference_type_2.png", width = 12, height = 8)

### 2016-04-25 simulation.R

@@ -194,16 +194,11 @@ batchExport(reg, create_network = create_network, ms_error = ms_error,
             rmvnorm_d = rmvnorm_d, unstack_vector = unstack_vector, diagnostic = diagnostic,
             overwrite = TRUE)
 
-resources <- list(walltime = 86400L, nodes = 1L, memory = "8gb")
+resources <- list(walltime = 86400L * 2, nodes = 1L, memory = "8gb")
 
-while (getJobNr(reg) > length(findDone(reg))) {
-  ids <- findNotStarted(reg)
-  ids <- ids[!ids %in% findOnSystem(reg)]
-  ## find how many chunked jobs are running
-  r <- as.integer(system("qstat -u zmj102 -r | wc -l", intern = TRUE))
-  submitJobs(reg, chunk(ids, chunk.size = 50)[seq_len(max(40 - r, 0))], resources)
-  Sys.sleep(60 * 5)
-}
+ids <- findNotStarted(reg)
+ids <- ids[!ids %in% findOnSystem(reg)]
+submitJobs(reg, chunk(ids, chunk.size = floor(c(100, round(length(ids) * .05)))), resources)
 
 reduce <- function(job, res) {
   if (any(class(res$fit) == "glm") & res$data$latent_space == -1 & res$data$family == "binomial") {
@@ -224,14 +219,8 @@ reduce <- function(job, res) {
 done <- summarizeExperiments(reg, findErrors(reg),
                              show = c("algo", "scale", "beta.var", "family", "eta", "nodes", "beta", "latent_space"))
 write.csv(done, "done.csv")
-
-## necessary for a weird bug
-ids <- chunk(unique(findDone(reg)), 500)
-results <- vector("list", length(ids))
-for (i in 1:length(ids)) {
-  try(results[[i]] <- reduceResultsExperiments(reg, ids[[i]], fun = reduce, progressbar = FALSE))
-}
-results <- do.call("rbind", results)
+res <- reduceResultsExperiments(reg, unique(findDone(reg)),
+                                fun = reduce, impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA))
 write.csv(results, "results.csv")
 
 ## generate a histogram plot for the off diagnoals of the covariance matrix used to generate

### 2016-04-26 .BatchJobs.R
separate out reduce code

@@ -1,5 +1,1 @@
 cluster.functions = makeClusterFunctionsTorque("template.tmpl")
-max.concurrent.jobs = 80L
-## debug = TRUE
-## cluster.functions = makeClusterFunctionsMulticore()
-## cluster.functions = makeClusterFunctionsInteractive(write.logs = TRUE)

### 2016-04-26 reduce.R

@@ -1,0 +1,37 @@
+library(BatchExperiments)
+library(plyr)
+
+reg_open <- loadRegistry("reg")
+reg_bruce <- loadRegistry("../lsm_applications/reg")
+
+reduce <- function(job, res) {
+  if (any(class(res$fit) == "glm") & res$data$latent_space == -1 & res$data$family == "binomial") {
+    estimate <- res$estimate * res$adjustment
+    coverage <- unname(ifelse(res$interval[1] * res$adjustment < 0 & res$interval[2] * res$adjustment > 0, 1, 0))
+  } else {
+    estimate <- res$estimate
+    coverage <- unname(ifelse(res$interval[1] < 0 & res$interval[2] > 0, 1, 0))
+  }
+  
+  list(
+    "estimate" = estimate,
+    "loss" = res$loss,
+    "coverage" = coverage
+  )
+}
+
+
+nodes <- 20L
+res_open <- batchMapQuick(reduceResultsExperiments, chunk(findDone(reg_open), n.chunks = nodes, shuffle = TRUE),
+                          more.args = list(reg = reg_open, fun = reduce, progressbar = FALSE,
+                                           impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
+                          resources = list(walltime = 86400L, nodes = 1L, memory = "8gb"))
+res_open <- reduceResults(res_open, fun = function(aggr, job, res) rbind.fill(aggr, res), init = data.frame())
+
+res_bruce <- batchMapQuick(reduceResultsExperiments, chunk(findDone(reg_bruce), n.chunks = nodes, shuffle = TRUE),
+                          more.args = list(reg = reg_bruce, fun = reduce, progressbar = FALSE,
+                                           impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
+                          resources = list(walltime = 86400L, nodes = 1L, memory = "8gb"))
+res_bruce <- reduceResults(res_bruce, fun = function(aggr, job, res) rbind.fill(aggr, res), init = data.frame())
+
+write.csv(rbind(res_open, res_bruce), "results.csv")

### 2016-04-26 simulation.R

@@ -200,29 +200,6 @@ ids <- findNotStarted(reg)
 ids <- ids[!ids %in% findOnSystem(reg)]
 submitJobs(reg, chunk(ids, chunk.size = floor(c(100, round(length(ids) * .05)))), resources)
 
-reduce <- function(job, res) {
-  if (any(class(res$fit) == "glm") & res$data$latent_space == -1 & res$data$family == "binomial") {
-    estimate <- res$estimate * res$adjustment
-    coverage <- unname(ifelse(res$interval[1] * res$adjustment < 0 & res$interval[2] * res$adjustment > 0, 1, 0))
-  } else {
-    estimate <- res$estimate
-    coverage <- unname(ifelse(res$interval[1] < 0 & res$interval[2] > 0, 1, 0))
-  }
-  
-  list(
-    "estimate" = estimate,
-    "loss" = res$loss,
-    "coverage" = coverage
-  )
-}
-
-done <- summarizeExperiments(reg, findErrors(reg),
-                             show = c("algo", "scale", "beta.var", "family", "eta", "nodes", "beta", "latent_space"))
-write.csv(done, "done.csv")
-res <- reduceResultsExperiments(reg, unique(findDone(reg)),
-                                fun = reduce, impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA))
-write.csv(results, "results.csv")
-
 ## generate a histogram plot for the off diagnoals of the covariance matrix used to generate
 ## d and X_stack
 eta_hist <- foreach(eta = c(.1, 1, 100, 1000000), .combine = "rbind") %:%

### 2016-06-24 manuscript.tex
small bugfixes and changes to submit/reduce code
polnet version

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 
@@
 \usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{a4paper}                   		% ... or a4paper or a5paper or ...
+% \geometry{a4paper}                   		% ... or a4paper or a5paper or ...
 %\geometry{landscape}                		% Activat\usetikzlibrary{arrows}e for for rotated page geometry
@@
 
-\noindent The latent space model (LSM) for network data is a popular model that combine a generalized linear model with a latent spatial embedding of the network to model ties.
-It has been used to decrease error in the estimation of and inference regarding the effect of observed covariates.
-In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounders that have an effect on the value of edges in a network.
-We investigate the latent space model's performance on this task via a Monte Carlo simulation wherein we show that in the presence of an unmeasured covariate which can be appropriately modeled using a latent space, estimation and inferential error decreases, however, the difference is not large, even at high levels of multicollinearity between the unmeasured and observed covariates.
-However, the prediction error of the LSM when an unmeasure covariate is present is substantially lower in most cases.
-When an omitted covariate is not present, the LSM can introduce bias and increased inferential error above that of a generalized linear model.
-We investigate solutions to this issue and conclude with recommendations as to the appropriate use of the latent space model for observational studies.
+\noindent The latent space model (LSM) for network data is a generative probabilistic model that combines a generalized linear model with a latent spatial embedding of the network.
+It has been used to decrease error in the estimation of and inference regarding the effects of observed covariates.
+In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network.
+As far as we know, there has been no research that considers the LSM's performance in adjusting for unmeasured structure to reduce estimation and inferential errors.
+We investigate the LSM's performance via a Monte Carlo study.
+In the presence of an unmeasured covariate that can be appropriately modeled using a latent space, estimation and inferential error remain high under even moderate confounding.
+Additionally, when omitted network structure is not present the LSM induces additional bias and inferential error.
+However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases.
+We conclude that the LSM is most appropriately used for exploratory or predictive tasks.
 
@@
 
-\subsection{Problem}
+Inferential analysis of political network data has grown increasingly sophisticated in recent years.
+Political networks scholars are well versed in the risks associated with ignoring unmodeled network structure.
+Dependencies such as reciprocity, transitivity, and homophily---if not accounted for---can lead to biased estimates and errors in hypothesis testing, much in the way that omitted variable bias can affect results in conventional regression models \citep{ward2007disputes,kinne2014,cranmerisq}.
+A number of statistical modeling frameworks have been proposed to account for confounding structure in network data that cannot be modeled with observed covariates.
+These include the exponential random graph model (ERGM) \citep[e.g., ][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model (LSM) \citep[e.g., ][]{ward2007disputes,ward2007persistent,kirkland2012multimember}, and the stochastic actor oriented model (SAOM) \citep[e.g., ][]{berardo2010ajps,kinne2014}.
 
-Latent variable inference, generally conceived, present the possibility of representing unmeasured data in statistical models.
-The latent space model (LSM) \cite{hoff2002latent} for network data is regularly used to estimate the effect of covariates in the presence of latent network structure.
-However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured are correlated with measured covariates.
+Despite their growing popularity, few studies exist that investigate the performance of these models in adjusting for confounders that can be represented as network structure.
+The approach to adjusting for dependencies in the ERGM and SAOM is quite similar to adjusting for confounding covariates in regression modeling.
+The researcher specifies a set of dependencies that (s)he hypothesizes to be important in the generative model for the network.
+These dependencies are then explicitly included in a model that simultaneously represents the effects of observed covariates \citep{cranmer2011pa}.
+The LSM takes a different approach, which involves the incorporation of latent variables to model network structure.
+The LSM has the advantage over ERGM and SAOM in that researchers need not develop a set of hypothesized dependencies in order to model network structure that is not reflected in observed covariates.
+In the current study, we focus on the LSM, examining its performance in reducing estimation and inferential errors regarding the effects of observed covariates, via adjustment for confounding network structure.
+
+\subsection{Central Problem}
+
+Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models.
+The LSM, introduced by \citet{hoff2002latent}, is used to estimate the effect of covariates in the presence of latent network structure.
+Here the distance function $|z_i - z_j|$ represents latent network structure as homophily with respect to latent variables.
+The distance function is additively combined with a regression on observed dyadic covariates, $x_{ij}$, to form a linear predictor for tie prediction.
+As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution.
+
+$$\mathbb{E}(y_{ij} | x_{ij}) = g^{-1}(\alpha + \beta x_{ij} - |z_i - z_j|)$$
+
+However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured, are correlated with measured covariates.
 There are two reasons that using the LSM may lead to increased error.
-First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spurriously correlated with the observed covariates, leading to reduced efficiency due to multicollinearity.
+First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spurriously correlated with the observed covariates, leading to reduced efficiency.
 Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable, while the latent space parameters are used to model other sources of variation.
 
-In what follows we identify the conditions under which using the LSM reduces the estimation and inferential error with respect to the effects of observed covariates.
-We compare the performance of LSM to the generalized linear model (GLM) as a baseline.
 We develop a simulation design in which we control the degree of collinearity between omitted network structure and observed covariates.
-We investigate whether the prior can be tuned to avoid efficiency losses in using the LSM.
-We conclude with recommendations regarding the ideal use of the LSM.
+We find that the LSM does not substantially reduce bias and inferential error with respect to the effects of observed covariates, in comparison to a generalized linear model (GLM).
+Additionally, the LSM may increase bias and inferential error when no omitted network structure is present, though we find that this effect may be reduced by appropriately scaling the prior of the latent positions.
+However, we find that the LSM substantially outperforms a GLM in terms of prediction.
+We conclude that the LSM should not be used to adjust for unmeasured confounders and is best used for exploratory and predictive tasks.
 
-\section{Background}
-
-\subsection{Applications and Development of the Latent Space Model}
+\section{Applications and Development of the Latent Space Model}
 
 The LSM has seen use in a variety of fields in which network data is common, particularly the social sciences.
-This appeal appears to have been driven primarily by the LSM's usefulness in modeling transitivity and homophily which appear to be ubiquitous in many (social) networks.
+The apparent appeal of the LSM appears to be driven primarily by the LSM's usefulness in modeling transitivity and homophily which are ubiquitous in social networks.
 In political science the LSM and variations on the form developed in \cite{hoff2002latent} have been used to estimate the effect of democracy on the probability of a militarized interstate dispute \citep{ward2007disputes}, the amount of portfolio investment between states \citep{cao2013democracies}, and the effect of multimember districts on the probability of collaboration between state legislators in the United States \citep{kirkland2012multimember}.
@@
 In ecology the LSM has been used to study the sociality of elephants \citep{vance2009social} and orcas \citep{fearnbach2014spatial}, to discover ecological communities \citep{fletcher2011social, fletcher2013network}, and to study food webs \citep{chiu2011unifying}.
-In epidemeology the LSM has been used to identify clusters of infected persons for later isolation \citep{zhang2015cluster} and to study patterns of interaction amongst physicians \citep{paul2014results}.
-In marketing and business research the LSM has been used to study inter-group trust \citep{dass2011impact}, optimal bundling and pricing of goods and brands for retailers \citep{dass2012assessing}.
-In neuroscience the LSM has been proposed as a method for modelling fMRI data \citep{simpson2013analyzing}.
+In epidemeology it has been used to identify clusters of infected persons for later isolation \citep{zhang2015cluster} and to study patterns of interaction amongst physicians \citep{paul2014results}.
+In marketing and business research it has been used to study inter-group trust \citep{dass2011impact}, optimal bundling and pricing of goods and brands for retailers \citep{dass2012assessing}.
+Lastly, in neuroscience it has been proposed as a method for modelling fMRI data \citep{simpson2013analyzing}.
 
@@
 
-Although the latent space model is arguably most useful as an exploratory or predictive model \citep{hoff2002latent, shmueli2010explain} it has been applied in some cases in a manner that suggests that users of the model assume that the latent space parameters reduce estimation and/or inferential error with respect to the effects of measured explanatory variables.
-Though it is not the case that all of the aforementioned articles explicitly or implicitly make such a claim (not all of the above applications of the LSM even include covariates), if the LSM does reduce inferential/estimation error under certain conditions, the LSM might find wider use due to the pervasiveness of network data where there remains unmeasured structure which either confounds relationships between measured covariates and edges, in which case the LSM might be applied to reuduce bias, or in which unmeasured structure is related to the strength/presence of edges and can be useful in decreasing estimator variance.
+Although the LSM is arguably most useful as an exploratory or predictive model (see \citet{shmueli2010explain} for a discussion of the differences between predictive and explanatory modeling), it has been applied in some cases to reduce estimation and/or inferential error with respect to the effects of measured covariates.
+If the LSM does reduce inferential/estimation error under certain conditions, it will serve as a valuable general model for explanatory analysis of network data, especially since its use does not require the researcher to specify a set of network dependencies from theory.
+However, as far as we know, there has been no research into the performance of the LSM in adjusting for confounding network structure.
 Additionally, the predictive performance of the LSM has only seen limited evaluation \citep{hoff2002latent}, despite having been used for model selection \citep{ward2013gravity, fletcher2011social, fletcher2013network, chiu2011unifying}.
-Hence, finding the condtions under which the LSM reduces prediction error also may affect future use of the LSM.
+Hence, finding the conditions under which the LSM reduces prediction error also may affect future use of the LSM.
 
-The LSM has also seen a substantial amount of subsequent development and extension.
+The LSM has seen a substantial amount of subsequent development and extension.
 The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors \citep{hoff2002latent, hoff2009multiplicative}.
@@
 \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
-
-\subsection{Collinearity in Generalized Linear Models}
-
-\citep{mackinnon1989collinearity} and \citep{weissfeld1991multicollinearity} investigate multicollinearity in generalized linear models and propose diagnostic criteria.
-Collinearity of the covariates causes the observed information matrix to be ill-conditioned, which produces numerical instability in the iterative estimation of $\hat{\mathbf{\beta}}$, the ``effects'' of covariates included in a regression.
-\citep{weissfeld1991multicollinearity}, extending the work of \citep{davis1986example}, propose the use of the condition index and the variance decomposition proportion of each condition index, to diagnose the presence and degree of multicollinearity in generalized linear models.
-The observed information matrix at the maximum likelihood estimate, $\mathcal{I}(\hat{\mathbf{\beta}})$ is scaled to have unit length and singular value decomposition is applied.
-The resulting right singular values are the eigenvectors and the associated singular values are the square root of the eigenvalues of the scaled observed information matrix since it is real Hermitian.
-The condition index is the ratio of the largest eigenvalue to each of the eigenvalues, and is at least 1.
-The square root of the condition number gives the potential magnification of the numerical error in inverting the observed information matrix.
-Hence our expectation is that as collinearity increases the variance of estimators of the effect of observed covariates will increase regardless of whether this is ommitted confounding structure in the data.
 
@@
 What priors are currently used and why?
-\cite{hoff2002latent} Propose the use of diffuse independent normal priors for the latent positions and regression parameters.
-The also recognize the fact that isolates' positions are weakly identified in the LSM, and in one application they actually just exclude the isolate.\footnote{Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space.
+\cite{hoff2002latent} propose the use of diffuse independent normal priors for the latent positions and regression parameters.
+They also recognize the fact that isolates' positions are weakly identified in the LSM, and in one application they actually just exclude the isolate.\footnote{Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space.
 However, such deletion would introduce bias in estimating covariate effects.} \cite{hoff2002latent} experiment with the use of an exponential prior for the intercept, which they argue sets a lower bound on the probability of a tie between nodes.
@@
 
-\citep{Krivitsky2009} implement a hierarchical model in which each node's latent space position is drawn from a mixture of $G$ normals. The covariate priors are single Gaussians.
+\citep{Krivitsky2009} implement a hierarchical model in which each node's latent position is drawn from a mixture of $G$ normals. The covariate priors are independent normals.
 
@@
 
-Would regularization/sparsity priors help with identifying the covariate effects?
-\cite{mitchell1988} propose a class of ``spike and slab'' prior as a class of priors for the coefficients in linear regression.
+Would regularizing/sparsity-inducing priors help with identifying the covariate effects?
+\cite{mitchell1988} propose a ``spike and slab'' prior as a class of priors for the coefficients in linear regression.
 These priors are well-suited to the problem of variable selection for linear regression.
@@
 This prior structure would build a preference into the model for using the observed covariates, rather than the latent coordinate parameters, to explain tie formation.
+This might reduce any bias that the LSM could introduce by preferentially using the latent positions to model variation due to the observed covariates.
 
-What priors could be used to avoid collinearity between observed covariates and latent space?
+% What priors could be used to avoid collinearity between observed covariates and latent space?
 Look to Bayesian Latent Trait models and specifically differential item functioning (DIF) in the IRT literature.
 
-Should priors be calibrated against observed covariates?
+Rather than using a mixture of distributions for the prior, we might instead use an informative prior.
 Using informative priors assures that the latent space coordinates are weakly identified, meaning that at some distance from the origin the increase in the likelihood from moving isolates away from the other nodes, or components away from each other, is offset by decreases in the prior from moving coordinates away from the origin.
@@
 
-$$ pr(y_{ij} = 1) = \text{logit}^{-1}(\beta_0 + \beta_1x_{ij} -  |z_i - z_j|).$$
+$$\mathbb{P}(y_{ij} = 1) = \text{logit}^{-1}(\alpha + \beta x_{ij} -  |z_i - z_j|)$$
 
-If we assume that the $z$ are independent and distributed $N(0,\sigma^2)$, it is straightforward to derive the distribution of $ |z_i - z_j|$.\footnote{The assumption that $z \sim N(0,\sigma^2$) is consistent with $z$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim N(0,2\sigma^2)$ \citep{devore2012}.
+If we assume that the $\mathbf{z}$ are independent and distributed $\mathcal{N}(0,\sigma^2)$, it is straightforward to derive the distribution of $|z_i - z_j|$.\footnote{The assumption that $\mathbf{z} \sim \mathcal{N}(0,\sigma^2$) is consistent with $\mathbf{z}$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim \mathcal{N}(0,2\sigma^2)$ \citep{devore2012}.
 The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}.
 Following from this result, we know that the variance of  $|z_i - z_j|$ is $$\sigma^2(2-4/\pi).
-$$ Let $\alpha^2$ be the empirical variance of $x$.
-Then the variance in the linear predictor is $\beta_1^2\alpha^2$.
-Given an estimate $\hat{\beta_1}$, we can set the variance of the latent space prior to approximately $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$ in order to tune the prior to avoid replicating $x$ via the distances between latent coordinates.
+$$ Let $\theta^2$ be the empirical variance of $\mathbf{x}$.
+Then the variance in the linear predictor is $\beta^2\theta^2$.
+Given an estimate $\hat{\beta}$, we can set the variance of the latent space prior to approximately $\hat{\beta}^2\theta^2/(2-4/\pi)$ in order to tune the prior to avoid replicating $\mathbf{x}$ via the distances between latent coordinates.
 This result is limited in that it requires the assumption of normally distributed latent coordinates and applies only to a single latent dimension.
@@
 
-\subsection{simulation design}
+\subsection{Simulation Design}
 
-To study how the LSM behaves in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space.
+We study how the LSM behaves in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space.
 For simplicity we consider a unidimensional latent space.
 To generate an observed covariate which has a controllable collinearity with the latent network structure we follow a three-step process.
-First, we simulate $k$-dimensional positions for each node, drawing each from a normal distribution, and calculate dyadic distances $d$ between each pair of positions.
-Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle X,d \rangle$, we derive the conditional mean vector and covariance, assuming that $X$ has a joint normal distribution given $d$ (see \cite[pp.
+First, we simulate unidimensional positions for each node, drawing from a normal distribution, and calculate the Euclidean distance $\mathbf{d}$ between each pair of positions.
+Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle \mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector and covariance, assuming that $\mathbf{x}$ has a normal distribution given $\mathbf{d}$ (see \cite[pp.
 116--117]{eaton1983} for the conditional normal derivation).
-Third, we simulate $X$ as a multivariate normal random variable with the respective conditional means and covariance.
-Finally we standardize each $X$ to have zero mean and unit variance.
-To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $X$ we utilize the C-vine method of \cite{lewandowski2009generating}.
+Third, we simulate $x$ as a normal random variable with the respective conditional means and covariance.
+Finally we standardize $x$ to have zero mean and unit variance.
+To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $\mathbf{x}$ we utilize the C-vine method of \cite{lewandowski2009generating}.
 % fill in a basic description of the method here
 
-We consider three exponential family distributions for the edges: Gaussian, Binomial, and Poisson.
-Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space $k$ at 1 for simplicity, and consider 4 values of the collinearity parameter $\eta$, $.1$, $1$, $100$, and $1000000$, where $.1$ a corresponds to near perfect collinearity and $1000000$ to independence.
+We consider three exponential family distributions for the edges: Gaussian, binomial, and poisson.
+Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space at 1 for simplicity, and consider three values of the collinearity parameter $\eta$, $1$, $100$, and $1000000$, where $1$ corresponds to a uniform distribution over $[0, 1]$, moderate collinearity, and $1,000,000$ to independence.
 See Figure \ref{fig:vine} for the distribution of the absolute value of the maximal off-diagnoal correlation generated at each value of $\eta$.
@@
 
-Each step in the simulation involves the estimation of a GLM, and the LSM.
 We consider the LSM with several different priors on the coefficient for the observed covariate and the latent space.
-We set the prior variance of $\beta_1$ to be either $1$ or $10$ and alternatively use a diffuse normal prior on the latent space or scale it by $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$, where $\alpha^2$ is the empirical variance of $X$.
+We set the prior variance of $\beta$ to be either $1$ or $10$ and alternatively use a diffuse normal prior on the latent space or scale it by $\hat{\beta}^2\theta^2/(2-4/\pi)$, where $\theta^2$ is the empirical variance of $x$.
 
-At each step in the simulation, the LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}.
+The LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}.
 An initial run of 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler.
@@
 If the convergence criteria is still not satisfied, then the aforementioned step in the simulation is flagged for review.
-At each point in the simulation's parameter space, we execute 500 Monte Carlo iterations.
+At each point in the simulation's parameter space, we execute 1,000 Monte Carlo iterations.
 
 One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM.
-We utilize the \texttt{BatchJobs} and \texttt{BatchExperiments} \R packages to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.
+We utilize the \texttt{BatchExperiments} \R package to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.
 
-We first compute the MLE estimated by iteratively reweighted least squares in \texttt{R}.
-For the LSM we use the posterior mode.
-In the cases where edges are Binomial distributed, we scale the estiamtes using the reciprocal of the bias of $\beta_1$ when $X$ and $d$ are independent: $\frac{\sqrt{3.28 + \beta_2^2 \text{Var(d)}}}{\sqrt{3.29}}$, where $3.29$ is the variance of a standard logistic distribution (See the derivation of the bias under an independent but ommitted covariante in \cite{mood2010logistic}).
+We compute the MLE estimated by iteratively reweighted least squares in \texttt{R}.
+For the LSM we use the posterior mode as our point estimate.
+In the cases where edges are binomial and there is omitted network structure, we scale the estiamtes using the reciprocal of the bias of $\beta$ when $\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 + \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the variance of a standard logistic distribution (See the derivation of the bias under an independent but ommitted covariate in \cite{mood2010logistic}).
 
@@
 number of nodes) and collinearity structure between the ommitted distance between nodes and the measured covariates.
-For each point in this space we evaluate the mean square prediction error of the model on new data drawn from the same distribution (i.e., the generalization error), the bias of the estimated coefficients for the measured covariate, the type 2 error rate (i.e., the probability that the null hypothesis that each regression parameter is $0$ is incorrectly accepted), and the type 1 error rate, that is, the probability of rejecting the null hypothesis that the effect of the measured covariate is 0 when it is in fact 0.
-Inference regarding $\beta_1$ is computed using the maximum a posteriori estimate (i.e., the mode of the posterior), and credible intervals for $\beta_1$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution for each regression parameter.
+For each point in this space we evaluate the mean square prediction error of the model on new edges drawn condtional on $\mathbf{x}$ and $\mathbf{d}$, the bias of the estimated coefficient for the measured covariate, and the Type-1 and 2 error rates.
+Inference regarding Credible intervals for $\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution.
 
-To evaluate estimation error we compute the bias: the difference between the true coefficient $\beta_1$ and the estimate.
+To evaluate estimation error we compute the bias.
 Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results.
-When a latent space is present (i.e., $\beta_2 = -1$, shown in Figure \ref{fig:estimation_ls}) the LSM outperforms the GLM in terms of bias across the board.
-However, when latent network structure is not present ($\beta_2 = 0$, shown in Figure \ref{fig:estimation_nls}) the LSM is biased upwards in the Binomial case and appears to have higher variance (perhaps due to efficiency losses) for Poisson and Gaussian edges (preliminary interpretation).
+In \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the measured covariate, all methods (except the true model wherein the distances are treated as observed )show substantial bias.
+In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$, however, in absolute terms the amount of bias is large.
+For both the GLM and the LSM bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases.
+In \ref{fig:estimation_nls} we can see that when there is no ommitted network structure the LSM exhibits bias greater than that of the GLM, though it does appear that scaling the prior of the latent positions to match the scale of the measured covariate decreases this tendency.
 
@@
 \includegraphics[width=\textwidth]{estimation_ls.png}
-\caption{The bias of estimates of the effect of the observed covariate $X$ when there is an ommitted variable.
-The $x$-axis gives the value of the parameter $\eta$ which controls the degree of dependence between $X$ and ommitted covariate.
+\caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is an ommitted variable.
+The $x$-axis gives the value of the parameter $\eta$ which controls the degree of dependence between $\mathbf{x}$ and ommitted covariate.
 Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate.
@@
 The number of nodes are indicated in the top panels, while the distributional family of the edges is shown on the right panel.
-Each panel represents 4 values of $\eta$ with 500 (partially complete) Monte Carlo iterations executed at each point.
+Each panel represents 4 values of $\eta$ with 1,000 Monte Carlo iterations executed at each point.
 \label{fig:estimation_ls}}
@@
 
-Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta_1$ and the latent space compared to a GLM.
-Type 1 error rates for the LSM are in general lower than those of a GLM when an unmeasured covariate exists, however, when one does not exist the GLM outperforms the LSM.
-However, scaling the prior of the latent space to match that of the observed covariates appears to make the LSM's error rate comparable to that of the GLM.
-Type 2 error rates are very similar for the GLM and the LSM with all evaluated prior settings; to the point that they are indistinguishable.
+Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta$ and the latent positions, as well as a GLM.
+Type-1 error rates shown in \ref{fig:inference_type_1} for the LSM are in general lower than those of a GLM when an unmeasured covariate exists.
+Under uniform correlation between the omitted network structure and the observed covariate the Type-1 error rate is still often more than 10 times greater than the nominal rate of .05.
+Again, scaling the prior of the latent space to match that of the observed covariate appears to make the LSM's error rate comparable to that of the GLM.
+Type-2 error rates, shown in \ref{fig:inference_type_2} are also substantially above their nominal rate when there is a uniform distribution on the strength of confounding, though the inflation is not as bad as that of the Type-1 rates.
 
@@
 \includegraphics[width=\textwidth]{inference_type_1.png}
-\caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $X$ are shown on the $y$$-axis.
-Here, $\beta_1 = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta_1 = 0$ is falsely rejected.
+\caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y$-axis.
+Here, $\beta = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta = 0$ is falsely rejected.
 \label{fig:inference_type_1}}
@@
 \includegraphics[width=\textwidth]{inference_type_2.png}
-\caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $X$ are shown on the $y-axis$.
-Here, $\beta_1 = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis.
+\caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y-axis$.
+Here, $\beta = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis.
 \label{fig:inference_type_2}}
@@
 
-Generalization error, in this case defined as the expected error on new edges generated using a fixed set of latent distances between the nodes and a node-level covariate is shown in figure \ref{fig:generalization}.
+Generalization error, in this case defined as the expected prediction error on new edges generated using a fixed set of latent positions for the nodes and a fixed observed covariate is shown in figure \ref{fig:generalization}.
 In nearly all cases the LSM substantially outperforms the GLM, especially when the ommitted covariate is not highly collinear with the observed covariate.
+In nearly all cases it performs as well as the true model.
 
@@
 \includegraphics[width=\textwidth]{generalization.png}
-\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the generative model using a model trained on independent data. \label{fig:generalization}}
+\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed.
+In the binomial case the Brier score is computed and in the poisson and normal cases the mean-square-error.
+\label{fig:generalization}}
 \end{figure}
@@
 
-Although the results are (very) preliminary (and have been plagued by my incompetence!), it seems increasingly clear that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space to uncover patterns of interest or to predict edges in networks where there is likely to be omitted structure that is not adequately modelled using measured covariates.
-However, inferential and estimation error are often smaller than that of a GLM when an omitted variable that corresponds to a form that can be modelled using a latent space is present, and can lead to an up to $25\%$ decrease in bias relative to a GLM.
-Whether or not there is a reduction in estimation and/or inferential error depends on the strength of the ommitted confounder, sample size (the LSM having many more parameters than an equivalent GLM), and relies on the ommitted confounder being representable as a latent space.
-Lastly, there is some (preliminary) evidence to suggest that scaling the prior of the latent space to match that of the measured covariates can reduce Type-1 error.
+We conclude that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space to uncover patterns of interest or the prediction of edges in networks where there is likely to be omitted structure that is not adequately modelled using measured covariates.
+Although inferential and estimation errors are somewhat smaller than that of a GLM when there is omitted network structure, even moderate collinearity between the omitted structure and the measured covariate leads to substantial bias and inferential error.
+The LSM cannot control for unmeasured confounders.
+Additionally, if such network structure does not exist, the LSM may induce additional bias and inferential error, though this effect can be moderated by scaling the prior on the latent positions to match that of the measured covariates.
 

### 2016-06-24 analysis.R

@@ -1,52 +1,79 @@
 pkgs <- c("dplyr", "reshape2", "ggplot2", "stringr", "coda", "scales")
 invisible(lapply(pkgs, library, character.only = TRUE))
 
-results <- read.csv("results.csv", stringsAsFactors = FALSE) %>%
-  select(-one_of("X", "id", "prob", "repl")) %>%
+results <- read.csv("results_bruce.csv", stringsAsFactors = FALSE) %>%
+  select(-one_of("X", "prob", "repl", "id")) %>%
   rename(model = algo) %>%
   mutate(model = ifelse(!is.na(scale) & scale, "lsm_scaled", model)) %>%
   mutate(model = ifelse(!is.na(beta.var), paste0(model, " (beta.var = ", beta.var, ")"), model)) %>%
+  filter(eta != .1 & !(model == "truth" & latent_space == 0)) %>%
   mutate(bias = estimate - beta, error = ifelse(beta == 0, 1 - coverage, coverage),
          type = ifelse(beta == 0, 1, 2)) %>%
   group_by(model, family, eta, beta, type, latent_space, nodes) %>%
-  summarise(estimate = mean(estimate), loss = mean(loss), coverage = mean(coverage), bias = mean(bias),
-            error = mean(error))
+  summarise(n = length(estimate[!is.na(estimate)]),
+            estimate = mean(estimate, na.rm = TRUE),
+            loss = mean(loss, na.rm = TRUE),
+            coverage = mean(coverage, na.rm = TRUE),
+            bias = mean(bias, na.rm = TRUE),
+            error = mean(error, na.rm = TRUE))
 
-ggplot(results[results$beta == 1, ], aes(eta, loss, color = model, linetype = model)) +
-  scale_x_log10(label = scientific_format()) +
+ggplot(results, aes(eta, n, color = model, linetype = model)) +
+  scale_x_log10(breaks = c(1, 100, 1000000), labels = c("uniform correlation", "low correlation", "independence")) +
+  scale_y_log10() +
+  geom_point() + geom_line() +
+  facet_grid(family + beta ~ nodes + latent_space, scales = "free", labeller = label_context) +
+  labs(x = expression(eta), y = "Iterations Completed") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+ggsave("iterations.png", width = 12, height = 8)
+
+ggplot(filter(results, beta == 1), aes(eta, loss, color = model, linetype = model)) +
[…]

### 2016-06-24 reduce.R

@@ -6,7 +6,8 @@ reg_bruce <- loadRegistry("../lsm_bruce/reg")
 
 reduce <- function(job, res) {
   if (!is.null(res)) {
-    if (any(class(res$fit) == "glm") & res$data$latent_space == -1 & res$data$family == "binomial") {
+    if (any(class(res$fit) == "glm") & res$data$latent_space == -1 & res$data$family == "binomial" &
+          length(coef(res$fit)) < 3 & length(coef(res$fit)) <= 2) {
       estimate <- res$estimate * res$adjustment
       coverage <- unname(ifelse(res$interval[1] * res$adjustment < 0 & res$interval[2] * res$adjustment > 0, 1, 0))
     } else {
@@ -19,9 +20,8 @@ reduce <- function(job, res) {
   }
 }
 
-nodes <- 20L
 res_reg_open <- batchMapQuick(reduceResultsExperiments,
-                              chunk(findDone(reg_open), n.chunks = nodes, shuffle = TRUE),
+                              chunk(findDone(reg_open), n.chunks = 80, shuffle = TRUE),
                               more.args = list(reg = reg_open, fun = reduce, progressbar = FALSE,
                                                impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
                               resources = list(walltime = 86400L / 2, nodes = 1L, memory = "8gb"))
@@ -29,11 +29,10 @@ waitForJobs(res_reg_open)
 res_open <- reduceResults(res_reg_open, fun = function(aggr, job, res) rbind.fill(aggr, res), init = data.frame())
 
 res_reg_bruce <- batchMapQuick(reduceResultsExperiments,
-                               chunk(findDone(reg_bruce), n.chunks = nodes, shuffle = TRUE),
+                               chunk(findDone(reg_bruce), n.chunks = 200, shuffle = TRUE),
                                more.args = list(reg = reg_bruce, fun = reduce, progressbar = FALSE,
                                                 impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
                                resources = list(walltime = 86400L, nodes = 1L, memory = "8gb"))
 waitForJobs(res_reg_bruce)
 res_bruce <- reduceResults(res_reg_bruce, fun = function(aggr, job, res) rbind.fill(aggr, res), init = data.frame())
-
 write.csv(rbind(res_open, res_bruce), "results.csv")

### 2016-06-24 simulation.R

@@ -1,6 +1,8 @@
 library("BatchExperiments")
 library("latentnet")
 library("foreach")
+library("clusterGeneration")
+library("iterators")
 
 create_network <- function(y, name = "eweight") {
   library("network")
@@ -141,7 +143,7 @@ lsm.wrapper <- function(static, dynamic, ...) {
   list(
     "pred" = pred,
     "estimate" = fit$mcmc.pmode$beta[2],
-    "interval" =  HPDinterval(as.mcmc(fit$sample$beta[, 2])),
+    "interval" = HPDinterval(as.mcmc(fit$sample$beta[, 2])),
     "loss" = ms_error(pred, dynamic$y_new),
     "convergence" = geweke,
     "data" = dynamic,
@@ -212,15 +214,18 @@ batchExport(reg, create_network = create_network, ms_error = ms_error,
             rmvnorm_d = rmvnorm_d, unstack_vector = unstack_vector, diagnostic = diagnostic,
             overwrite = TRUE)
 
-resources <- list(walltime = 86400L * 5, nodes = 1L, memory = "8gb")
+resources <- list(walltime = 86400L * 1, nodes = 1L, memory = "8gb")
+
+## from earlier version
+removeExperiments(reg, findExperiments(reg, prob.pars = (eta < 1)))
 
 ids <- findNotStarted(reg)
 ids <- ids[!ids %in% findOnSystem(reg)]
-submitJobs(reg, chunk(ids, n.chunks = 1000, shuffle = TRUE)[1:100], resources)
+submitJobs(reg, chunk(ids, n.chunks = 80, shuffle = TRUE), resources)
 
 ## generate a histogram plot for the off diagnoals of the covariance matrix used to generate
 ## d and X_stack
-eta_hist <- foreach(eta = c(.1, 1, 100, 1000000), .combine = "rbind") %:%
+eta_hist <- foreach(eta = c(1, 100, 1000000), .combine = "rbind") %:%
   foreach(icount(1000), .combine = "rbind") %do% {
   c(eta, abs(cov2cor(genPositiveDefMat(2, "c-vine", eta = eta)$Sigma))[1, 2])
 }
@@ -229,6 +234,8 @@ eta_hist <- data.frame(eta_hist)
 colnames(eta_hist) <- c("eta", "cor")
 p <- ggplot(eta_hist, aes(x = cor))
 p <- p + geom_histogram()
-p <- p + facet_wrap(~ eta)
[…]

### 2016-07-26 manuscript.tex
post polmeth update

@@
 In the presence of an unmeasured covariate that can be appropriately modeled using a latent space, estimation and inferential error remain high under even moderate confounding.
-Additionally, when omitted network structure is not present the LSM induces additional bias and inferential error.
 However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases.
-We conclude that the LSM is most appropriately used for exploratory or predictive tasks.
+We conclude that the LSM is most appropriately used for exploratory or predictive tasks.\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219.
+Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}
 
@@
 
-Despite their growing popularity, few studies exist that investigate the performance of these models in adjusting for confounders that can be represented as network structure.
-The approach to adjusting for dependencies in the ERGM and SAOM is quite similar to adjusting for confounding covariates in regression modeling.
+Despite their growing popularity, few studies exist that investigate the performance of these models in adjusting for confounding network structure.
+The approach to adjusting for dependencies in the two other models commonly used for network data---ERGM and SAOM---is quite similar to adjusting for confounding covariates in regression modeling.
 The researcher specifies a set of dependencies that (s)he hypothesizes to be important in the generative model for the network.
@@
 The LSM has the advantage over ERGM and SAOM in that researchers need not develop a set of hypothesized dependencies in order to model network structure that is not reflected in observed covariates.
+However, this advantage hinges upon the capacity for the LSM to discover unmodeled structure that could otherwise be attributed to the observed covariates (i.e., inferring confounding structure).
 In the current study, we focus on the LSM, examining its performance in reducing estimation and inferential errors regarding the effects of observed covariates, via adjustment for confounding network structure.
@@
 
-We develop a simulation design in which we control the degree of collinearity between omitted network structure and observed covariates.
-We find that the LSM does not substantially reduce bias and inferential error with respect to the effects of observed covariates, in comparison to a generalized linear model (GLM).
-Additionally, the LSM may increase bias and inferential error when no omitted network structure is present, though we find that this effect may be reduced by appropriately scaling the prior of the latent positions.
-However, we find that the LSM substantially outperforms a GLM in terms of prediction.
-We conclude that the LSM should not be used to adjust for unmeasured confounders and is best used for exploratory and predictive tasks.
-
 \section{Applications and Development of the Latent Space Model}
@@
 The LSM has seen use in a variety of fields in which network data is common, particularly the social sciences.
-The apparent appeal of the LSM appears to be driven primarily by the LSM's usefulness in modeling transitivity and homophily which are ubiquitous in social networks.
+The apparent appeal of the LSM appears to be driven primarily by the LSM's usefulness in modeling transitivity (i.e., clustering) and homophily, which are ubiquitous in social networks.
 In political science the LSM and variations on the form developed in \cite{hoff2002latent} have been used to estimate the effect of democracy on the probability of a militarized interstate dispute \citep{ward2007disputes}, the amount of portfolio investment between states \citep{cao2013democracies}, and the effect of multimember districts on the probability of collaboration between state legislators in the United States \citep{kirkland2012multimember}.
 Variations of the LSM developed for networks measured over time have been applied to the study of international trade, wherein the effects of various features of trading partners are estimated \citep{ward2013gravity}.
-In ecology the LSM has been used to study the sociality of elephants \citep{vance2009social} and orcas \citep{fearnbach2014spatial}, to discover ecological communities \citep{fletcher2011social, fletcher2013network}, and to study food webs \citep{chiu2011unifying}.
+In ecology the LSM has been used to study the sociality of elephants \citep{vance2009social} and orcas \citep{fearnbach2014spatial}, birds \citep{nomano2015unrelated}, to discover ecological communities \citep{fletcher2011social, fletcher2013network}, and to study food webs \citep{chiu2011unifying}.
 In epidemeology it has been used to identify clusters of infected persons for later isolation \citep{zhang2015cluster} and to study patterns of interaction amongst physicians \citep{paul2014results}.
 In marketing and business research it has been used to study inter-group trust \citep{dass2011impact}, optimal bundling and pricing of goods and brands for retailers \citep{dass2012assessing}.
+It has been used to describe topic-specific patterns of interaction in e-mail communication networks \citep{krafft2012topic}.
 Lastly, in neuroscience it has been proposed as a method for modelling fMRI data \citep{simpson2013analyzing}.
@@
 
-Although the LSM is arguably most useful as an exploratory or predictive model (see \citet{shmueli2010explain} for a discussion of the differences between predictive and explanatory modeling), it has been applied in some cases to reduce estimation and/or inferential error with respect to the effects of measured covariates.
+Although the LSM is arguably most useful as an exploratory or predictive model (see \citet{shmueli2010explain} for a discussion of the differences between predictive and explanatory modeling), it has been applied in some cases to reduce estimation and/or inferential error with respect to the effects of observed covariates.
+For example \citet{ward2007disputes}, in a high profile example, argue that the the LSM improves inference about the effects of democracy, international trade, and participation in international organizations on the probability of inter-state conflict.
+
+\begin{quote}
+``The history of international disputes, and consequently the extant data on militarized interstate disputes, is replete with \ldots dependencies.
+We formally incorporate and estimate the extent of these \ldots dependencies in our model of the Kantian peace in order to more precisely determine the effects of the Kantian tripod on international conflict'' \citep[p.
+585]{ward2007disputes}
+\end{quote}
+
+Likewise, similar claims are made in \citet{dorff2013networks, cao2013democracies, ward2013gravity, kirkland2012multimember}.
+
+\begin{quote}
+``For the most part, however, most dyadic research in international relations ignores the essential features of dyads in that they fail to satisfy the assumption of independence or, by construction, have missing data but ignore its effects: both of these bias the results in a fundamental way'' \citep[p.
+2]{dorff2013networks}.
+\end{quote}
+
+\begin{quote}
+``This approach combines a network analysis with a standard-looking regression to permit us to access the importance of our explanatory factors without having them biased by the interdependencies in the network we are studying'' \citep[p.
+15]{cao2013democracies}.
+\end{quote}
+
+\begin{quote}
+``The presence of \ldots dependence implies misspecification and a high likelihood of bias in most current applications.
+Building on the latent space framework we model the world trading system without assuming a particular network structure or the sufficiency of particular network statistics'' \citep[p.
+20]{ward2013gravity}.
+\end{quote}
+
+\begin{quote}
+``\ldots the latent space model allows for the assessment of distance between two unconnected actors while simultaneously controlling for the interdependence inherent in network data.
+This interdependence in latent space positions allows the model to control for common network effects like reciprocity or transitivity that would ordinarily bias results'' \citep[p.
+336]{kirkland2012multimember}.
+\end{quote}
+
+The above examples are drawn from political science, but we see a similar logic for using the LSM articulated in recent work in ecology by \citet[p. 989]{nomano2015unrelated}.
+
+\begin{quote}
+``\ldots can create artificially exaggerated synchrony rates regardless of motivation for signaling, which can be modeled with a term known as ``transitivity'' in the social network literature.
+A latent space model (Krivitsky et al.
+2009) was used to examine the propensities for synchrony between helper males and the primary male while accounting for the variability deriving from the transitivity.''
+\end{quote}
+
+Though this list of quotations is by no means complete, the last example to which we point comes from the business literature, and uses the bilinear form of the latent space model \citep[p.
+7]{dass2011impact}.
+
+\begin{quote}
+``This model controls for higher order team dynamics using the bilinear component $z_i'z_j$ and also allows
+both trustor ($x_{tor,i}$ ) and trusted ($x_{ted,j}$ ) characteristics to be investigated along with the dyadic covariates ($x_{d,i,j}$).''
+\end{quote}
+
 If the LSM does reduce inferential/estimation error under certain conditions, it will serve as a valuable general model for explanatory analysis of network data, especially since its use does not require the researcher to specify a set of network dependencies from theory.
@@
 \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
+Most of these developments have represented new forms for the latent space and/or the further development of the LSM as a Bayesian hierarchical model.
+In the next section we propose an approach to specifying the Bayesian prior in order to improve inference regarding covariate parameters.
 
-\section{Research Design}
+\section{Bayesian Inference and Priors in the Latent Space Model}
 
-\subsection{Priors for the Latent Space Model}
-
-\subsubsection{Current Implementations}
-
-What priors are currently used and why?
+Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions), which means that the ML estimate is unidentified.
+As such, inference in the LSM is conducted using a Bayesian approach.
+Latent positions and other parameters are sampled using a Metropolis-Hastings algorithm.
+After sampling, latent space positions are rotated to match a common set of positions, via Procrustes transformation \citep{borg2005modern}, in order to eliminate variance due to rotation of the latent space.
 \cite{hoff2002latent} propose the use of diffuse independent normal priors for the latent positions and regression parameters.
-They also recognize the fact that isolates' positions are weakly identified in the LSM, and in one application they actually just exclude the isolate.\footnote{Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space.
-However, such deletion would introduce bias in estimating covariate effects.} \cite{hoff2002latent} experiment with the use of an exponential prior for the intercept, which they argue sets a lower bound on the probability of a tie between nodes.
-Of course, the latent space positions could simply drift further apart in order to compensate for the high intercept.
-The following paragraph on priors appears on page 1097.
-
-\begin{quote}
-We have not discussed in detail the choice of a prior distribution for latent positions in this article.
+They also recognize the fact that isolates' (i.e., nodes with no ties) finite positions are only identified through the prior in the LSM.
+In one application they actually just exclude the isolates.
+Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space, but for representative explanatory analyses excluding isolates would amount to selecting on the dependent variable.
+Unlike in many practical applications of Bayesian inference, it is not possible to use an improper ``flat'' prior \citep{tibshirani1989noninformative} with the LSM, as an informative prior is necessary to assure finite latent positions.
+%\begin{quote}
+%We have not discussed in detail the choice of a prior distribution for latent positions in this article.
 Although simple, the diffuse independent normal priors presented in the examples may not accurately represent prior beliefs about the structure of social networks.
@@
 This would add another level of hierarchy to the analysis, although the resulting model would be more flexible and perhaps more accurately represent any tendencies of populations to form segregating groups.
-\end{quote}
+%\end{quote}
+Since it is necessary to use an informative prior with the LSM, we ask whether it is possible to specify the prior in a way that improves inference regarding covariate parameters.
+We propose to calibrate the prior to be on a scale comparable to the linear predictor estimated from a GLM.
+This scaling should discourage the latent space from replicating the linear predictor, encouraging the discovery of structure that remains in the data after identifying the effects of measured covariates.
 
-\citep{Krivitsky2009} implement a hierarchical model in which each node's latent position is drawn from a mixture of $G$ normals. The covariate priors are independent normals.
-
-\subsubsection{Alternatives to consider}
-
-Would regularizing/sparsity-inducing priors help with identifying the covariate effects?
-\cite{mitchell1988} propose a ``spike and slab'' prior as a class of priors for the coefficients in linear regression.
-These priors are well-suited to the problem of variable selection for linear regression.
-The prior is formed as a mixture between a diffuse uniform prior and a point mass at zero.
-Such a prior would not be appropriate for latent space coordinates since weakly identified coordinates of isolates or separate components would likely exhibit strange behavior at the boundary of the diffuse uniform.
-We might consider instead mixing, e.g., a point mass at zero with a distribution such as the Cauchy, which is fairly flat in the tails, but is defined on the entire real line and still exhibits a surface that descends from the center of the distribution.
-This prior structure would build a preference into the model for using the observed covariates, rather than the latent coordinate parameters, to explain tie formation.
-This might reduce any bias that the LSM could introduce by preferentially using the latent positions to model variation due to the observed covariates.
-
-% What priors could be used to avoid collinearity between observed covariates and latent space?
-Look to Bayesian Latent Trait models and specifically differential item functioning (DIF) in the IRT literature.
-
-Rather than using a mixture of distributions for the prior, we might instead use an informative prior.
-Using informative priors assures that the latent space coordinates are weakly identified, meaning that at some distance from the origin the increase in the likelihood from moving isolates away from the other nodes, or components away from each other, is offset by decreases in the prior from moving coordinates away from the origin.
-However, we may also use the priors to penalize the model for inferring latent coordinates that result in distances that are comparable in scale to the observed covariates.
-This would embed a tendency for the model to use the observed covariates, rather than the latent space, to model the outcome.
-Consider the following example of a one-dimensional spatial model with an additional dyadic covariate.
-
-$$\mathbb{P}(y_{ij} = 1) = \text{logit}^{-1}(\alpha + \beta x_{ij} -  |z_i - z_j|)$$
-
+To derive our scaling rule, we follow the thought experiment of a one-dimensional latent space $\mathbf{z}$ with a dyadic covariate $\mathbf{x}$.
 If we assume that the $\mathbf{z}$ are independent and distributed $\mathcal{N}(0,\sigma^2)$, it is straightforward to derive the distribution of $|z_i - z_j|$.\footnote{The assumption that $\mathbf{z} \sim \mathcal{N}(0,\sigma^2$) is consistent with $\mathbf{z}$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim \mathcal{N}(0,2\sigma^2)$ \citep{devore2012}.
@@
 Then the variance in the linear predictor is $\beta^2\theta^2$.
-Given an estimate $\hat{\beta}$, we can set the variance of the latent space prior to approximately $\hat{\beta}^2\theta^2/(2-4/\pi)$ in order to tune the prior to avoid replicating $\mathbf{x}$ via the distances between latent coordinates.
-This result is limited in that it requires the assumption of normally distributed latent coordinates and applies only to a single latent dimension.
-However, it may provide a reasonable approximation in the case of non-normal coordinates and/or multiple dimensions.
-\cite{thirey2015} derive the distribution of Euclidean Distances between two $k$-dimensional points where each coordinate is drawn from a standard normal distribution.
+Given an estimate $\hat{\beta}$, we can set the variance of the latent space prior to a multiple of $\hat{\beta}^2\theta^2/(2-4/\pi)$ in order to tune the prior to avoid replicating the linear predictor via the distances between latent coordinates.
 
-\section{Analysis}
+\subsection{Simulation Study}
 
-\subsection{Simulation Design}
+We use a simulation study to evaluate the LSM's performance at inferring covariate parameters in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space.
+We are particularly interested in whether the LSM can adjust for confounding when the confounding variable is unmeasured.
+In the simulation design, we study how the level of confounding affects the performance of the LSM relative to the GLM.
+We evaluate performance in terms of bias, inferential error, and prediction.
 
-We study how the LSM behaves in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space.
-For simplicity we consider a unidimensional latent space.
+\subsection{Simulation Setup}
+
+For simplicity and computational efficiency we consider a unidimensional latent space.
+We have no reason to suspect that our results should be unique to unidimensional LSMs.
 To generate an observed covariate which has a controllable collinearity with the latent network structure we follow a three-step process.
@@
 
-We consider three exponential family distributions for the edges: Gaussian, binomial, and poisson.
-Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space at 1 for simplicity, and consider three values of the collinearity parameter $\eta$, $1$, $100$, and $1000000$, where $1$ corresponds to a uniform distribution over $[0, 1]$, moderate collinearity, and $1,000,000$ to independence.
-See Figure \ref{fig:vine} for the distribution of the absolute value of the maximal off-diagnoal correlation generated at each value of $\eta$.
+We consider three exponential family distributions from which the adjacency matrix entries are drawn: Gaussian, binomial, and Poisson.
+Additionally we control the number of nodes in the network $n = 25, 50, 100$.
+To vary the degree of confounding attributable to the latent positions, we consider three values of the collinearity parameter $\eta$, $1$, $100$, and $1,000,000$, where $1$ corresponds to a standard uniform distribution of the correlation  (i.e.,  moderate collinearity) between the observed covariate and the latent distances, and $1,000,000$ to independence between the observed covariate and the latent distances.
+See Figure \ref{fig:vine} for the distribution of the absolute value of the correlation generated at each value of $\eta$.
 
 \begin{figure}
-\includegraphics[width=\textwidth]{max_r_vine.png}
-\caption{The distribution of the absolute value of the maximum off-diagonal correlation. \label{fig:vine}}
+\includegraphics[width=\textwidth]{figures/max_r_vine.png}
+\caption{The distribution of the absolute value of the correlation between the observed covariate and the latent distances. \label{fig:vine}}
 \end{figure}
@@
 
-The LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}.
+The LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R} \citep{latentnet}.
 An initial run of 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler.
 Every 100th iteration is saved.
-Convergence in the log probability of the model is assessed using the Geweke diagnostic.
-If the convergence criteria is satisfied the simulation continues to the next set of arguments, otherwise the number of iterations is doubled.
+Convergence in the log probability of the model is assessed using the Geweke diagnostic in the \texttt{coda} package \citep{coda, geweke1991evaluating}.
+If the convergence criterion is satisfied the simulation continues to the next set of arguments, otherwise the number of iterations is doubled.
 If the convergence criteria is still not satisfied, then the aforementioned step in the simulation is flagged for review.
 At each point in the simulation's parameter space, we execute 1,000 Monte Carlo iterations.
+\footnote{One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM.
+We utilize the \texttt{BatchExperiments} \R package to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.}
 
-One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM.
-We utilize the \texttt{BatchExperiments} \R package to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.
+We compute the MLE estimated by iteratively reweighted least squares via the \texttt{glm} function in \texttt{R} \citep{rcore}.
+For the LSM we use the posterior mode as our point estimate.
+In the cases where edges are binomial and there is omitted network structure, we scale the estimates using the reciprocal of the bias of $\beta$ when $\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 + \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the variance of a standard logistic distribution.
+We do this to adjust for bias in the coefficient estimate that arises due to the lack of a scale coefficient in logistic regression (See the derivation of the bias under an independent but omitted covariate in \cite{mood2010logistic}).
 
-We compute the MLE estimated by iteratively reweighted least squares in \texttt{R}.
-For the LSM we use the posterior mode as our point estimate.
-In the cases where edges are binomial and there is omitted network structure, we scale the estiamtes using the reciprocal of the bias of $\beta$ when $\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 + \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the variance of a standard logistic distribution (See the derivation of the bias under an independent but ommitted covariate in \cite{mood2010logistic}).
+For each combination of simulation conditions we evaluate the mean square prediction error of the model on new edges drawn condtional on $\mathbf{x}$ and $\mathbf{d}$, the bias of the estimated coefficient for the measured covariate, and the Type-1 and 2 error rates.
+Inference regarding Credible intervals for $\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution \cite{turkkan1993computation}.
+We consider $\beta$ to be statistically significant at the 0.05 level if the credible interval does not contain 0.
 
-We consider networks of varying size (i.e.
-number of nodes) and collinearity structure between the ommitted distance between nodes and the measured covariates.
-For each point in this space we evaluate the mean square prediction error of the model on new edges drawn condtional on $\mathbf{x}$ and $\mathbf{d}$, the bias of the estimated coefficient for the measured covariate, and the Type-1 and 2 error rates.
-Inference regarding Credible intervals for $\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution.
+\subsection{Results}
 
@@
 Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results.
-In \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the measured covariate, all methods (except the true model wherein the distances are treated as observed )show substantial bias.
-In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$, however, in absolute terms the amount of bias is large.
-For both the GLM and the LSM bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases.
-In \ref{fig:estimation_nls} we can see that when there is no ommitted network structure the LSM exhibits bias greater than that of the GLM, though it does appear that scaling the prior of the latent positions to match the scale of the measured covariate decreases this tendency.
+In Figure \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are treated as observed) show substantial bias.
+In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$.
+However, in absolute terms the amount of bias is large.
+Our results demonstrate that using the LSM will not permit the discovery of and adjustment for the true latent positions.
+For both the GLM and the LSM, bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases.
+In \ref{fig:estimation_nls} we can see that when there is no ommitted network structure, the LSM exhibits bias greater than that of the GLM, though it does appear that scaling the prior of the latent positions to match the scale of the linear predictor decreases this tendency.
 
 \begin{figure}
-\includegraphics[width=\textwidth]{estimation_ls.png}
+\includegraphics[width=\textwidth]{figures/estimation_ls.png}
 \caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is an ommitted variable.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{estimation_nls.png}
-\caption{The bias of estimates of the effect of the observed covariate $X$ when there is no ommitted variable.
+\includegraphics[width=\textwidth]{figures/estimation_nls.png}
+\caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is no ommitted variable.
 \label{fig:estimation_nls}}
@@
 Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta$ and the latent positions, as well as a GLM.
-Type-1 error rates shown in \ref{fig:inference_type_1} for the LSM are in general lower than those of a GLM when an unmeasured covariate exists.
-Under uniform correlation between the omitted network structure and the observed covariate the Type-1 error rate is still often more than 10 times greater than the nominal rate of .05.
+Type-1 error rates shown in Figure \ref{fig:inference_type_1} for the LSM are in general lower than those of a GLM when an unmeasured covariate exists.
+Under uniform correlation between the omitted network structure and the observed covariate the Type-1 error rate is often more than 10 times greater than the nominal rate of 0.05.
 Again, scaling the prior of the latent space to match that of the observed covariate appears to make the LSM's error rate comparable to that of the GLM.
-Type-2 error rates, shown in \ref{fig:inference_type_2} are also substantially above their nominal rate when there is a uniform distribution on the strength of confounding, though the inflation is not as bad as that of the Type-1 rates.
+Type-2 error rates, shown in Figure \ref{fig:inference_type_2} are also substantially above their nominal rate when there is a uniform distribution on the strength of confounding, though the inflation is not as bad as that of the Type-1 rates.
+These results indicate that, in the presence of unobserved confounders that can be represented through the LSM, the LSM does not adequately correct the inferential errors that would arise due to omitted variables in the conventional regression framework.
 
 \begin{figure}
-\includegraphics[width=\textwidth]{inference_type_1.png}
+\includegraphics[width=\textwidth]{figures/inference_type_1.png}
 \caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y$-axis.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{inference_type_2.png}
+\includegraphics[width=\textwidth]{figures/inference_type_2.png}
 \caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y-axis$.
@@
 
-Generalization error, in this case defined as the expected prediction error on new edges generated using a fixed set of latent positions for the nodes and a fixed observed covariate is shown in figure \ref{fig:generalization}.
-In nearly all cases the LSM substantially outperforms the GLM, especially when the ommitted covariate is not highly collinear with the observed covariate.
+We estimate generalization error as the expected prediction error on new edges generated using a fixed set of latent positions for the nodes and a fixed observed covariate.
+Generalization error results are shown in Figure \ref{fig:generalization}.
+In nearly all cases the LSM substantially outperforms the GLM, especially when the omitted covariate is not highly collinear with the observed covariate.
 In nearly all cases it performs as well as the true model.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{generalization.png}
+\includegraphics[width=\textwidth]{figures/generalization.png}
 \caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed.
@@
 
-We conclude that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space to uncover patterns of interest or the prediction of edges in networks where there is likely to be omitted structure that is not adequately modelled using measured covariates.
+Based on our simulation study, the primary advantage of the LSM is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network.
+Considering both (1) that the LSM does not substantially reduce bias or inferential error relative to the GLM, and (2) that the LSM exhibits strong predictive performance, we infer that the latent positions are fit to explain the systematic variation that cannot be explained through the observed covariates.
+This is of considerable use in research focused on developing a predictive model, or fitting and exploring latent positions.
+However, since the latent positions are inferred to complement the observed covariates---explaining residual variation---the LSM is ill-suited to adjust for confounding network structure, as adjusting for confounding would require that the latent positions explain variation that could otherwise be attributed to the observed covariates.
+
+We conclude that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space model without covariates to explore latent structure in the network through a principled approach to embedding, or the prediction of edges in networks where there is likely to be omitted structure that is not adequately modeled using observed covariates.
 Although inferential and estimation errors are somewhat smaller than that of a GLM when there is omitted network structure, even moderate collinearity between the omitted structure and the measured covariate leads to substantial bias and inferential error.
 The LSM cannot control for unmeasured confounders.
-Additionally, if such network structure does not exist, the LSM may induce additional bias and inferential error, though this effect can be moderated by scaling the prior on the latent positions to match that of the measured covariates.
+Furthermore, if such network structure does not exist, the LSM may induce additional bias and inferential error, though this effect can be moderated by scaling the prior on the latent positions to match that of the measured covariates.
+If the researcher is interested in identifying causal effects of observed covariates, but is concerned that unmeasured variables or network structure could confound the relationship between observed variables and the network, the LSM does not represent an advisable alternative to measuring the confounding variables and/or explicitly modeling the endogenous network structure.
 

### 2016-08-30 manuscript.tex
updated abstract again

@@
 
-\noindent The latent space model (LSM) for network data is a generative probabilistic model that combines a generalized linear model with a latent spatial embedding of the network.
+\noindent In order to account for the complex interdependence found in relational data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference.
+The latent space model (LSM) for network data is a generative probabilistic model that combines a generalized linear model with a latent spatial embedding of the network.
 It has been used to decrease error in the estimation of and inference regarding the effects of observed covariates.

### 2016-08-30 manuscript.tex
updated first sections

--- 2016-07-26 manuscript.tex
@@
 
-\noindent The latent space model (LSM) for network data is a generative probabilistic model that combines a generalized linear model with a latent spatial embedding of the network.
-It has been used to decrease error in the estimation of and inference regarding the effects of observed covariates.
-In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network.
-As far as we know, there has been no research that considers the LSM's performance in adjusting for unmeasured structure to reduce estimation and inferential errors.
-We investigate the LSM's performance via a Monte Carlo study.
-In the presence of an unmeasured covariate that can be appropriately modeled using a latent space, estimation and inferential error remain high under even moderate confounding.
+\noindent Due to the complex interdependence found in relational data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference.
+The latent space model (LSM) for network data combines a generalized linear model with a latent spatial embedding of the network.
+It is assumed that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network.
+There has been little to no research that considers the LSM's performance in adjusting for unmeasured structure.
+We investigate the LSM's performance via a simulation study.
+In the presence of an unmeasured covariate that can be modeled using a latent space, estimation and inferential error remain high under even moderate confounding.
 However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases.
-We conclude that the LSM is most appropriately used for exploratory or predictive tasks.\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219.
+We conclude that the LSM is most appropriate for exploration or prediction.\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219.
 Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}
@@
 The LSM takes a different approach, which involves the incorporation of latent variables to model network structure.
-The LSM has the advantage over ERGM and SAOM in that researchers need not develop a set of hypothesized dependencies in order to model network structure that is not reflected in observed covariates.
-However, this advantage hinges upon the capacity for the LSM to discover unmodeled structure that could otherwise be attributed to the observed covariates (i.e., inferring confounding structure).
-In the current study, we focus on the LSM, examining its performance in reducing estimation and inferential errors regarding the effects of observed covariates, via adjustment for confounding network structure.
+The LSM has an advantage over ERGM and SAOM in that researchers need not develop a set of hypothesized dependencies in order to model network structure that is not reflected in observed covariates.
+However, this advantage hinges upon the capacity for the LSM to discover unmodeled structure that could otherwise be attributed to the observed covariates.
+In the current study, we focus on the LSM, examining its performance in reducing estimation and inferential errors regarding the effects of observed covariates via adjustment for confounding network structure.
 
@@
 There are two reasons that using the LSM may lead to increased error.
-First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spurriously correlated with the observed covariates, leading to reduced efficiency.
-Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable, while the latent space parameters are used to model other sources of variation.
+First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates.
+Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects.
+Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable.
+Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias.
 

### 2016-08-31 manuscript.tex
added yet another quote, from cao2016transnational

@@
 
-Likewise, similar claims are made in \citet{dorff2013networks, cao2013democracies, ward2013gravity, kirkland2012multimember}.
+Likewise, similar claims are made in \citet{dorff2013networks, cao2013democracies, ward2013gravity, kirkland2012multimember}, and \citet{cao2016transnational}.
 
@@
 
-The above examples are drawn from political science, but we see a similar logic for using the LSM articulated in recent work in ecology by \citet[p. 989]{nomano2015unrelated}.
+\begin{quote}
+``Ignoring third-order dependence in dyadic data and treating dyads Germany-France, France-Italy, and Germany-Italy as independent observations can cause bias in parameter estimates (Hoff 2005).
+The statistical literature has proposed a series of latent space to control for autocorrelation among dyadic observations (Hoff et al.
+2002; Hoff 2005).
+Countries?
+unobserved characteristics are captured by latent vectors...'' \citep[p.
+17]{cao2016transnational}.
+\end{quote}
+
+The above examples are drawn from political science, but the results of our analysis have implications in other fields.
+We see a similar logic for using the LSM articulated in recent work in ecology by \citet[p.
+989]{nomano2015unrelated}.
 

### 2016-09-01 manuscript.tex
nearly finished prior section

@@
 
-Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions), which means that the ML estimate is unidentified.
+Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions).
+Invariance to rotation means that the ML estimate is unidentified.
 As such, inference in the LSM is conducted using a Bayesian approach.
 Latent positions and other parameters are sampled using a Metropolis-Hastings algorithm.
-After sampling, latent space positions are rotated to match a common set of positions, via Procrustes transformation \citep{borg2005modern}, in order to eliminate variance due to rotation of the latent space.
+After sampling, latent space positions are rotated to match a common set of positions, via Procrustes transformation \citep{borg2005modern}.
+Procrustes transformation variance that is due to rotation of the latent space.
 \cite{hoff2002latent} propose the use of diffuse independent normal priors for the latent positions and regression parameters.
@@
 In one application they actually just exclude the isolates.
-Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space, but for representative explanatory analyses excluding isolates would amount to selecting on the dependent variable.
+Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space (for, e.g., visualization).
+However, if the goal is to conduct explanatory analysis that is representative of the full set of nodes, excluding isolates would amount to selecting on the dependent variable.
 Unlike in many practical applications of Bayesian inference, it is not possible to use an improper ``flat'' prior \citep{tibshirani1989noninformative} with the LSM, as an informative prior is necessary to assure finite latent positions.

### 2016-09-01 manuscript.tex
finished prior section

@@
 
-To derive our scaling rule, we follow the thought experiment of a one-dimensional latent space $\mathbf{z}$ with a dyadic covariate $\mathbf{x}$.
+Our scaling rule is expressed as the prior variance for the latent positions, denoted $\eta^2$.
+We focus on a one-dimensional latent space $\mathbf{z}$ with a single dyadic covariate $\mathbf{x}$.
+The rule we derive generalizes naturally to more than one dimension, as the prior variance for each dimension could simply be set to $\eta^2/d$ where $d$ is the number of dimensions.
 If we assume that the $\mathbf{z}$ are independent and distributed $\mathcal{N}(0,\sigma^2)$, it is straightforward to derive the distribution of $|z_i - z_j|$.\footnote{The assumption that $\mathbf{z} \sim \mathcal{N}(0,\sigma^2$) is consistent with $\mathbf{z}$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim \mathcal{N}(0,2\sigma^2)$ \citep{devore2012}.
 The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}.
-Following from this result, we know that the variance of  $|z_i - z_j|$ is $$\sigma^2(2-4/\pi).
-$$ Let $\theta^2$ be the empirical variance of $\mathbf{x}$.
+Following from this result, we know that the variance of  $|z_i - z_j|$ is $\sigma^2(2-4/\pi)$.
+Let $\theta^2$ be the empirical variance of $\mathbf{x}$.
 Then the variance in the linear predictor is $\beta^2\theta^2$.
-Given an estimate $\hat{\beta}$, we can set the variance of the latent space prior to a multiple of $\hat{\beta}^2\theta^2/(2-4/\pi)$ in order to tune the prior to avoid replicating the linear predictor via the distances between latent coordinates.
-
+Given an estimate $\hat{\beta}$, we can set the variance of the latent position prior to $$\eta^2=\hat{\beta}^2\theta^2/(2-4/\pi).$$  This prior would actively penalize latent distance distributions that match or exceed the variance of the observed linear predictor.
 \subsection{Simulation Study}

### 2016-09-01 manuscript.tex
changed variance symbol

@@
 
-Our scaling rule is expressed as the prior variance for the latent positions, denoted $\eta^2$.
+Our scaling rule is expressed as the prior variance for the latent positions, denoted $\nu^2$.
 We focus on a one-dimensional latent space $\mathbf{z}$ with a single dyadic covariate $\mathbf{x}$.
-The rule we derive generalizes naturally to more than one dimension, as the prior variance for each dimension could simply be set to $\eta^2/d$ where $d$ is the number of dimensions.
+The rule we derive generalizes naturally to more than one dimension, as the prior variance for each dimension could simply be set to $\nu^2/d$ where $d$ is the number of dimensions.
 If we assume that the $\mathbf{z}$ are independent and distributed $\mathcal{N}(0,\sigma^2)$, it is straightforward to derive the distribution of $|z_i - z_j|$.\footnote{The assumption that $\mathbf{z} \sim \mathcal{N}(0,\sigma^2$) is consistent with $\mathbf{z}$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim \mathcal{N}(0,2\sigma^2)$ \citep{devore2012}.
@@
 Then the variance in the linear predictor is $\beta^2\theta^2$.
-Given an estimate $\hat{\beta}$, we can set the variance of the latent position prior to $$\eta^2=\hat{\beta}^2\theta^2/(2-4/\pi).$$  This prior would actively penalize latent distance distributions that match or exceed the variance of the observed linear predictor.
+Given an estimate $\hat{\beta}$, we can set the variance of the latent position prior to $$\nu^2=\hat{\beta}^2\theta^2/(2-4/\pi).$$  This prior actively penalizes latent distance distributions that match or exceed the variance of the observed linear predictor.
 \subsection{Simulation Study}

### 2016-09-02 manuscript.tex
almost through the simulation design

@@
 
-\subsection{Simulation Setup}
+\subsection{Simulation Design}
 
 For simplicity and computational efficiency we consider a unidimensional latent space.
-We have no reason to suspect that our results should be unique to unidimensional LSMs.
-To generate an observed covariate which has a controllable collinearity with the latent network structure we follow a three-step process.
+We expect our results to generalize to higher dimensional LSMs.
+To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a three-step process.
 First, we simulate unidimensional positions for each node, drawing from a normal distribution, and calculate the Euclidean distance $\mathbf{d}$ between each pair of positions.
@@
 Finally we standardize $x$ to have zero mean and unit variance.
-To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $\mathbf{x}$ we utilize the C-vine method of \cite{lewandowski2009generating}.
+To generate the covariance matrix $\Sigma$, which controls the dependence between the omitted network structure and the observed covariate $\mathbf{x}$ we utilize the C-vine method of \cite{lewandowski2009generating}.
 % fill in a basic description of the method here
 
-We consider three exponential family distributions from which the adjacency matrix entries are drawn: Gaussian, binomial, and Poisson.
-Additionally we control the number of nodes in the network $n = 25, 50, 100$.
+We consider three exponential family distributions from which the adjacency matrix entries are drawn: Gaussian, Bernoilli, and Poisson.
+We also vary the size of the network under study, considering networks with $n = 25, 50,$ and $100$ nodes.
 To vary the degree of confounding attributable to the latent positions, we consider three values of the collinearity parameter $\eta$, $1$, $100$, and $1,000,000$, where $1$ corresponds to a standard uniform distribution of the correlation  (i.e.,  moderate collinearity) between the observed covariate and the latent distances, and $1,000,000$ to independence between the observed covariate and the latent distances.
@@
 The LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R} \citep{latentnet}.
-An initial run of 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler.
-Every 100th iteration is saved.
+For each iteration of the simulation, we run 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler.
+Every 100th post burn in iteration is saved.
 Convergence in the log probability of the model is assessed using the Geweke diagnostic in the \texttt{coda} package \citep{coda, geweke1991evaluating}.

### 2016-09-02 manuscript.tex
working on conclusion

@@
 Convergence in the log probability of the model is assessed using the Geweke diagnostic in the \texttt{coda} package \citep{coda, geweke1991evaluating}.
-If the convergence criterion is satisfied the simulation continues to the next set of arguments, otherwise the number of iterations is doubled.
-If the convergence criteria is still not satisfied, then the aforementioned step in the simulation is flagged for review.
+If the convergence criterion is satisfied, the simulation continues to the next set of arguments, otherwise the number of iterations is doubled.
+If the convergence criterion is still not satisfied, then the aforementioned step in the simulation is flagged for review [ZACH, would you elaborate on this..what did we do with these iterations, and how often did this happen].
 At each point in the simulation's parameter space, we execute 1,000 Monte Carlo iterations.
@@
 For the LSM we use the posterior mode as our point estimate.
-In the cases where edges are binomial and there is omitted network structure, we scale the estimates using the reciprocal of the bias of $\beta$ when $\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 + \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the variance of a standard logistic distribution.
+In the cases where edges are Bernoulli and there is omitted network structure, we scale the estimates using the reciprocal of the bias of $\beta$ when $\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 + \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the variance of a standard logistic distribution.
 We do this to adjust for bias in the coefficient estimate that arises due to the lack of a scale coefficient in logistic regression (See the derivation of the bias under an independent but omitted covariate in \cite{mood2010logistic}).
@@
 Inference regarding Credible intervals for $\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution \cite{turkkan1993computation}.
-We consider $\beta$ to be statistically significant at the 0.05 level if the credible interval does not contain 0.
+We consider $\beta$ to be statistically significant at the 0.05 level if the 95\% credible interval excludes 0.
 
@@
 Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results.
-In Figure \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are treated as observed) show substantial bias.
+In Figure \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are included via an observed covariate) show substantial bias.
 In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$.
-However, in absolute terms the amount of bias is large.
-Our results demonstrate that using the LSM will not permit the discovery of and adjustment for the true latent positions.
+However, in absolute terms, the amount of bias is large.
+Our results demonstrate that using the LSM will not permit the discovery of and adjustment for the true latent positions when the observed covariates and latent distances are confounded.
 For both the GLM and the LSM, bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases.
-In \ref{fig:estimation_nls} we can see that when there is no ommitted network structure, the LSM exhibits bias greater than that of the GLM, though it does appear that scaling the prior of the latent positions to match the scale of the linear predictor decreases this tendency.
+In \ref{fig:estimation_nls} we can see that when there is no omitted network structure, the LSM exhibits bias greater than that of the GLM, though it does appear that scaling the prior of the latent positions to match the scale of the linear predictor decreases this tendency.
 
@@
 Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta$ and the latent positions, as well as a GLM.
-Type-1 error rates shown in Figure \ref{fig:inference_type_1} for the LSM are in general lower than those of a GLM when an unmeasured covariate exists.
-Under uniform correlation between the omitted network structure and the observed covariate the Type-1 error rate is often more than 10 times greater than the nominal rate of 0.05.
-Again, scaling the prior of the latent space to match that of the observed covariate appears to make the LSM's error rate comparable to that of the GLM.
+Type-1 error rates shown in Figure \ref{fig:inference_type_1} for the LSM are in general somewhat lower than those of a GLM when an unmeasured covariate exists.
+Under uniform correlation between the omitted network structure and the observed covariate, the Type-1 error rate is often more than 10 times greater than the nominal rate of 0.05.
+This illustrates that the LSM cannot be considered a viable substitute to measuring the omitted structure, as the true model exhibits a Type-1 error rate that matches the nominal rate.
+Scaling the prior of the latent space to match that of the observed covariate appears to make the LSM's error rate comparable to that of the GLM.
 Type-2 error rates, shown in Figure \ref{fig:inference_type_2} are also substantially above their nominal rate when there is a uniform distribution on the strength of confounding, though the inflation is not as bad as that of the Type-1 rates.
-These results indicate that, in the presence of unobserved confounders that can be represented through the LSM, the LSM does not adequately correct the inferential errors that would arise due to omitted variables in the conventional regression framework.
+These results indicate that, in the presence of unobserved confounders that can be represented through the LSM, the LSM does not adequately correct the inferential errors that would arise due to omitted structure among the dyads in the network.
 
@@
 
+The last property we consider in comparing the GLM and LSM is generalization error---the performance exhibited by the models in predicting new data that was drawn from the same model that generated the data used for estimation (i.e., out of sample predictive performance).
 We estimate generalization error as the expected prediction error on new edges generated using a fixed set of latent positions for the nodes and a fixed observed covariate.
@@
 \caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed.
-In the binomial case the Brier score is computed and in the poisson and normal cases the mean-square-error.
+In the Bernoulli case the Brier score is computed and in the poisson and normal cases the mean-square-error.
 \label{fig:generalization}}
@@
 
-Based on our simulation study, the primary advantage of the LSM is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network.
-Considering both (1) that the LSM does not substantially reduce bias or inferential error relative to the GLM, and (2) that the LSM exhibits strong predictive performance, we infer that the latent positions are fit to explain the systematic variation that cannot be explained through the observed covariates.
+Based on our simulation study, we conclude that the primary advantage of using the LSM---relative to a dyadic regression model in which the network is fit using observed covariates only---is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network.
+Considering both (1) that the LSM does not substantially reduce bias or inferential error relative to the GLM, and (2) that the LSM exhibits strong predictive performance, we conclude that the latent positions are used to explain the systematic variation that cannot be explained through the observed covariates.
 This is of considerable use in research focused on developing a predictive model, or fitting and exploring latent positions.
 However, since the latent positions are inferred to complement the observed covariates---explaining residual variation---the LSM is ill-suited to adjust for confounding network structure, as adjusting for confounding would require that the latent positions explain variation that could otherwise be attributed to the observed covariates.
+
+The key recommendation arising from our results is that researchers not use the LSM as an alternative to measuring potential confounding variables and/or explicitly modeling the endogenous network structure.
+The LSM cannot control for unmeasured confounders.
+We do not argue that this property is unique to LSM.
+Rather, we suspect it is the case with most, if not all, latent variable models (e.g., ideal point models, topic models, .
 

### 2016-09-02 analysis.R
updates to figures

@@ -1,7 +1,7 @@
 pkgs <- c("dplyr", "reshape2", "ggplot2", "stringr", "coda", "scales")
 invisible(lapply(pkgs, library, character.only = TRUE))
 
-results <- read.csv("results_bruce.csv", stringsAsFactors = FALSE) %>%
+results <- read.csv("results.csv", stringsAsFactors = FALSE) %>%
   select(-one_of("X", "prob", "repl", "id")) %>%
   rename(model = algo) %>%
   mutate(model = ifelse(!is.na(scale) & scale, "lsm_scaled", model)) %>%
@@ -16,6 +16,7 @@ results <- read.csv("results_bruce.csv", stringsAsFactors = FALSE) %>%
             coverage = mean(coverage, na.rm = TRUE),
             bias = mean(bias, na.rm = TRUE),
             error = mean(error, na.rm = TRUE))
+results$family[results$family == "binomial"] <- "bernoulli"
 
 ggplot(results, aes(eta, n, color = model, linetype = model)) +
   scale_x_log10(breaks = c(1, 100, 1000000), labels = c("uniform correlation", "low correlation", "independence")) +
@@ -24,7 +25,7 @@ ggplot(results, aes(eta, n, color = model, linetype = model)) +
   facet_grid(family + beta ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = "Iterations Completed") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")
 ggsave("iterations.png", width = 12, height = 8)
 
 ggplot(filter(results, beta == 1), aes(eta, loss, color = model, linetype = model)) +
@@ -33,7 +34,7 @@ ggplot(filter(results, beta == 1), aes(eta, loss, color = model, linetype = mode
   facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = "Mean Squared Error", title = "Generalization Error") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")
 ggsave("generalization.png", width = 12, height = 8)
 
 ggplot(filter(results, beta == 1, latent_space == -1),
[…]

### 2016-09-02 manuscript.tex

@@
 The LSM cannot control for unmeasured confounders.
-We do not argue that this property is unique to LSM.
-Rather, we suspect it is the case with most, if not all, latent variable models (e.g., ideal point models, topic models, .
-
-We conclude that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space model without covariates to explore latent structure in the network through a principled approach to embedding, or the prediction of edges in networks where there is likely to be omitted structure that is not adequately modeled using observed covariates.
-Although inferential and estimation errors are somewhat smaller than that of a GLM when there is omitted network structure, even moderate collinearity between the omitted structure and the measured covariate leads to substantial bias and inferential error.
-The LSM cannot control for unmeasured confounders.
-Furthermore, if such network structure does not exist, the LSM may induce additional bias and inferential error, though this effect can be moderated by scaling the prior on the latent positions to match that of the measured covariates.
-If the researcher is interested in identifying causal effects of observed covariates, but is concerned that unmeasured variables or network structure could confound the relationship between observed variables and the network, the LSM does not represent an advisable alternative to measuring the confounding variables and/or explicitly modeling the endogenous network structure.
+This property is, perhaps, not surprising.
+Given that all priors are centered at zero, the parameter regions that exhibit high posterior probability will be those that explain the observed network (i.e., high likelihood) using minimal departures from zero (i.e., high prior probability).
+Since explaining the observed network using a covariate requires moving one parameter value away from zero, whereas explaining the network using a latent dimension requires moving many latent coordinates away from zero, the LSM exhibits an inherent preference to explain network structure using observed covariates over latent coordinates.
+One way to address this issue would be to use a highly diffuse prior for the latent positions, but as we discuss above, that is not possible with the LSM.
+If the LSM is to be used as a tool for drawing inferences regarding observed covariate effects while adjusting for confounding network structure, further development is needed to enable the LSM to avoid bias and inferential error under confounding.
 

### 2016-09-07 manuscript.tex
update to iterations line

@@
 If the convergence criterion is satisfied, the simulation continues to the next set of arguments, otherwise the number of iterations is doubled.
-If the convergence criterion is still not satisfied, then the aforementioned step in the simulation is flagged for review [ZACH, would you elaborate on this..what did we do with these iterations, and how often did this happen].
+If the convergence criterion is still not satisfied, then the aforementioned step in the simulation is flagged for review.
+However, this was not necessary in any cases.
 At each point in the simulation's parameter space, we execute 1,000 Monte Carlo iterations.

### 2016-09-07 reduce.R

@@ -14,25 +14,30 @@ reduce <- function(job, res) {
       estimate <- res$estimate
       coverage <- unname(ifelse(res$interval[1] < 0 & res$interval[2] > 0, 1, 0))
     }
-    list("estimate" = estimate, "loss" = res$loss, "coverage" = coverage)
+    if (!any(class(res$fit) == "glm")) {
+      iter <- nrow(res$fit$sample$beta)
+    } else {
+      iter <- NA
+    }
+    list("estimate" = estimate, "loss" = res$loss, "coverage" = coverage, "iter" = iter)
   } else {
-    list("estimate" = NA, "loss" = NA, "coverage" = NA)
+    list("estimate" = NA, "loss" = NA, "coverage" = NA, "iter" = NA)
   }
 }
 
 res_reg_open <- batchMapQuick(reduceResultsExperiments,
-                              chunk(findDone(reg_open), n.chunks = 80, shuffle = TRUE),
-                              more.args = list(reg = reg_open, fun = reduce, progressbar = FALSE,
-                                               impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
-                              resources = list(walltime = 86400L / 2, nodes = 1L, memory = "8gb"))
+  chunk(findDone(reg_open), n.chunks = 80, shuffle = TRUE),
+  more.args = list(reg = reg_open, fun = reduce, progressbar = FALSE,
+    impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
+  resources = list(walltime = 86400L / 2, nodes = 1L, memory = "8gb"))
 waitForJobs(res_reg_open)
 res_open <- reduceResults(res_reg_open, fun = function(aggr, job, res) rbind.fill(aggr, res), init = data.frame())
 
 res_reg_bruce <- batchMapQuick(reduceResultsExperiments,
-                               chunk(findDone(reg_bruce), n.chunks = 200, shuffle = TRUE),
-                               more.args = list(reg = reg_bruce, fun = reduce, progressbar = FALSE,
-                                                impute.val = list("estimate" = NA, "loss" = NA, "coverage" = NA)),
-                               resources = list(walltime = 86400L, nodes = 1L, memory = "8gb"))
+  chunk(findDone(reg_bruce), n.chunks = 200, shuffle = TRUE),
+  more.args = list(reg = reg_bruce, fun = reduce, progressbar = FALSE,
[…]

### 2016-09-26 manuscript.tex
updated bibliography

@@
 Political networks scholars are well versed in the risks associated with ignoring unmodeled network structure.
-Dependencies such as reciprocity, transitivity, and homophily---if not accounted for---can lead to biased estimates and errors in hypothesis testing, much in the way that omitted variable bias can affect results in conventional regression models \citep{ward2007disputes,kinne2014,cranmerisq}.
+Dependencies such as reciprocity, transitivity, and homophily---if not accounted for---can lead to biased estimates and errors in hypothesis testing, much in the way that omitted variable bias can affect results in conventional regression models \citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}.
 A number of statistical modeling frameworks have been proposed to account for confounding structure in network data that cannot be modeled with observed covariates.

### 2017-03-10 response.tex
add response memo draft

\section{Response Memo}\label{response-memo}

\subsection{Response to Reviewer 1}\label{response-to-reviewer-1}

\begin{quote}
``The paper does not accurately characterize or analytically understand
the latent space approaches employed by the work that is cited. For
example Dorff and Ward (2013) do not use LATENTNET. LATENTNET is
extremely outdated and is not the software used in much of the other
work cited herein. This undermines much of the criticisms in the papers
and raises a red flag that suggests the authors do not grasp the
modeling approaches as deeply as they would present.''
\end{quote}

While we appreciate that the \texttt{latentnet} package is not the
implementation most commonly used in political science, it is used
outside of political science. We have made this point more clearly in
the revised manuscript. Most importantly, our criticism is not specific
to the implementation in \texttt{latentnet}, and we argue (again, more
clearly in the revised manuscript), that our argument does not depend on
the specific software implementation or parameterization used, and hence
this criticism does not speak to our point, which is that that the LSM
cannot adjust for unobserved confounders, even in a synthetic scenario
designed to make this as easy as possible.

\begin{quote}
``The authors are not up-to-date on the state of latent modeling in the
field and have ignored more recent work (which addresses many of the
concerns raised by the authors) particularly by Peter Hoff and Shahryar
Minhas. Additionally, the progress made by the newly developed AMEN
software is completely overlooked by this article.''
\end{quote}

We have added discussion of the AMEN package, which we were aware of,
but again, does not apply to our argument for the same reason.

\begin{quote}
``In that vein, authors identify a problem (e.g., collinearity between
latent distance parameters and exogenous variables in the latentnet
framework) and do not solve it nor discuss any recent work that tries to
address this problem.''
\end{quote}

Our point is that when the measured covariates are \emph{not}
independent of the unobserved structure in the network, that estimating
that structure using latent variables cannot adjust for any confounding.
We are unaware of any work that attempts to tackle this problem and we
furthermore consider the problem as we've posed it intractable for
theoretical reasons, which we have elucidated more clearly in our
revised version. In generally latent variables cannot be used to adjust
the effects of measured covariates. In the case where the measured
covariates are independent of any unmeasured structure, the LSM does no
worse than standard models, as we show in our simulation.

\subsection{Response to Reviewer 2}\label{response-to-reviewer-2}

\begin{quote}
``Yet, in their efforts to contextualize their research the authors seem
to disregard or not be aware of notable differences between existing
latent variable approaches. Specifically, much of the work they cite
(e.g., Ward Siverson and Cao 2007, Ward, Ahlquist and Rozenas 2013,
etc.) does not even use the type of latent variable model that the
authors focus on in their paper. In fact, with the exception of a 2012
piece by Kirkland, I am not aware of any published political science
work that actually uses the latent space model for network analysis the
authors study here.''
\end{quote}

While this is true, it is orthogonal to our point, which is that the
LSM, regardless of how the latent variables are parameterized, cannot
adjust for unobserved network confounders. We chose the first
formulation of the LSM to make our synthetic demonstration as
understandable as possible. We argue that it shows that even in the most
ideal scenario, the LSM cannot adjust for confounders that are
unmeasured. We have made the point that this behavior is independent of
the parameterization of the latent variables more clear in the revised
manuscript. While we certainly could have chosen any number of the
variations of the LSM for our simulation, absent clear reasons to do so
we do not believe that it is reasonable to expend the substantial
computational resources necessary to expand the simulation.

\begin{quote}
``The difference between these approaches has been shown to be
consequential in a number of works. Hoff \& Ward (2004) and Hoff (2005)
detail the ways in which the GBME is different than the latent distance
model that is employed in the latentnet package. Hoff (2008) explicitly
discusses how the latent distance model confounds transitivity and
stochastic equivalence (two types of third order interdependencies) in
problematic ways that the latent factor model is able to avoid. The
issue this paper seems most concerned with showing is that the latent
distance model does not solve the problem of bias and inferential error
relative to GLM for observed covariates. They do so through a well
executed simulation exercise and the results are certainly compelling in
favor of showcasing a shortcoming of the latent distance model. At the
same time, this result is not very surprising. For the latent distance
model we expect there to be some collinearity between covariates for
which there is homophily and latent position. In theory, a weak prior on
the covariate coefficients and a strong (tight) prior on the
coefficients should minimize this effect, but at the cost of less
expressiveness for latent positions.''
\end{quote}

We agree that the parameterization of the latent variables can affect
what can be learned about omitted structure in the network, as
suggested. However, this is unrelated to our point which is that it is
not possible to adjust the estimates of the effect of observed
covariates using a latent variable model that can represent unobserved
structure. We argue more clearly in the revised manuscript that this is
not a property of the Euclidean distance formulation we chose for
expository simplicity.

### 2017-03-16 manuscript.tex
finished response to R1

@@
 The distance function is additively combined with a regression on observed dyadic covariates, $x_{ij}$, to form a linear predictor for tie prediction.
-As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution.
+As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution, giving
 
-$$\mathbb{E}(y_{ij} | x_{ij}) = g^{-1}(\alpha + \beta x_{ij} - |z_i - z_j|)$$
+$$\mathbb{E}(y_{ij} | x_{ij}) = g^{-1}(\alpha + \beta x_{ij} - |z_i - z_j|).$$
 
-However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured, are correlated with measured covariates.
-There are two reasons that using the LSM may lead to increased error.
+This ``Euclidean'' LSM is the original form proposed by \citet{hoff2002latent}, and, as far as we can tell, the most commonly used specification in the literature.
+However, the LSM has been extended in several important directions.
+The  The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors with a multiplicative metric \citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa, hoff2009multiplicative}.
+Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}.
+Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process.
+\cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models.
+\cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
+\citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R.
+
+Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure.
+There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and other Type 2.
 First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates.
-Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects.
+Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects, and result in a high Type 2 error rate.
 Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable.
-Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias.
+Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias, which leads to an inflated Type 1 inferential error rate.
+
+Recent work on the LSM has explicitly addressed the potential problem of Type 2 error.
+The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates.
+\citet[p.
+43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp.
+12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error.
+However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks.
+Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.
 
@@
 
-The LSM has seen a substantial amount of subsequent development and extension.
-The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors \citep{hoff2002latent, hoff2009multiplicative}.
-Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}.
-Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process.
-\cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models.
-\cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
  Most of these developments have represented new forms for the latent space and/or the further development of the LSM as a Bayesian hierarchical model.

### 2017-03-17 manuscript.tex
revised to remove prior discussion and generalize beyond Euclidean

@@
 
-\section{Applications and Development of the Latent Space Model}
+\section{Applications of the Latent Space Model}
 
@@
 It has been used to describe topic-specific patterns of interaction in e-mail communication networks \citep{krafft2012topic}.
+It has been used to estimate the effects of education policy interventions on the structure of friendship networks among students \citep{sweet2011modeling}.
 Lastly, in neuroscience it has been proposed as a method for modelling fMRI data \citep{simpson2013analyzing}.
@@
 
-Although the LSM is arguably most useful as an exploratory or predictive model (see \citet{shmueli2010explain} for a discussion of the differences between predictive and explanatory modeling), it has been applied in some cases to reduce estimation and/or inferential error with respect to the effects of observed covariates.
+Although the LSM has often been used as an exploratory or predictive model, it has been applied in many cases for explanatory causal modeling---to reduce estimation and/or inferential error with respect to the effects of observed covariates.
 For example \citet{ward2007disputes}, in a high profile example, argue that the the LSM improves inference about the effects of democracy, international trade, and participation in international organizations on the probability of inter-state conflict.
@@
 
-If the LSM does reduce inferential/estimation error under certain conditions, it will serve as a valuable general model for explanatory analysis of network data, especially since its use does not require the researcher to specify a set of network dependencies from theory.
-However, as far as we know, there has been no research into the performance of the LSM in adjusting for confounding network structure.
-Additionally, the predictive performance of the LSM has only seen limited evaluation \citep{hoff2002latent}, despite having been used for model selection \citep{ward2013gravity, fletcher2011social, fletcher2013network, chiu2011unifying}.
-Hence, finding the conditions under which the LSM reduces prediction error also may affect future use of the LSM.
+The latent space framework offers an attractive general purpose approach to adjusting for confounding structure in network models, as, unlike the major alternative framework for network modeling (ERGM), using the latent space model does not require the researcher to specify a set of network dependencies for which to control.
+However,  the latent variables in the LSM are free parameters that will be inferred to explain variation that is not explained by the observed covariates.
+As such, the LSM framework is not designed to encourage latent network structure to displace observed covariate effects when the latent structure includes confounders.
+As we note above, recent work on latent space approaches for network modeling describes exactly this modeling characteristic---that the latent variables are used to explain variation that cannot be explained by observed covariates.
+In a simulation study that follows, we examine whether the LSM can be used to adjust for confounding in a simple and ideal scenario..
 
- Most of these developments have represented new forms for the latent space and/or the further development of the LSM as a Bayesian hierarchical model.
-In the next section we propose an approach to specifying the Bayesian prior in order to improve inference regarding covariate parameters.
+%\section{Bayesian Inference and Priors in the Latent Space Model}
 
-\section{Bayesian Inference and Priors in the Latent Space Model}
-
-Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions).
+%Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions).
 Invariance to rotation means that the ML estimate is unidentified.
@@
 %\end{quote}
-Since it is necessary to use an informative prior with the LSM, we ask whether it is possible to specify the prior in a way that improves inference regarding covariate parameters.
+%Since it is necessary to use an informative prior with the LSM, we ask whether it is possible to specify the prior in a way that improves inference regarding covariate parameters.
 We propose to calibrate the prior to be on a scale comparable to the linear predictor estimated from a GLM.
@@
 
-Our scaling rule is expressed as the prior variance for the latent positions, denoted $\nu^2$.
+%Our scaling rule is expressed as the prior variance for the latent positions, denoted $\nu^2$.
 We focus on a one-dimensional latent space $\mathbf{z}$ with a single dyadic covariate $\mathbf{x}$.
@@
 In the simulation design, we study how the level of confounding affects the performance of the LSM relative to the GLM.
-We evaluate performance in terms of bias, inferential error, and prediction.
+We evaluate performance in terms of bias and inferential error.
+Since the LSM has also been used for predictive model selection, we also evaluate predictive performance  \citep{ward2013gravity, fletcher2011social, fletcher2013network, chiu2011unifying}.
 
@@
 For simplicity and computational efficiency we consider a unidimensional latent space.
-We expect our results to generalize to higher dimensional LSMs.
+We expect our results to generalize to higher dimensions and other variants of the LSM since each variant involves the use of unconstrained latent variables to represent unmeasured network structure.
 To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a three-step process.

### 2017-03-17 manuscript.tex
removed all mention of prior scaling

@@
 We consider the LSM with several different priors on the coefficient for the observed covariate and the latent space.
-We set the prior variance of $\beta$ to be either $1$ or $10$ and alternatively use a diffuse normal prior on the latent space or scale it by $\hat{\beta}^2\theta^2/(2-4/\pi)$, where $\theta^2$ is the empirical variance of $x$.
+We set the prior variance of $\beta$ to be either $1$ or $10$ and use a diffuse normal prior on the latent space.
 
@@
 For both the GLM and the LSM, bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases.
-In \ref{fig:estimation_nls} we can see that when there is no omitted network structure, the LSM exhibits bias greater than that of the GLM, though it does appear that scaling the prior of the latent positions to match the scale of the linear predictor decreases this tendency.
+In \ref{fig:estimation_nls} we can see that when there is no omitted network structure, the LSM exhibits bias greater than that of the GLM.
 
@@
 
-Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta$ and the latent positions, as well as a GLM.
+Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta$, as well as a GLM.
 Type-1 error rates shown in Figure \ref{fig:inference_type_1} for the LSM are in general somewhat lower than those of a GLM when an unmeasured covariate exists.
@@
 This illustrates that the LSM cannot be considered a viable substitute to measuring the omitted structure, as the true model exhibits a Type-1 error rate that matches the nominal rate.
-Scaling the prior of the latent space to match that of the observed covariate appears to make the LSM's error rate comparable to that of the GLM.
 Type-2 error rates, shown in Figure \ref{fig:inference_type_2} are also substantially above their nominal rate when there is a uniform distribution on the strength of confounding, though the inflation is not as bad as that of the Type-1 rates.

### 2017-04-03 src/simulation.R
implementation of bilinear dgp w/ latentnet

@@ -1,11 +1,8 @@
-library("BatchExperiments")
-library("latentnet")
-library("foreach")
-library("clusterGeneration")
-library("iterators")
+pkgs <- c("BatchExperiments", "latentnet", "amen", "foreach", "clusterGeneration",
+  "iterators", "network", "mvtnorm", "coda")
+invisible(lapply(pkgs, library, character.only = TRUE))
 
 create_network <- function(y, name = "eweight") {
-  library("network")
   net <- network(y, directed = FALSE)
   el <- as.matrix(net, "edgelist")
   w <- y[el]
@@ -14,8 +11,6 @@ create_network <- function(y, name = "eweight") {
 }
 
 rmvnorm_d <- function(sigma, d) {
-  library("mvtnorm")
-  
   var_d <- var(d)
   mu_d <- mean(d)
   n_var <- ncol(sigma)
@@ -58,10 +53,15 @@ diagnostic <- function(fit) {
 }
 
 create_data <- function(static, family, eta, nodes, beta, latent_space) {
-  library("clusterGeneration")
-
   C <- abs(cov2cor(genPositiveDefMat(static$p, "c-vine", eta = eta)$Sigma))
-  d <- as.matrix(dist(replicate(static$p, rnorm(nodes, 0, static$sigma))))
+  if (static$billinear) {
+    u <- rnorm(nodes, 0, static$sigma)
+    v <- rnorm(nodes, 0, static$sigma)
+    d <- u %o% v
+  } else {
+    d <- as.matrix(dist(replicate(static$p, rnorm(nodes, 0, static$sigma))))
+  }
+
   d <- d[lower.tri(d)]
   d <- d / (sd(d) / sqrt(diag(C)[nrow(C)]))
   X <- rmvnorm_d(C, d)
@@ -91,42 +91,55 @@ create_data <- function(static, family, eta, nodes, beta, latent_space) {
     family = family,
     nodes = nodes,
     beta = beta,
-    latent_space = latent_space
+    latent_space = latent_space,
+    bilinear = bilinear
   )
   ret
 }
 
-lsm.wrapper <- function(static, dynamic, ...) {
-  library("latentnet")
-  library("coda")
-
+amen.wrapper <- function(static, dynamic, ...) {
   dots <- list(...)
+  lf <- c("gaussian" = "nrm", "binomial" = "bin", "poisson" = "Poisson")
 
+  fit <- ame(dynamic$y, dynamic$X, R = 2,
[…]

### 2017-04-03 response.tex
updated response memo

@@
 
+We updated the Problem section of the manuscript to represent, (1) that we know about extensions to the LSM, (2) to be clear that we think the problem affects all versions of the LSM, and (3) to quote from AMEN to echo the problem we see with using LSM as a method of adjusting for confounders.
+We cannot possibly analyze all of the individual variants of the LSM, but our point is more broadly directed at the use of latent variables to adjust for confounding network dependencies.
+
 \begin{quote}
@@
 \end{quote}
+
+We added discussion of the way in which AMEN addresses the problem of correlation between the observed covariates and the latent factors.
+We point out, however, that the AMEN approach is designed to address Type 2 error by prioritizing the observed covariates and using the latent factors to explain the variation in the residuals.
+This does not address Type 1 error (which is more troubling than Type 2), and probably makes Type 1 worse.
 

### 2017-04-04 src/simulation.R
setup for amen component

@@ -40,6 +40,8 @@ unstack_vector <- function(x, n) {
   mat
 }
 
+stack_matrix <- function(x) x[lower.tri(x)]
+
 diagnostic <- function(fit) {
   stopifnot(any(class(fit) %in% "glm"))
   SVD <- svd(scale(solve(vcov(fit)), center = FALSE))
@@ -56,8 +58,8 @@ create_data <- function(static, family, eta, nodes, beta, latent_space) {
   C <- abs(cov2cor(genPositiveDefMat(static$p, "c-vine", eta = eta)$Sigma))
   if (static$billinear) {
     u <- rnorm(nodes, 0, static$sigma)
-    v <- rnorm(nodes, 0, static$sigma)
-    d <- u %o% v
+    ## v <- rnorm(nodes, 0, static$sigma)
+    d <- u %o% u ## symmetric
   } else {
     d <- as.matrix(dist(replicate(static$p, rnorm(nodes, 0, static$sigma))))
   }
@@ -99,12 +101,41 @@ create_data <- function(static, family, eta, nodes, beta, latent_space) {
 
 amen.wrapper <- function(static, dynamic, ...) {
   dots <- list(...)
-  lf <- c("gaussian" = "nrm", "binomial" = "bin", "poisson" = "Poisson")
+  lf <- c("gaussian" = "nrm", "binomial" = "bin")
 
-  fit <- ame(dynamic$y, dynamic$X, R = 2,
-    rvar = FALSE, cvar = FALSE, dcor = FALSE, nvar = FALSE,
-    symmetric = TRUE, plot = FALSE, model = "nrm", seed = static$seed, print = FALSE,
-    nscan = static$ame_mcmc)
+  flag <- TRUE
+  iter <- 1
+  m <- 200
+
+  while (flag) {
+    fit <- ame(dynamic$y, dynamic$X, R = 1,
+      rvar = FALSE, cvar = FALSE, dcor = FALSE, nvar = FALSE,
+      symmetric = TRUE, plot = FALSE, model = unname(lf[names(lf) %in% dynamic$family]),
+      seed = static$seed, print = FALSE,
+      nscan = static$ame_mcmc, odens = static$thin)
+
+    raft <- as.numeric(max(raftery.diag(as.mcmc(fit$BETA))$resmatrix[, 2]))
+    if (m > raft | iter == static$maxit) {
+      flag <- FALSE
+    } else {
+      iter <- iter + 1
[…]

### 2017-04-05 src/.BatchJobs.R
batchtools rewrite

@@ -1,1 +1,0 @@
-cluster.functions = makeClusterFunctionsTorque("template.tmpl")

### 2017-04-05 src/simulation.R

@@ -1,4 +1,4 @@
-pkgs <- c("BatchExperiments", "latentnet", "amen", "foreach", "clusterGeneration",
+pkgs <- c("batchtools", "latentnet", "amen", "foreach", "clusterGeneration",
   "iterators", "network", "mvtnorm", "coda")
 invisible(lapply(pkgs, library, character.only = TRUE))
 
@@ -54,14 +54,14 @@ diagnostic <- function(fit) {
   round(out, 3)
 }
 
-create_data <- function(static, family, eta, nodes, beta, latent_space) {
-  C <- abs(cov2cor(genPositiveDefMat(static$p, "c-vine", eta = eta)$Sigma))
-  if (static$billinear) {
-    u <- rnorm(nodes, 0, static$sigma)
-    ## v <- rnorm(nodes, 0, static$sigma)
+create_data <- function(data, job, p, eta, nodes, family, beta, latent_space, ...) {
+  C <- abs(cov2cor(genPositiveDefMat(data$p, "c-vine", eta = eta)$Sigma))
+  if (data$bilinear) {
+    u <- rnorm(nodes, 0, data$sigma)
+    ## v <- rnorm(nodes, 0, data$sigma)
     d <- u %o% u ## symmetric
   } else {
-    d <- as.matrix(dist(replicate(static$p, rnorm(nodes, 0, static$sigma))))
+    d <- as.matrix(dist(replicate(data$p, rnorm(nodes, 0, data$sigma))))
   }
 
   d <- d[lower.tri(d)]
@@ -70,14 +70,14 @@ create_data <- function(static, family, eta, nodes, beta, latent_space) {
   X <- (X - colMeans(X)) / apply(X, 2, sd)
   lp <- beta * X + latent_space * d
   y_stack <- switch(family,
-                    gaussian = rnorm(length(lp), lp, static$sigma),
+                    gaussian = rnorm(length(lp), lp, data$sigma),
                     binomial = rbinom(length(lp), 1, 1 / (1 + exp(-(lp)))),
                     poisson = rpois(length(lp), exp(lp)))
   y <- unstack_vector(y_stack, nodes)
   X_stack <- data.frame(X)
   X <- unstack_vector(X, nodes)
   y_stack_new <- switch(family,
-                        gaussian = rnorm(length(lp), lp, static$sigma),
+                        gaussian = rnorm(length(lp), lp, data$sigma),
[…]

### 2017-04-05 src/submit.R

@@ -1,7 +1,9 @@
-library("BatchExperiments")
+library("batchtools")
 
 ## load registry defined in simulation.R
-reg <- loadRegistry("reg")
+reg <- loadRegistry("lsm")
+reg$cluster.functions = makeClusterFunctionsTORQUE("template.tmpl")
+
 ## per-job resources for torque cluster with 80 cores
 resources <- list(walltime = 86400L * 1, nodes = 1L, memory = "8gb")
 

### 2017-05-02 src/analysis_amen.R
bugfixes and some selective submission code

@@ -3,10 +3,11 @@ invisible(lapply(pkgs, library, character.only = TRUE))
 
 results = fread("results_amen.csv") %>%
   filter(eta != .1 & !(algorithm == "truth" & latent_space == 0) &
-           algorithm != "lsm" & !(latent_space == -1)) %>%
+    algorithm != "lsm" & !(latent_space == -1) & family != "poisson" & nodes != 50) %>%
   mutate(bias = estimate - beta,
     error = ifelse(beta == 0, 1 - coverage, coverage),
-    type = ifelse(beta == 0, 1, 2)) %>%
+    type = ifelse(beta == 0, 1, 2),
+    family = ifelse(family == "binomial", "bernoulli", family)) %>%
   group_by(algorithm, family, eta, beta, type, latent_space, nodes) %>%
   summarise(n = length(estimate[!is.na(estimate)]),
             estimate = mean(estimate, na.rm = TRUE),
@@ -14,7 +15,6 @@ results = fread("results_amen.csv") %>%
             coverage = mean(coverage, na.rm = TRUE),
             bias = mean(bias, na.rm = TRUE),
             error = mean(error, na.rm = TRUE))
-results$family[results$family == "binomial"] = "bernoulli"
 
 ggplot(results, aes(eta, n, color = algorithm, linetype = algorithm)) +
   scale_x_log10(breaks = c(1, 100, 1000000), labels = c("uniform correlation", "low correlation", "independence")) +

### 2017-05-02 src/simulation.R

@@ -2,6 +2,12 @@ pkgs = c("batchtools", "latentnet", "amen", "foreach", "clusterGeneration",
   "iterators", "network", "mvtnorm", "coda")
 invisible(lapply(pkgs, library, character.only = TRUE))
 
+predict.ame = function(object, instance) {
+  ret = colMeans(object$BETA %*% t(cbind(1, instance$X_stack))) +
+    stack_matrix(object$ULUPM)
+  if (instance$family == "binomial") pnorm(ret) else ret
+}
+
 create_network = function(y, name = "eweight") {
   net = network(y, directed = FALSE)
   el = as.matrix(net, "edgelist")
@@ -24,12 +30,7 @@ rmvnorm_d = function(sigma, d) {
     t(apply(mu_x, 2, function(mu) rmvnorm(1, mu, sigma_x)))
 }
 
-ms_error = function(x, y, each_column = FALSE) {
-  if (each_column)
-    sapply((x - y)^2, mean)
-  else
-    mean((x - y)^2)
-}
+ms_error = function(x, y) mean((x - y)^2)
 
 unstack_vector = function(x, n) {
   idx = t(combn(1:n, 2))
@@ -70,16 +71,16 @@ create_data = function(data, job, p, eta, nodes, family, beta, latent_space, ...
   X = (X - colMeans(X)) / apply(X, 2, sd)
   lp = beta * X + latent_space * d
   y_stack = switch(family,
-                    gaussian = rnorm(length(lp), lp, data$sigma),
-                    binomial = rbinom(length(lp), 1, 1 / (1 + exp(-(lp)))),
-                    poisson = rpois(length(lp), exp(lp)))
+    gaussian = rnorm(length(lp), lp, data$sigma),
+    binomial = rbinom(length(lp), 1, 1 / (1 + exp(-(lp)))),
+    poisson = rpois(length(lp), exp(lp)))
   y = unstack_vector(y_stack, nodes)
   X_stack = data.frame(X)
   X = unstack_vector(X, nodes)
   y_stack_new = switch(family,
-                        gaussian = rnorm(length(lp), lp, data$sigma),
-                        binomial = rbinom(length(lp), 1, 1 / (1 + exp(-(lp)))),
-                        poisson = rpois(length(lp), exp(lp)))
+    gaussian = rnorm(length(lp), lp, data$sigma),
[…]

### 2017-05-02 src/submit.R

@@ -11,11 +11,17 @@ resources = list(walltime = 86400L * 1, nodes = 1L, memory = "8gb")
 ## find ids for jobs not executed or on system, chunk them, and send them
 ## to the scheduler
 ids = getJobTable()[, .(job.id, problem, algorithm, nodes)]
-exclusions = rbind(findOnSystem(), findDone())
-ids[, chunk := chunk(job.id, chunk.size = 100), by = "problem"]
-ids[, chunk := .GRP, by = c("problem", "chunk")]
-chunks.to.submit = sort(unique(ids[ids$problem == "bilinear.test", chunk]))
-submitJobs(ids[ids$chunk %in% chunks.to.submit & !ids$job.id %in% exclusions$job.id & ids$nodes != 50, ], resources = resources)
+exclusions = rbind(findOnSystem(), findDone(), findStarted())
+ids = ids[!job.id %in% exclusions$job.id, ]
+amen.experiments = findExperiments("bilinear.test", algo.name = "amen",
+  ids = ids[!job.id %in% exclusions$job.id, ])
+## amen.pars = getJobPars(amen.experiments)
+## amen.to.time = amen.pars[, .SD[1, ], by = nodes]$job.id
+## amen.timing = sapply(amen.to.time, function(x) system.time(testJob(x)))
+## amen.chunksize = sapply(((24 * 60 * 60) / amen.timing) * .75, floor)
+amen.experiments[, chunk := chunk(job.id, chunk.size = 25)]
+
+submitJobs(amen.experiments[chunk < 100], resources = resources)
 getStatus(ids[ids$nodes != 50 & ids$problem == "bilinear.test", "job.id"])
 
 reducer = function(x) {
@@ -27,7 +33,9 @@ reducer = function(x) {
     ret = tryCatch(list("estimate" = x$estimate * adjustment), error = function(e) e)
     ret$coverage = tryCatch(unname(ifelse(x$interval[1] * adjustment < 0 &
       x$interval[2] * adjustment > 0, 1, 0)), error = function(e) e)
-    ret$loss = tryCatch(x$loss, error = function(e) e)
+    ret$loss = tryCatch(ifelse(all(class(x$fit) == "ame"),
+      ms_error(unstack_vector(predict.ame(x$fit, x$data), x$data$nodes), x$data$y_new),
+      x$loss), error = function(e) e)
   }
   return(ret)
 }
@@ -36,13 +44,17 @@ done = findDone()
[…]

### 2017-05-03 manuscript.tex
amen added to manuscript

@@
 \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
-\citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R.
+\citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R \citep{minhas2016inferential, hoff2015dyadic}.
 
@@
 However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks.
-Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.
+Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM as implemented in the R package \texttt{latentnet} as well as \texttt{amen} performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.
 
@@
 
+We also repeat this simulation study using \texttt{amen} \citep{hoff2015dyadic, minhas2016inferential}.
+We omit the poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes.
+Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor.
+We take 100,000 MCMC samples after x burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.
+
 \subsection{Results}
@@
 
+In our repetition of this simulation with \texttt{amen} we find similar results, with the exception of type-2 error rates, which, as expected, are substantially lower for \texttt{amen}.
+However, type-1 error rates remain elevated, as is bias in the case where there is a latent dependence structure that is correlated with the observed covariate.
+\texttt{amen} does perform better in the case where it is used despite the absence of such a latent dependence structure though, relative to \texttt{latentnet}, which is to be expected given that it outperforms \texttt{latentnet} in terms of type-2 error.
+Again, \texttt{amen} performs well in terms of generalization error, coming close to the theoretical lower bound.
+
 \section{Conclusion}
 
-Based on our simulation study, we conclude that the primary advantage of using the LSM---relative to a dyadic regression model in which the network is fit using observed covariates only---is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network.
+Based on our simulation study, we conclude that the primary advantage of using the LSM---relative to a dyadic regression model in which the network is fit using observed covariates only---is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network; \texttt{amen} increases this advantage.
 Considering both (1) that the LSM does not substantially reduce bias or inferential error relative to the GLM, and (2) that the LSM exhibits strong predictive performance, we conclude that the latent positions are used to explain the systematic variation that cannot be explained through the observed covariates.

### 2017-05-06 manuscript.tex

@@
 Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure.
-There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and other Type 2.
+We do not claim that the methodological literature in which the LSM was introduced or extended has presented the LSM as capable of adjusting for unmeasured confounding structure.
+However, as we document below, many researchers who have applied the LSM claim that it has the capacity to adjust for confounders, and use it for this purpose.
+There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and the other in Type 2 error.
 First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates.

### 2017-05-06 manuscript.tex
added online appendix

@@
 Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor.
-We take 100,000 MCMC samples after x burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.
+We take 100,000 MCMC samples after 500 burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.
 
@@
 Again, \texttt{amen} performs well in terms of generalization error, coming close to the theoretical lower bound.
+The results of our simulations with \texttt{amen} are available in the Online Appendix.
 
@@
 
+\newpage
+\section*{Online Appendix}
+Below we replicate the simulation study results figures using the results from our simulation with the recently developed \texttt{amen} package.
+Note that these simulations exclude (1) the Poisson family, which is not available in the \texttt{amen} package, and (2) the 50-node networks, which we left out in the interest of lessening the computational burden of the simulation.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/estimation_ls.png}
+\caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is an ommitted variable.
+The $x$-axis gives the value of the parameter $\eta$ which controls the degree of dependence between $\mathbf{x}$ and ommitted covariate.
+Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate.
+The $y$-axis gives a Monte Carlo estimate of the bias.
+The number of nodes are indicated in the top panels, while the distributional family of the edges is shown on the right panel.
+Each panel represents 4 values of $\eta$ with 1,000 Monte Carlo iterations executed at each point.
+\label{fig:estimation_ls}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/estimation_nls.png}
+\caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is no ommitted variable.
+\label{fig:estimation_nls}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/inference_type_1.png}
+\caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y$-axis.
+Here, $\beta = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta = 0$ is falsely rejected.
+\label{fig:inference_type_1}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/inference_type_2.png}
+\caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y-axis$.
+Here, $\beta = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis.
+\label{fig:inference_type_2}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/generalization.png}
+\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed.
+In the Bernoulli case the Brier score is computed and in the poisson and normal cases the mean-square-error.
+\label{fig:generalization}}
+\end{figure}
+
 \end{document}

### 2017-05-07 manuscript.tex
making final edits

@@
 
-\noindent Due to the complex interdependence found in relational data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference.
+\noindent Due to the complex interdependence found in network data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference.
 The latent space model (LSM) for network data combines a generalized linear model with a latent spatial embedding of the network.
-It is assumed that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network.
+In numerous applications, researchers assume that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network.
 There has been little to no research that considers the LSM's performance in adjusting for unmeasured structure.
@@
 However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases.
-We conclude that the LSM is most appropriate for exploration or prediction.\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219.
+We conclude that the LSM is most appropriate for exploration or prediction.%\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219.
 Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}
@@
 However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks.
-Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM as implemented in the R package \texttt{latentnet} as well as \texttt{amen} performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.
+Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounding structure.
+Through a simulation study we find that the LSM, as implemented in both the R \citep{R} packages \texttt{latentnet} \citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.
 
@@
 We also repeat this simulation study using \texttt{amen} \citep{hoff2015dyadic, minhas2016inferential}.
-We omit the poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes.
+We omit the Poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes.
 Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor.
@@
 Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results.
-In Figure \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are included via an observed covariate) show substantial bias.
+In Figure \ref{fig:estimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are included via an observed covariate) show substantial bias.
 In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$.
@@
 \caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed.
-In the Bernoulli case the Brier score is computed and in the poisson and normal cases the mean-square-error.
+In the Bernoulli case the Brier score is computed and in the Poisson and normal cases the mean-square-error.
 \label{fig:generalization}}
@@
 \texttt{amen} does perform better in the case where it is used despite the absence of such a latent dependence structure though, relative to \texttt{latentnet}, which is to be expected given that it outperforms \texttt{latentnet} in terms of type-2 error.
-Again, \texttt{amen} performs well in terms of generalization error, coming close to the theoretical lower bound.
+Like the LSM estimated with \texttt{latentnet}, \texttt{amen} performs well in terms of generalization error.
 The results of our simulations with \texttt{amen} are available in the Online Appendix.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/estimation_ls.png}
+\includegraphics[width=\textwidth]{figures/estimation_ls_amen.png}
 \caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is an ommitted variable.
@@
 Each panel represents 4 values of $\eta$ with 1,000 Monte Carlo iterations executed at each point.
-\label{fig:estimation_ls}}
+\label{fig:estimation_ls_amen}}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/estimation_nls.png}
+\includegraphics[width=\textwidth]{figures/estimation_nls_amen.png}
 \caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is no ommitted variable.
-\label{fig:estimation_nls}}
+\label{fig:estimation_nls_amen}}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/inference_type_1.png}
+\includegraphics[width=\textwidth]{figures/inference_type_1_amen.png}
 \caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y$-axis.
 Here, $\beta = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta = 0$ is falsely rejected.
-\label{fig:inference_type_1}}
+\label{fig:inference_type_1_amen}}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/inference_type_2.png}
+\includegraphics[width=\textwidth]{figures/inference_type_2_amen.png}
 \caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y-axis$.
 Here, $\beta = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis.
-\label{fig:inference_type_2}}
+\label{fig:inference_type_2_amen}}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/generalization.png}
+\includegraphics[width=\textwidth]{figures/generalization_amen.png}
 \caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed.
-In the Bernoulli case the Brier score is computed and in the poisson and normal cases the mean-square-error.
-\label{fig:generalization}}
+In the Bernoulli case the Brier score is computed and in the normal case the mean-square-error.
+\label{fig:generalization_amen}}
 \end{figure}

### 2017-05-08 src/analysis_amen.R
change font size and title position

@@ -23,8 +23,9 @@ ggplot(results, aes(eta, n, color = algorithm, linetype = algorithm)) +
   facet_grid(family + beta ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = "Iterations Completed") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")
-ggsave("iterations_amen.png", width = 12, height = 8)
+  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
+    plot.title = element_text(hjust = 0.5))
+ggsave("iterations_amen.png", width = 12, height = 8, scale = .75)
 
 ggplot(filter(results, beta == 1), aes(eta, loss,
   color = algorithm, linetype = algorithm)) +
@@ -33,8 +34,9 @@ ggplot(filter(results, beta == 1), aes(eta, loss,
   facet_grid(family ~ nodes + latent_space, scales = "free", labeller = label_context) +
   labs(x = expression(eta), y = "Mean Squared Error", title = "Generalization Error") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")
-ggsave("generalization_amen.png", width = 12, height = 8)
+  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
+    plot.title = element_text(hjust = 0.5))
+ggsave("generalization_amen.png", width = 12, height = 8, scale = .75)
 
 ggplot(filter(results, beta == 1, latent_space == 1),
   aes(eta, bias, color = algorithm, linetype = algorithm)) +
@@ -44,8 +46,9 @@ ggplot(filter(results, beta == 1, latent_space == 1),
   geom_hline(aes(yintercept = 0), linetype = "dashed") +
   labs(x = expression(eta), y = "Bias", title = "Estimation Error w/ Latent Space") +
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")
-ggsave("estimation_ls_amen.png", width = 10, height = 8)
+  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
+    plot.title = element_text(hjust = 0.5))
+ggsave("estimation_ls_amen.png", width = 10, height = 8, scale = .75)
 
[…]

### 2017-05-08 src/analysis_amen.R
more figure scaling

@@ -25,7 +25,7 @@ ggplot(results, aes(eta, n, color = algorithm, linetype = algorithm)) +
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
     plot.title = element_text(hjust = 0.5))
-ggsave("iterations_amen.png", width = 12, height = 8, scale = .75)
+ggsave("iterations_amen.png", width = 12, height = 8, scale = .65)
 
 ggplot(filter(results, beta == 1), aes(eta, loss,
   color = algorithm, linetype = algorithm)) +
@@ -36,7 +36,7 @@ ggplot(filter(results, beta == 1), aes(eta, loss,
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
     plot.title = element_text(hjust = 0.5))
-ggsave("generalization_amen.png", width = 12, height = 8, scale = .75)
+ggsave("generalization_amen.png", width = 12, height = 8, scale = .65)
 
 ggplot(filter(results, beta == 1, latent_space == 1),
   aes(eta, bias, color = algorithm, linetype = algorithm)) +
@@ -48,7 +48,7 @@ ggplot(filter(results, beta == 1, latent_space == 1),
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
     plot.title = element_text(hjust = 0.5))
-ggsave("estimation_ls_amen.png", width = 10, height = 8, scale = .75)
+ggsave("estimation_ls_amen.png", width = 10, height = 8, scale = .65)
 
 ggplot(filter(results, beta == 1, latent_space == 0),
   aes(eta, bias, color = algorithm, linetype = algorithm)) +
@@ -60,7 +60,7 @@ ggplot(filter(results, beta == 1, latent_space == 0),
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom",
     plot.title = element_text(hjust = 0.5))
-ggsave("estimation_nls_amen.png", width = 10, height = 8, scale = .75)
+ggsave("estimation_nls_amen.png", width = 10, height = 8, scale = .65)
 
 ggplot(filter(results, type == 1),
   aes(eta, error, color = algorithm, linetype = algorithm)) +
@@ -72,7 +72,7 @@ ggplot(filter(results, type == 1),
   theme_bw() +
[…]

### 2017-09-06 hanna/paper.tex
Hanna's edits.

\documentclass[11pt]{article}
\usepackage[pdftex]{graphics}
\usepackage{amsmath,amssymb,rotating,multirow}
\usepackage{times}
\usepackage{url}

\usepackage{booktabs}

\usepackage{subfigure}

\usepackage[longnamesfirst]{natbib}

\usepackage[plain]{algorithm}
\usepackage{algpseudocode}

%\usepackage{parskip}
\usepackage{setspace}

\usepackage{geometry}
\geometry{verbose,letterpaper}

\setlength{\parskip}{0pt}

\renewenvironment{quote}{\list{}{\listparindent=1.5em\itemindent=\listparindent\rightmargin=1.65cm\leftmargin=1.65cm}\item\relax}{\endlist}

\newcommand{\comment}[1]{}

\title{Inferring the Effects of Observed Covariates in\\ Latent Space
  Models for Networks}

\author{Zachary Jones, Matthew Denny, Bruce Desmarais, Hanna Wallach}
\date{\today}

\bibliographystyle{chicago}

\begin{document}

\singlespacing
\maketitle

\thispagestyle{empty}

% TODO: Consider rewriting the abstract to talk about type I vs. type
% II error and the fact that only type II error has been addressed.

\begin{abstract}
  \noindent Due to the complex dependencies found in network data,
  scholars draw upon an increasingly sophisticated toolkit for
  statistical inference. The latent space model (LSM) models the edges
  in a network via a combination of the generalized linear model (GLM)
  and a latent spatial embedding of the network's nodes. In many
  applications, researchers have assumed that this embedding can
  control for unmeasured confounding network structure. However, there
  has been little research that considers whether this is indeed the
  case. Via a simulation study, we investigate the LSM's ability to
  control for unmeasured confounding network structure. We find that
  under even moderate confounding, the LSM does not exhibit lower
  estimation error or inference error than the GLM; it does, however,
  exhibit substantially lower prediction error. We therefore conclude
  that the LSM is most appropriate for predictive or exploratory
  analyses.
\end{abstract}

% latent spatial embedding of the network's nodes

% complex dependencies
% dependencies such as reciprocity, transitivity, and homophily

% unmeasured counfounding variables
% unmeasured confounding network structure

% estimation error
% biased estimates of covariate effects

% inference error
% errors in hypothesis testing

% prediction error

% exploratory analyses, inferential analysis, predictive analysis

% observed covariates
% observed covariates of interest

% NOTE: ``confounding variables'' are correlated with both the
% dependent (edge values) and independent variables (covariates).

% NOTE: ``unmeasured confounding variables'' means that the
% confounding variables are omitted from the model.

% NOTE: Controlling for unmeasured confounding variables often means
% measuring them and including them in the model.

\section{Introduction}

% LSM wasn't developed to account for unmeasured confounding network
% structure, so should NOT talk about it in that same sentence. TODO:
% Rewrite to say that although it wasn't developed for this purpose,
% the LSM is often assumed to control for confounding variables.

% Regarding this, Bruce says: ERGM and SAOM were designed for fully
% specifying the network model, including how covariates effect the
% edges, and how edges relate to each other. ERGM and SAOM do not
% involve any latent variables. In other words, when the analyst is
% using ERGM or SAOM (s)he is assuming to know the correct form of the
% model, and measure all of the data needed to estimate the model.

% The ``unmeasured'' part does not apply to ERGM and SAOM. All of
% these models were designed to account for structural predictors of
% tie formation (e.g., you and I are tied because we have a mutual
% friend---also in the network---who introduced us, not because of our
% background covariates). In ERGM and SAOM you need to explicitly
% specify the functional form of these dependencies in order to
% account for them. By specifying them exactly they are measured on
% the network while fitting the model. The downside is that ERGM does
% not account for any dependencies that are not specified by the
% researcher. With the LSM, the big matrix of latent variables is
% supposed to absorb these dependencies, as well as omitted features.

Inferential analyses of network data have grown increasingly
sophisticated in recent years. Scholars are well-versed in the risks
associated with ignoring unmeasured confounding network structure. If
not accounted for, dependencies such as reciprocity, transitivity, and
homophily can lead to biased estimates of covariate effects and to
hypothesis testing errors, in much the same way that omitted-variable
bias can affect conventional regression analyses
\citep{ward07disputes,kinne14dependent,cranmer17critique,hays10spatial}. Researchers
have therefore proposed a number of statistical models to account for
unmeasured confounding network structure, including the exponential
random graph model \citep[ERGM;
  e.g.,][]{lazer10coevolution,cranmer11inferential,desmarais12micro-level},
the stochastic actor-oriented model \citep[SAOM;
  e.g.,][]{berardo10self-organizing,kinne14dependent}, and the latent
space model \citep[LSM;
  e.g.,][]{ward07disputes,ward07persistent,kirkland12multimember}.

The ERGM and the SAOM both account for dependencies using an approach
similar to that used to control for unmeasured confounding variables
in regression analyses. The researcher specifies a set of dependencies
that (s)he hypothesizes to be important to the network. These
dependencies are then explicitly included in a model that
simultaneously represents the observed covariates of interest. In
contrast, the LSM takes a different approach and uses a latent spatial
embedding to control for unmeasured confounding network structure. The
LSM therefore has an advantage over the ERGM and the SAOM in that the
researcher does not need to hypothesize a set of dependencies that may
be important. However, this advantage is contingent upon the LSM's
ability to identify confounding network structure that would otherwise
be attributed to the observed covariates.

Despite the growing popularity of these models, there have been few
studies that investigate their performance at accounting for
unmeasured confounding network structure. In this article, we focus on
the LSM and study its ability to reduce estimation error and inference
error using a latent spatial embedding of network structure.

TODO: Something about type I error.

\section{The Latent Space Model}

What the LSM is and why it was developed.

What the LSM is used for (currently in 1, 1.1, and 2).

Concerns re. its use, type I and II error, type II addressed.

\section{}

\section*{Acknowledgements}

This work was supported in part by National Science Foundation (NSF)
grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any
opinions, findings, and conclusions or recommendations are those of
the authors and do not necessarily reflect those of the sponsor.

\bibliography{references}

\end{document}

### 2017-09-06 manuscript_rearranged.tex
Hanna's edits.

--- 2017-05-07 manuscript.tex
@@
 
-\noindent Due to the complex interdependence found in network data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference.
-The latent space model (LSM) for network data combines a generalized linear model with a latent spatial embedding of the network.
-In numerous applications, researchers assume that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network.
-There has been little to no research that considers the LSM's performance in adjusting for unmeasured structure.
-We investigate the LSM's performance via a simulation study.
-In the presence of an unmeasured covariate that can be modeled using a latent space, estimation and inferential error remain high under even moderate confounding.
-However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases.
-We conclude that the LSM is most appropriate for exploration or prediction.%\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219.
-Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}
-
+\noindent Due to the complex interdependence found in network data,
+political networks scholars draw upon an increasingly sophisticated
+toolkit for statistical inference. The latent space model (LSM) for
+network data combines a generalized linear model with a latent spatial
+embedding of the network. In numerous applications, researchers assume
+that the latent spatial embedding can control for unmeasured
+confounding structure that is related to the values of edges in the
+network. There has been little to no research that considers the LSM's
+performance in adjusting for unmeasured structure. We investigate the
+LSM's performance via a simulation study. In the presence of an
+unmeasured covariate that can be modeled using a latent space,
+estimation and inferential error remain high under even moderate
+confounding. However, the prediction error of the LSM when unmeasured
+network structure is present is substantially lower in most cases. We
+conclude that the LSM is most appropriate for exploration or
+prediction.
+%\footnote{This work was supported by National Science Foundation
+%grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any
+%opinions, findings, and conclusions or recommendations are those of
+%the authors and do not necessarily reflect those of the sponsor.}
 \end{abstract}
@@
 
-Inferential analysis of political network data has grown increasingly sophisticated in recent years.
-Political networks scholars are well versed in the risks associated with ignoring unmodeled network structure.
-Dependencies such as reciprocity, transitivity, and homophily---if not accounted for---can lead to biased estimates and errors in hypothesis testing, much in the way that omitted variable bias can affect results in conventional regression models \citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}.
-A number of statistical modeling frameworks have been proposed to account for confounding structure in network data that cannot be modeled with observed covariates.
-These include the exponential random graph model (ERGM) \citep[e.g., ][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model (LSM) \citep[e.g., ][]{ward2007disputes,ward2007persistent,kirkland2012multimember}, and the stochastic actor oriented model (SAOM) \citep[e.g., ][]{berardo2010ajps,kinne2014}.
+Inferential analysis of political network data has grown increasingly
+sophisticated in recent years. Political networks scholars are well
+versed in the risks associated with ignoring unmodeled network
+structure. Dependencies such as reciprocity, transitivity, and
+homophily---if not accounted for---can lead to biased estimates and
+errors in hypothesis testing, much in the way that omitted variable
+bias can affect results in conventional regression models
+\citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}. A
+number of statistical modeling frameworks have been proposed to
+account for confounding structure in network data that cannot be
+modeled with observed covariates. These include the exponential random
+graph model (ERGM) \citep[e.g.,
+][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model
+(LSM) \citep[e.g.,
+][]{ward2007disputes,ward2007persistent,kirkland2012multimember}, and
+the stochastic actor oriented model (SAOM) \citep[e.g.,
+][]{berardo2010ajps,kinne2014}.
 
-Despite their growing popularity, few studies exist that investigate the performance of these models in adjusting for confounding network structure.
-The approach to adjusting for dependencies in the two other models commonly used for network data---ERGM and SAOM---is quite similar to adjusting for confounding covariates in regression modeling.
-The researcher specifies a set of dependencies that (s)he hypothesizes to be important in the generative model for the network.
-These dependencies are then explicitly included in a model that simultaneously represents the effects of observed covariates \citep{cranmer2011pa}.
-The LSM takes a different approach, which involves the incorporation of latent variables to model network structure.
-The LSM has an advantage over ERGM and SAOM in that researchers need not develop a set of hypothesized dependencies in order to model network structure that is not reflected in observed covariates.
-However, this advantage hinges upon the capacity for the LSM to discover unmodeled structure that could otherwise be attributed to the observed covariates.
-In the current study, we focus on the LSM, examining its performance in reducing estimation and inferential errors regarding the effects of observed covariates via adjustment for confounding network structure.
+The approach to adjusting for dependencies in the ERGM and SAOM is
+quite similar to adjusting for confounding covariates in regression
+modeling. The researcher specifies a set of dependencies that (s)he
+hypothesizes to be important in the generative model for the
+network. These dependencies are then explicitly included in a model
+that simultaneously represents the effects of observed covariates
+\citep{cranmer2011pa}. The LSM takes a different approach, which
+involves the incorporation of latent variables to model network
+structure. The LSM has an advantage over ERGM and SAOM in that
+researchers need not develop a set of hypothesized dependencies in
+order to model network structure that is not reflected in observed
+covariates. However, this advantage hinges upon the capacity for the
+LSM to discover unmodeled structure that could otherwise be attributed
+to the observed covariates.
 
-\subsection{Central Problem}
+Despite their growing popularity, few studies exist that investigate
+the performance of these models in adjusting for confounding network
+structure. In the current study, we focus on the LSM, examining its
+performance in reducing estimation and inferential errors regarding
+the effects of observed covariates via adjustment for confounding
+network structure. There are two reasons that using the LSM may lead
+to increased error---one may result in Type 1 inferential error, and
+the other in Type 2 error. First, if the unobserved (i.e., latent)
+network structure is truly correlated with the observed covariates,
+the unobserved structure that can be correlated with the observed
+variable may be attributed to the observed variable. Under this
+condition the latent space parameters are used to model sources of
+variation that cannot be attributed to the observed covariate, and
+inferences regarding the observed covariates remain subject to omitted
+variable bias, which leads to an inflated Type 1 inferential error
+rate. Second, the latent configurations inferred may result in a
+representation of the network wherein a node's position in the latent
+space is spuriously correlated with the observed covariates. Inferring
+latent variables that are correlated with observed covariates would
+lead to reduced efficiency in estimating observed covariate effects,
+and result in a high Type 2 error rate.
 
-Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models.
-The LSM, introduced by \citet{hoff2002latent}, is used to estimate the effect of covariates in the presence of latent network structure.
-Here the distance function $|z_i - z_j|$ represents latent network structure as homophily with respect to latent variables.
-The distance function is additively combined with a regression on observed dyadic covariates, $x_{ij}$, to form a linear predictor for tie prediction.
-As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution, giving
+Recent work on the LSM has explicitly addressed the potential problem
+of Type 2 error. The AMEN framework for latent variable modeling of
+networks is explicitly designed to model structure in the
+residuals---structure that is leftover after accounting for the
+observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
+latent factors represent patterns in the network that, ``aren't
+explained by the known regressors.''
+\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
+designed such that the multiplicative effects, ``capture higher-order
+dependence patterns that are left over [in the stochastic linear
+  predictor] after accounting for any known covariate information.''
+The approach incorporated in AMEN is effective at avoiding correlation
+between the observed covariates and latent factors, which would result
+in Type 2 error. However, there have been no methodological
+innovations to avoid Type 1 error in latent variable modeling with
+networks. Through a simulation study we find that the LSM, as
+implemented in both the R \citep{R} packages \texttt{latentnet}
+\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
+terms of Type 1 error when the unmeasured network structure confounds
+the relationship between the observed covariate and the network.
+
+% CUT: not sure where to put this
+%Despite the power of the LSM in representing network structure that is
+%not modeled by the observed covariates, and its performance in
+%predictive ties predicting ties, we have reason to doubt the LSM's
+%capacity to adjust for confounding network structure.
+
+\section{The Latent Space Model}
+
+The LSM, introduced by \citet{hoff2002latent}, is used to estimate the
+effect of covariates in the presence of latent network structure. Here
+the distance function $|z_i - z_j|$ represents latent network
+structure as homophily with respect to latent variables.  The distance
+function is additively combined with a regression on observed dyadic
+covariates, $x_{ij}$, to form a linear predictor for tie prediction.
+As with a GLM, a link function, $g^{-1}$, maps this linear predictor
+to the appropriate edge distribution, giving
 
@@
 
-This ``Euclidean'' LSM is the original form proposed by \citet{hoff2002latent}, and, as far as we can tell, the most commonly used specification in the literature.
-However, the LSM has been extended in several important directions.
-The  The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors with a multiplicative metric \citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa, hoff2009multiplicative}.
-Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}.
-Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process.
-\cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models.
-\cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa.
-\citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R \citep{minhas2016inferential, hoff2015dyadic}.
+This ``Euclidean'' LSM is the original form proposed by
+\citet{hoff2002latent}, and, as far as we can tell, the most commonly
+used specification in the literature.
 
-Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure.
-We do not claim that the methodological literature in which the LSM was introduced or extended has presented the LSM as capable of adjusting for unmeasured confounding structure.
-However, as we document below, many researchers who have applied the LSM claim that it has the capacity to adjust for confounders, and use it for this purpose.
-There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and the other in Type 2 error.
-First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates.
-Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects, and result in a high Type 2 error rate.
-Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable.
-Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias, which leads to an inflated Type 1 inferential error rate.
+% CUT: what's the point of this paragraph?
+%However, the LSM has been extended in several important
+%directions. The latent space has been represented as a $k$-dimensional
+%Euclildean space and by latent factors with a multiplicative metric
+%\citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa,
+%  hoff2009multiplicative}. Additional structure has been introduced by
+%adding random effects, which, for example, may involve sender or
+%receiver specific effects for directed networks which capture
+%differential activity rates amongst nodes
+%\citep{hoff2003random}. Within this framework \cite{westveld2011mixed}
+%models dynamic network data by treating the latent space as a
+%stochastic process. \cite{handcock2007model} enable the LSM to model
+%clustering that is not representable as homophily (i.e., stochastic
+%equivalence) by combining the LSM with latent cluster
+%models. \cite{hoff2008modeling} shows that the LSM and latent cluster
+%models are special cases of an ``eigenmodel.'' That is, an
+%eigendecomposition of a symmetric sociomatrix can be used to represent
+%both latent space and cluster models, but not
+%vice-versa. \citet{hoff2015dyadic} consolidates, extends, and
+%implements several of these extensions in the \texttt{amen} package
+%\citep{amen} for \R \citep{minhas2016inferential, hoff2015dyadic}.
 
-Recent work on the LSM has explicitly addressed the potential problem of Type 2 error.
-The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates.
-\citet[p.
-43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp.
-12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error.
-However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks.
-Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounding structure.
-Through a simulation study we find that the LSM, as implemented in both the R \citep{R} packages \texttt{latentnet} \citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.
+The LSM has seen use in a variety of fields in which network data is
+common, particularly the social sciences. The apparent appeal of the
+LSM appears to be driven primarily by the LSM's usefulness in modeling
+transitivity (i.e., clustering) and homophily, which are ubiquitous in
+social networks. In political science the LSM and variations on the
+form developed in \cite{hoff2002latent} have been used to estimate the
+effect of democracy on the probability of a militarized interstate
+dispute \citep{ward2007disputes}, the amount of portfolio investment
+between states \citep{cao2013democracies}, and the effect of
+multimember districts on the probability of collaboration between
+state legislators in the United States
+\citep{kirkland2012multimember}. Variations of the LSM developed for
+networks measured over time have been applied to the study of
+international trade, wherein the effects of various features of
+trading partners are estimated \citep{ward2013gravity}. In ecology the
+LSM has been used to study the sociality of elephants
+\citep{vance2009social} and orcas \citep{fearnbach2014spatial}, birds
+\citep{nomano2015unrelated}, to discover ecological communities
+\citep{fletcher2011social, fletcher2013network}, and to study food
+webs \citep{chiu2011unifying}. In epidemeology it has been used to
+identify clusters of infected persons for later isolation
+\citep{zhang2015cluster} and to study patterns of interaction amongst
+physicians \citep{paul2014results}. In marketing and business research
+it has been used to study inter-group trust \citep{dass2011impact},
+optimal bundling and pricing of goods and brands for retailers
+\citep{dass2012assessing}. It has been used to describe topic-specific
+patterns of interaction in e-mail communication networks
+\citep{krafft2012topic}.  It has been used to estimate the effects of
+education policy interventions on the structure of friendship networks
+among students \citep{sweet2011modeling}. Lastly, in neuroscience it
+has been proposed as a method for modelling fMRI data
+\citep{simpson2013analyzing}.
 
-\section{Applications of the Latent Space Model}
+%Below we review several studies in which researchers have relied on
+%the latent space modeling framework to adjust for unmeasured
+%confounding structure.
 
-The LSM has seen use in a variety of fields in which network data is common, particularly the social sciences.
-The apparent appeal of the LSM appears to be driven primarily by the LSM's usefulness in modeling transitivity (i.e., clustering) and homophily, which are ubiquitous in social networks.
-In political science the LSM and variations on the form developed in \cite{hoff2002latent} have been used to estimate the effect of democracy on the probability of a militarized interstate dispute \citep{ward2007disputes}, the amount of portfolio investment between states \citep{cao2013democracies}, and the effect of multimember districts on the probability of collaboration between state legislators in the United States \citep{kirkland2012multimember}.
-Variations of the LSM developed for networks measured over time have been applied to the study of international trade, wherein the effects of various features of trading partners are estimated \citep{ward2013gravity}.
-In ecology the LSM has been used to study the sociality of elephants \citep{vance2009social} and orcas \citep{fearnbach2014spatial}, birds \citep{nomano2015unrelated}, to discover ecological communities \citep{fletcher2011social, fletcher2013network}, and to study food webs \citep{chiu2011unifying}.
-In epidemeology it has been used to identify clusters of infected persons for later isolation \citep{zhang2015cluster} and to study patterns of interaction amongst physicians \citep{paul2014results}.
-In marketing and business research it has been used to study inter-group trust \citep{dass2011impact}, optimal bundling and pricing of goods and brands for retailers \citep{dass2012assessing}.
-It has been used to describe topic-specific patterns of interaction in e-mail communication networks \citep{krafft2012topic}.
-It has been used to estimate the effects of education policy interventions on the structure of friendship networks among students \citep{sweet2011modeling}.
-Lastly, in neuroscience it has been proposed as a method for modelling fMRI data \citep{simpson2013analyzing}.
-
-% need to add marketing stuff and probably should look again for applications
-% also need to read and fit in the newer ward student stuff
-
-Although the LSM has often been used as an exploratory or predictive model, it has been applied in many cases for explanatory causal modeling---to reduce estimation and/or inferential error with respect to the effects of observed covariates.
-For example \citet{ward2007disputes}, in a high profile example, argue that the the LSM improves inference about the effects of democracy, international trade, and participation in international organizations on the probability of inter-state conflict.
+Although the LSM has often been used as an exploratory or predictive
+model, it has been applied in many cases for explanatory causal
+modeling---to reduce estimation and/or inferential error with respect
+to the effects of observed covariates. We do not claim that the
+methodological literature in which the LSM was introduced or extended
+has presented the LSM as capable of adjusting for unmeasured
+confounding structure. However, as we document below, many researchers
+who have applied the LSM claim that it has the capacity to adjust for
+confounders, and use it for this purpose. For example
+\citet{ward2007disputes}, in a high profile example, argue that the
+the LSM improves inference about the effects of democracy,
+international trade, and participation in international organizations
+on the probability of inter-state conflict.
 
 \begin{quote}
-``The history of international disputes, and consequently the extant data on militarized interstate disputes, is replete with \ldots dependencies.
-We formally incorporate and estimate the extent of these \ldots dependencies in our model of the Kantian peace in order to more precisely determine the effects of the Kantian tripod on international conflict'' \citep[p.
-585]{ward2007disputes}
+``The history of international disputes, and consequently the extant
+  data on militarized interstate disputes, is replete with \ldots
+  dependencies. We formally incorporate and estimate the extent of
+  these \ldots dependencies in our model of the Kantian peace in order
+  to more precisely determine the effects of the Kantian tripod on
+  international conflict'' \citep[p. 585]{ward2007disputes}
 \end{quote}
 
-Likewise, similar claims are made in \citet{dorff2013networks, cao2013democracies, ward2013gravity, kirkland2012multimember}, and \citet{cao2016transnational}.
+Likewise, similar claims are made in \citet{dorff2013networks,
+  cao2013democracies, ward2013gravity, kirkland2012multimember}, and
+\citet{cao2016transnational}.
 
 \begin{quote}
-``For the most part, however, most dyadic research in international relations ignores the essential features of dyads in that they fail to satisfy the assumption of independence or, by construction, have missing data but ignore its effects: both of these bias the results in a fundamental way'' \citep[p.
-2]{dorff2013networks}.
+``For the most part, however, most dyadic research in international
+  relations ignores the essential features of dyads in that they fail
+  to satisfy the assumption of independence or, by construction, have
+  missing data but ignore its effects: both of these bias the results
+  in a fundamental way'' \citep[p. 2]{dorff2013networks}.
 \end{quote}
@@
 \begin{quote}
-``This approach combines a network analysis with a standard-looking regression to permit us to access the importance of our explanatory factors without having them biased by the interdependencies in the network we are studying'' \citep[p.
-15]{cao2013democracies}.
+``This approach combines a network analysis with a standard-looking
+  regression to permit us to access the importance of our explanatory
+  factors without having them biased by the interdependencies in the
+  network we are studying'' \citep[p. 15]{cao2013democracies}.
 \end{quote}
@@
 \begin{quote}
-``The presence of \ldots dependence implies misspecification and a high likelihood of bias in most current applications.
-Building on the latent space framework we model the world trading system without assuming a particular network structure or the sufficiency of particular network statistics'' \citep[p.
-20]{ward2013gravity}.
+``The presence of \ldots dependence implies misspecification and a
+  high likelihood of bias in most current applications. Building on
+  the latent space framework we model the world trading system without
+  assuming a particular network structure or the sufficiency of
+  particular network statistics'' \citep[p. 20]{ward2013gravity}.
 \end{quote}
@@
 \begin{quote}
-``\ldots the latent space model allows for the assessment of distance between two unconnected actors while simultaneously controlling for the interdependence inherent in network data.
-This interdependence in latent space positions allows the model to control for common network effects like reciprocity or transitivity that would ordinarily bias results'' \citep[p.
-336]{kirkland2012multimember}.
+``\ldots the latent space model allows for the assessment of distance
+  between two unconnected actors while simultaneously controlling for
+  the interdependence inherent in network data. This interdependence
+  in latent space positions allows the model to control for common
+  network effects like reciprocity or transitivity that would
+  ordinarily bias results'' \citep[p. 336]{kirkland2012multimember}.
 \end{quote}
@@
 \begin{quote}
-``Ignoring third-order dependence in dyadic data and treating dyads Germany-France, France-Italy, and Germany-Italy as independent observations can cause bias in parameter estimates (Hoff 2005).
-The statistical literature has proposed a series of latent space to control for autocorrelation among dyadic observations (Hoff et al.
-2002; Hoff 2005).
-Countries?
-unobserved characteristics are captured by latent vectors...'' \citep[p.
-17]{cao2016transnational}.
+``Ignoring third-order dependence in dyadic data and treating dyads
+  Germany-France, France-Italy, and Germany-Italy as independent
+  observations can cause bias in parameter estimates (Hoff 2005). The
+  statistical literature has proposed a series of latent space to
+  control for autocorrelation among dyadic observations (Hoff et
+  al. 2002; Hoff 2005). Countries? unobserved characteristics are
+  captured by latent vectors...'' \citep[p. 17]{cao2016transnational}.
 \end{quote}
 
-The above examples are drawn from political science, but the results of our analysis have implications in other fields.
-We see a similar logic for using the LSM articulated in recent work in ecology by \citet[p.
-989]{nomano2015unrelated}.
+The above examples are drawn from political science, but the results
+of our analysis have implications in other fields. We see a similar
+logic for using the LSM articulated in recent work in ecology by
+\citet[p. 989]{nomano2015unrelated}.
 
 \begin{quote}
-``\ldots can create artificially exaggerated synchrony rates regardless of motivation for signaling, which can be modeled with a term known as ``transitivity'' in the social network literature.
-A latent space model (Krivitsky et al.
-2009) was used to examine the propensities for synchrony between helper males and the primary male while accounting for the variability deriving from the transitivity.''
+``\ldots can create artificially exaggerated synchrony rates
+  regardless of motivation for signaling, which can be modeled with a
+  term known as ``transitivity'' in the social network literature. A
+  latent space model (Krivitsky et al. 2009) was used to examine the
+  propensities for synchrony between helper males and the primary male
+  while accounting for the variability deriving from the
+  transitivity.''
 \end{quote}
 
-Though this list of quotations is by no means complete, the last example to which we point comes from the business literature, and uses the bilinear form of the latent space model \citep[p.
-7]{dass2011impact}.
+Though this list of quotations is by no means complete, the last
+example to which we point comes from the business literature, and uses
+the bilinear form of the latent space model
+\citep[p. 7]{dass2011impact}.
 
 \begin{quote}
-``This model controls for higher order team dynamics using the bilinear component $z_i'z_j$ and also allows
-both trustor ($x_{tor,i}$ ) and trusted ($x_{ted,j}$ ) characteristics to be investigated along with the dyadic covariates ($x_{d,i,j}$).''
+``This model controls for higher order team dynamics using the
+  bilinear component $z_i'z_j$ and also allows both trustor
+  ($x_{tor,i}$ ) and trusted ($x_{ted,j}$ ) characteristics to be
+  investigated along with the dyadic covariates ($x_{d,i,j}$).''
 \end{quote}
 
-The latent space framework offers an attractive general purpose approach to adjusting for confounding structure in network models, as, unlike the major alternative framework for network modeling (ERGM), using the latent space model does not require the researcher to specify a set of network dependencies for which to control.
-However,  the latent variables in the LSM are free parameters that will be inferred to explain variation that is not explained by the observed covariates.
-As such, the LSM framework is not designed to encourage latent network structure to displace observed covariate effects when the latent structure includes confounders.
-As we note above, recent work on latent space approaches for network modeling describes exactly this modeling characteristic---that the latent variables are used to explain variation that cannot be explained by observed covariates.
-In a simulation study that follows, we examine whether the LSM can be used to adjust for confounding in a simple and ideal scenario..
+The latent space framework offers an attractive general purpose
+approach to adjusting for confounding structure in network models, as,
+unlike the major alternative framework for network modeling (ERGM),
+using the latent space model does not require the researcher to
+specify a set of network dependencies for which to control.  However,
+the latent variables in the LSM are free parameters that will be
+inferred to explain variation that is not explained by the observed
+covariates. As such, the LSM framework is not designed to encourage
+latent network structure to displace observed covariate effects when
+the latent structure includes confounders. As we note above, recent
+work on latent space approaches for network modeling describes exactly
+this modeling characteristic---that the latent variables are used to
+explain variation that cannot be explained by observed covariates. In
+a simulation study that follows, we examine whether the LSM can be
+used to adjust for confounding in a simple and ideal scenario..
 
-%\section{Bayesian Inference and Priors in the Latent Space Model}
+\section{Simulation Study}
 
-%Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions).
-Invariance to rotation means that the ML estimate is unidentified.
-As such, inference in the LSM is conducted using a Bayesian approach.
-Latent positions and other parameters are sampled using a Metropolis-Hastings algorithm.
-After sampling, latent space positions are rotated to match a common set of positions, via Procrustes transformation \citep{borg2005modern}.
-Procrustes transformation variance that is due to rotation of the latent space.
-\cite{hoff2002latent} propose the use of diffuse independent normal priors for the latent positions and regression parameters.
-They also recognize the fact that isolates' (i.e., nodes with no ties) finite positions are only identified through the prior in the LSM.
-In one application they actually just exclude the isolates.
-Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space (for, e.g., visualization).
-However, if the goal is to conduct explanatory analysis that is representative of the full set of nodes, excluding isolates would amount to selecting on the dependent variable.
-Unlike in many practical applications of Bayesian inference, it is not possible to use an improper ``flat'' prior \citep{tibshirani1989noninformative} with the LSM, as an informative prior is necessary to assure finite latent positions.
-%\begin{quote}
-%We have not discussed in detail the choice of a prior distribution for latent positions in this article.
-Although simple, the diffuse independent normal priors presented in the examples may not accurately represent prior beliefs about the structure of social networks.
-More appropriate might be clustered point processes or mixtures of normals with an unknown number of components.
-Such priors could allow one to incorporate prior information on tendencies for clustering, without specifying cluster membership.
-This would add another level of hierarchy to the analysis, although the resulting model would be more flexible and perhaps more accurately represent any tendencies of populations to form segregating groups.
-%\end{quote}
-%Since it is necessary to use an informative prior with the LSM, we ask whether it is possible to specify the prior in a way that improves inference regarding covariate parameters.
-We propose to calibrate the prior to be on a scale comparable to the linear predictor estimated from a GLM.
-This scaling should discourage the latent space from replicating the linear predictor, encouraging the discovery of structure that remains in the data after identifying the effects of measured covariates.
-
-%Our scaling rule is expressed as the prior variance for the latent positions, denoted $\nu^2$.
-We focus on a one-dimensional latent space $\mathbf{z}$ with a single dyadic covariate $\mathbf{x}$.
-The rule we derive generalizes naturally to more than one dimension, as the prior variance for each dimension could simply be set to $\nu^2/d$ where $d$ is the number of dimensions.
-If we assume that the $\mathbf{z}$ are independent and distributed $\mathcal{N}(0,\sigma^2)$, it is straightforward to derive the distribution of $|z_i - z_j|$.\footnote{The assumption that $\mathbf{z} \sim \mathcal{N}(0,\sigma^2$) is consistent with $\mathbf{z}$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim \mathcal{N}(0,2\sigma^2)$ \citep{devore2012}.
-The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}.
-Following from this result, we know that the variance of  $|z_i - z_j|$ is $\sigma^2(2-4/\pi)$.
-Let $\theta^2$ be the empirical variance of $\mathbf{x}$.
-Then the variance in the linear predictor is $\beta^2\theta^2$.
-Given an estimate $\hat{\beta}$, we can set the variance of the latent position prior to $$\nu^2=\hat{\beta}^2\theta^2/(2-4/\pi).$$  This prior actively penalizes latent distance distributions that match or exceed the variance of the observed linear predictor.
-\subsection{Simulation Study}
-
-We use a simulation study to evaluate the LSM's performance at inferring covariate parameters in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space.
-We are particularly interested in whether the LSM can adjust for confounding when the confounding variable is unmeasured.
-In the simulation design, we study how the level of confounding affects the performance of the LSM relative to the GLM.
-We evaluate performance in terms of bias and inferential error.
-Since the LSM has also been used for predictive model selection, we also evaluate predictive performance  \citep{ward2013gravity, fletcher2011social, fletcher2013network, chiu2011unifying}.
+We use a simulation study to evaluate the LSM's performance at
+inferring covariate parameters in the situation in which we have some
+observed covariates as well as omitted network structure which can be
+represented using a Euclidean latent space. We are particularly
+interested in whether the LSM can adjust for confounding when the
+confounding variable is unmeasured. In the simulation design, we study
+how the level of confounding affects the performance of the LSM
+relative to the GLM. We evaluate performance in terms of bias and
+inferential error. Since the LSM has also been used for predictive
+model selection, we also evaluate predictive performance
+\citep{ward2013gravity, fletcher2011social, fletcher2013network,
+  chiu2011unifying}.
 
@@
 
-For simplicity and computational efficiency we consider a unidimensional latent space.
-We expect our results to generalize to higher dimensions and other variants of the LSM since each variant involves the use of unconstrained latent variables to represent unmeasured network structure.
-To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a three-step process.
-First, we simulate unidimensional positions for each node, drawing from a normal distribution, and calculate the Euclidean distance $\mathbf{d}$ between each pair of positions.
-Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle \mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector and covariance, assuming that $\mathbf{x}$ has a normal distribution given $\mathbf{d}$ (see \cite[pp.
-116--117]{eaton1983} for the conditional normal derivation).
-Third, we simulate $x$ as a normal random variable with the respective conditional means and covariance.
-Finally we standardize $x$ to have zero mean and unit variance.
-To generate the covariance matrix $\Sigma$, which controls the dependence between the omitted network structure and the observed covariate $\mathbf{x}$ we utilize the C-vine method of \cite{lewandowski2009generating}.
-% fill in a basic description of the method here
+For simplicity and computational efficiency we consider a
+unidimensional latent space. We expect our results to generalize to
+higher dimensions and other variants of the LSM since each variant
+involves the use of unconstrained latent variables to represent
+unmeasured network structure. To generate an observed covariate that
+exhibits a controllable degree of colinearity with the latent network
+structure we follow a three-step process. First, we simulate
+unidimensional positions for each node, drawing from a normal
+distribution, and calculate the Euclidean distance $\mathbf{d}$
+between each pair of positions. Second, given a target covariance
+matrix ($\Sigma$) among the covariates and distances, $\langle
+\mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector
+and covariance, assuming that $\mathbf{x}$ has a normal distribution
+given $\mathbf{d}$ (see \cite[pp. 116--117]{eaton1983} for the
+conditional normal derivation). Third, we simulate $x$ as a normal
+random variable with the respective conditional means and
+covariance. Finally we standardize $x$ to have zero mean and unit
+variance. To generate the covariance matrix $\Sigma$, which controls
+the dependence between the omitted network structure and the observed
+covariate $\mathbf{x}$ we utilize the C-vine method of
+\cite{lewandowski2009generating}.
 
-We consider three exponential family distributions from which the adjacency matrix entries are drawn: Gaussian, Bernoilli, and Poisson.
-We also vary the size of the network under study, considering networks with $n = 25, 50,$ and $100$ nodes.
-To vary the degree of confounding attributable to the latent positions, we consider three values of the collinearity parameter $\eta$, $1$, $100$, and $1,000,000$, where $1$ corresponds to a standard uniform distribution of the correlation  (i.e.,  moderate collinearity) between the observed covariate and the latent distances, and $1,000,000$ to independence between the observed covariate and the latent distances.
-See Figure \ref{fig:vine} for the distribution of the absolute value of the correlation generated at each value of $\eta$.
+We consider three exponential family distributions from which the
+adjacency matrix entries are drawn: Gaussian, Bernoilli, and
+Poisson. We also vary the size of the network under study, considering
+networks with $n = 25, 50,$ and $100$ nodes. To vary the degree of
+confounding attributable to the latent positions, we consider three
+values of the collinearity parameter $\eta$, $1$, $100$, and
+$1,000,000$, where $1$ corresponds to a standard uniform distribution
+of the correlation (i.e., moderate collinearity) between the observed
+covariate and the latent distances, and $1,000,000$ to independence
+between the observed covariate and the latent distances. See Figure
+\ref{fig:vine} for the distribution of the absolute value of the
+correlation generated at each value of $\eta$.
 
@@
 
-We consider the LSM with several different priors on the coefficient for the observed covariate and the latent space.
-We set the prior variance of $\beta$ to be either $1$ or $10$ and use a diffuse normal prior on the latent space.
+We consider the LSM with several different priors on the coefficient
+for the observed covariate and the latent space. We set the prior
+variance of $\beta$ to be either $1$ or $10$ and use a diffuse normal
+prior on the latent space.
 
-The LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R} \citep{latentnet}.
-For each iteration of the simulation, we run 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler.
-Every 100th post burn in iteration is saved.
-Convergence in the log probability of the model is assessed using the Geweke diagnostic in the \texttt{coda} package \citep{coda, geweke1991evaluating}.
-If the convergence criterion is satisfied, the simulation continues to the next set of arguments, otherwise the number of iterations is doubled.
-If the convergence criterion is still not satisfied, then the aforementioned step in the simulation is flagged for review.
-However, this was not necessary in any cases.
-At each point in the simulation's parameter space, we execute 1,000 Monte Carlo iterations.
-\footnote{One of the primary difficulties in executing the above simulation design is the computational cost of estimating the LSM.
-We utilize the \texttt{BatchExperiments} \R package to construct and execute our computational experiments on a Torque cluster \cite{bischl2015batchjobs}.}
+The LSM is estimated using the canonical implementation in the
+\texttt{latentnet} package in \texttt{R} \citep{latentnet}. For each
+iteration of the simulation, we run 10,000 burn in iterations,
+followed by 1,000,000 iterations of the sampler. Every 100th post burn
+in iteration is saved. Convergence in the log probability of the model
+is assessed using the Geweke diagnostic in the \texttt{coda} package
+\citep{coda, geweke1991evaluating}. If the convergence criterion is
+satisfied, the simulation continues to the next set of arguments,
+otherwise the number of iterations is doubled. If the convergence
+criterion is still not satisfied, then the aforementioned step in the
+simulation is flagged for review. However, this was not necessary in
+any cases. At each point in the simulation's parameter space, we
+execute 1,000 Monte Carlo iterations. \footnote{One of the primary
+  difficulties in executing the above simulation design is the
+  computational cost of estimating the LSM. We utilize the
+  \texttt{BatchExperiments} \R package to construct and execute our
+  computational experiments on a Torque cluster
+  \cite{bischl2015batchjobs}.}
 
-We compute the MLE estimated by iteratively reweighted least squares via the \texttt{glm} function in \texttt{R} \citep{rcore}.
-For the LSM we use the posterior mode as our point estimate.
-In the cases where edges are Bernoulli and there is omitted network structure, we scale the estimates using the reciprocal of the bias of $\beta$ when $\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 + \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the variance of a standard logistic distribution.
-We do this to adjust for bias in the coefficient estimate that arises due to the lack of a scale coefficient in logistic regression (See the derivation of the bias under an independent but omitted covariate in \cite{mood2010logistic}).
+We compute the MLE estimated by iteratively reweighted least squares
+via the \texttt{glm} function in \texttt{R} \citep{rcore}. For the LSM
+we use the posterior mode as our point estimate. In the cases where
+edges are Bernoulli and there is omitted network structure, we scale
+the estimates using the reciprocal of the bias of $\beta$ when
+$\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 +
+    \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the
+variance of a standard logistic distribution. We do this to adjust for
+bias in the coefficient estimate that arises due to the lack of a
+scale coefficient in logistic regression (See the derivation of the
+bias under an independent but omitted covariate in
+\cite{mood2010logistic}).
 
-For each combination of simulation conditions we evaluate the mean square prediction error of the model on new edges drawn condtional on $\mathbf{x}$ and $\mathbf{d}$, the bias of the estimated coefficient for the measured covariate, and the Type-1 and 2 error rates.
-Inference regarding Credible intervals for $\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution \cite{turkkan1993computation}.
-We consider $\beta$ to be statistically significant at the 0.05 level if the 95\% credible interval excludes 0.
+For each combination of simulation conditions we evaluate the mean
+square prediction error of the model on new edges drawn condtional on
+$\mathbf{x}$ and $\mathbf{d}$, the bias of the estimated coefficient
+for the measured covariate, and the Type-1 and 2 error
+rates. Inference regarding Credible intervals for $\beta$ are defined
+by using the region of highest posterior density which covers $95\%$
+of the marginal posterior distribution
+\cite{turkkan1993computation}. We consider $\beta$ to be statistically
+significant at the 0.05 level if the 95\% credible interval excludes
+0.
 
-We also repeat this simulation study using \texttt{amen} \citep{hoff2015dyadic, minhas2016inferential}.
-We omit the Poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes.
-Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor.
-We take 100,000 MCMC samples after 500 burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.
+We also repeat this simulation study using \texttt{amen}
+\citep{hoff2015dyadic, minhas2016inferential}. We omit the Poisson
+family, and, due to the computational cost of the simulation, we
+consider only networks with 25 nodes or 100 nodes. Rather than
+constructing an omitted variable as described above, we draw from a
+standard normal distribution $\mathbf{u}$, and, because we are
+considering symmetric networks, take the outer product of $\mathbf{u}$
+with itself to construct the latent factor. We take 100,000 MCMC
+samples after 500 burn-in iterations, and we assess convergence using
+the diagnostic of \cite{raftery1992practical}.
 
@@
 
-To evaluate estimation error we compute the bias.
-Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results.
-In Figure \ref{fig:estimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are included via an observed covariate) show substantial bias.
-In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$.
-However, in absolute terms, the amount of bias is large.
-Our results demonstrate that using the LSM will not permit the discovery of and adjustment for the true latent positions when the observed covariates and latent distances are confounded.
-For both the GLM and the LSM, bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases.
-In \ref{fig:estimation_nls} we can see that when there is no omitted network structure, the LSM exhibits bias greater than that of the GLM.
+To evaluate estimation error we compute the bias. Figures
+\ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the
+results. In Figure \ref{fig:estimation_ls} we can see that with a
+uniform distribution over the correlation between the distances and
+the observed covariate, all methods (except the true model wherein the
+distances are included via an observed covariate) show substantial
+bias. In most cases the LSM performs better than the GLM, sometimes by
+as much as $50\%$. However, in absolute terms, the amount of bias is
+large. Our results demonstrate that using the LSM will not permit the
+discovery of and adjustment for the true latent positions when the
+observed covariates and latent distances are confounded.  For both the
+GLM and the LSM, bias decays rapidly as the degree of correlation
+between the distances and the measured covariate decreases. In
+\ref{fig:estimation_nls} we can see that when there is no omitted
+network structure, the LSM exhibits bias greater than that of the GLM.
 

### 2018-02-16 manuscript.tex
adding quotes from sociology

@@
 
+And in sociology \citet[p. 157]{spillane2017elephant},
+
+ \begin{quote} ``the models include latent space positions that implicitly control for structural aspects of networks, such as reciprocity and centrality, allowing for estimation of covariate effects (Hoff, Raftery, and Handcock 2002).''
+ \end{quote}
+
+ and  \citet[p. 107]{berlusconi2017determinants}
+
+ \begin{quote} ``Higher order dependencies can be taken into account by assigning each country a position in a ?social space,?
+based on its (unobserved) characteristics and the presence of other ties.
+Conditional independence between observations can then subsequently be assumed, given the countries?
+latent position within the social space (Hoff et al., 2002).''
+ \end{quote}
+
 Though this list of quotations is by no means complete, the last example to which we point comes from the business literature, and uses the bilinear form of the latent space model \citep[p.

### 2018-02-16 manuscript_rearranged.tex
adding quotes from sociology

@@
 
+And in sociology \citet[p. 157]{spillane2017elephant},
+
+ \begin{quote} ``the models include latent space positions that implicitly control for structural aspects of networks, such as reciprocity and centrality, allowing for estimation of covariate effects (Hoff, Raftery, and Handcock 2002).''
+ \end{quote}
+
+ and  \citet[p. 107]{berlusconi2017determinants}
+
+ \begin{quote} ``Higher order dependencies can be taken into account by assigning each country a position in a ?social space,?
+based on its (unobserved) characteristics and the presence of other ties.
+Conditional independence between observations can then subsequently be assumed, given the countries?
+latent position within the social space (Hoff et al., 2002).''
+ \end{quote}
+
 Though this list of quotations is by no means complete, the last

### 2018-02-16 manuscript_rearranged.tex
revising abstract

@@
 \noindent Due to the complex interdependence found in network data,
-political networks scholars draw upon an increasingly sophisticated
-toolkit for statistical inference. The latent space model (LSM) for
-network data combines a generalized linear model with a latent spatial
-embedding of the network. In numerous applications, researchers assume
-that the latent spatial embedding can control for unmeasured
-confounding structure that is related to the values of edges in the
-network. There has been little to no research that considers the LSM's
-performance in adjusting for unmeasured structure. We investigate the
+networks scholars draw upon an increasingly sophisticated
+toolkit for statistical inference. The Latent space model (LSM) for
+network data combines a conventional regression model with an embedding of actors in a latent space.
+In the LSM, the expected relationship between two actors is dependent on observed covariates and a function of their positions in the latent space.
+In numerous applications, researchers have assumed
+that the latent spatial embedding can control for latent
+confounding structure.
+Latent confounding structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity.
+There has been little research that considers the LSM's
+performance in adjusting for latent confounding structure. We investigate the
 LSM's performance via a simulation study. In the presence of an
-unmeasured covariate that can be modeled using a latent space,
-estimation and inferential error remain high under even moderate
-confounding. However, the prediction error of the LSM when unmeasured
-network structure is present is substantially lower in most cases. We
+unmeasured actor attribute that can be modeled perfectly using a latent space,
+we find that the LSM exhibits unacceptable levels of bias and Type 1 error for inferences on measured covariates.
+However, the prediction error of the LSM when latent confounding structure is present is substantially lower in most cases.
+We
 conclude that the LSM is most appropriate for exploration or
-prediction.
+prediction, unless it can be assumed that all potential confounders have been measured.
 %\footnote{This work was supported by National Science Foundation

### 2018-02-16 manuscript_rearranged.tex
making revisions to the intro

@@
 networks scholars draw upon an increasingly sophisticated
-toolkit for statistical inference. The Latent space model (LSM) for
+toolkit for building models of networks. The Latent space model (LSM) for
 network data combines a conventional regression model with an embedding of actors in a latent space.
@@
 confounding structure.
-Latent confounding structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity.
+latent network structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity.
 There has been little research that considers the LSM's
-performance in adjusting for latent confounding structure. We investigate the
+performance in adjusting for latent network structure when making inferences about covariates. We investigate the
 LSM's performance via a simulation study. In the presence of an
@@
 we find that the LSM exhibits unacceptable levels of bias and Type 1 error for inferences on measured covariates.
-However, the prediction error of the LSM when latent confounding structure is present is substantially lower in most cases.
+However, the prediction error of the LSM when latent network structure is present is substantially lower in most cases.
 We
-conclude that the LSM is most appropriate for exploration or
-prediction, unless it can be assumed that all potential confounders have been measured.
+conclude that the LSM is most appropriate for exploratory or predictive modeling, unless it can be assumed that all potential confounders have been measured.
 %\footnote{This work was supported by National Science Foundation
@@
 
-Inferential analysis of political network data has grown increasingly
-sophisticated in recent years. Political networks scholars are well
-versed in the risks associated with ignoring unmodeled network
-structure. Dependencies such as reciprocity, transitivity, and
-homophily---if not accounted for---can lead to biased estimates and
+Statistical models in which the outcome data (i.e., the dependent variable) consists of network data have proliferated, and grown increasingly
+sophisticated, over the past decade.
+The defining feature of network data that renders conventional statistical models---namely, regression models---inappropriate, is that ties in a network exhibit complex forms of interdependence such as reciprocity, transitivity, and
+homophily. If not accounted for, these complex dependencies can lead to biased estimates and
 errors in hypothesis testing, much in the way that omitted variable
 bias can affect results in conventional regression models
-\citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}. A
+\citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}.  We refer to network dependencies that cannot be modeled with observed covariates as {\em latent network structure}. A
 number of statistical modeling frameworks have been proposed to
-account for confounding structure in network data that cannot be
-modeled with observed covariates. These include the exponential random
-graph model (ERGM) \citep[e.g.,
+account for latent network structure in network data. These include, but are not limited to, the exponential random
+graph model (ERGM) {\bf [UPDATE STRING CITES TO LOOK OUTSIDE POLISCI]} \citep[e.g.,
 ][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model
@@
 
-The approach to adjusting for dependencies in the ERGM and SAOM is
+The approach to adjusting for latent network structure in the ERGM and SAOM is
 quite similar to adjusting for confounding covariates in regression
@@
 \citep{cranmer2011pa}. The LSM takes a different approach, which
-involves the incorporation of latent variables to model network
-structure. The LSM has an advantage over ERGM and SAOM in that
-researchers need not develop a set of hypothesized dependencies in
-order to model network structure that is not reflected in observed
-covariates. However, this advantage hinges upon the capacity for the
-LSM to discover unmodeled structure that could otherwise be attributed
-to the observed covariates.
+involves the incorporation of latent variables to model latent network structure. The LSM has an advantage over ERGM and SAOM in that
+researchers need not develop a precise set of hypothesized dependencies in
+order to model latent network structure. However, this advantage hinges upon the capacity for the
+LSM to differentiate the effects of latent network structure 
+from those of the observed covariates.
 
@@
 to increased error---one may result in Type 1 inferential error, and
-the other in Type 2 error. First, if the unobserved (i.e., latent)
+the other in Type 2 error. First, if the latent
 network structure is truly correlated with the observed covariates,
@@
 representation of the network wherein a node's position in the latent
-space is spuriously correlated with the observed covariates. Inferring
+space is spuriously correlated with the observed covariates (i.e., the latent space crowds out the effects of observed covariates). Inferring
 latent variables that are correlated with observed covariates would
@@
 
-Recent work on the LSM has explicitly addressed the potential problem
-of Type 2 error. The AMEN framework for latent variable modeling of
-networks is explicitly designed to model structure in the
-residuals---structure that is leftover after accounting for the
-observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
-latent factors represent patterns in the network that, ``aren't
-explained by the known regressors.''
-\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
-designed such that the multiplicative effects, ``capture higher-order
-dependence patterns that are left over [in the stochastic linear
-  predictor] after accounting for any known covariate information.''
-The approach incorporated in AMEN is effective at avoiding correlation
-between the observed covariates and latent factors, which would result
-in Type 2 error. However, there have been no methodological
-innovations to avoid Type 1 error in latent variable modeling with
-networks. Through a simulation study we find that the LSM, as
-implemented in both the R \citep{R} packages \texttt{latentnet}
-\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
-terms of Type 1 error when the unmeasured network structure confounds
-the relationship between the observed covariate and the network.
+%HERE, ADD OVERVIEW OF OUR PAPER
 
@@
 
+Recent work on the LSM has explicitly addressed the potential problem
+of Type 2 error. The AMEN framework for latent variable modeling of
+networks is explicitly designed to model structure in the
+residuals---structure that is leftover after accounting for the
+observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
+latent factors represent patterns in the network that, ``aren't
+explained by the known regressors.''
+\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
+designed such that the multiplicative effects, ``capture higher-order
+dependence patterns that are left over [in the stochastic linear
+  predictor] after accounting for any known covariate information.''
+The approach incorporated in AMEN is effective at avoiding correlation
+between the observed covariates and latent factors, which would result
+in Type 2 error. However, there have been no methodological
+innovations to avoid Type 1 error in latent variable modeling with
+networks. Through a simulation study we find that the LSM, as
+implemented in both the R \citep{R} packages \texttt{latentnet}
+\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
+terms of Type 1 error when the unmeasured network structure confounds
+the relationship between the observed covariate and the network.
+
 We also repeat this simulation study using \texttt{amen}

### 2018-03-12 manuscript_rearranged.tex
updates

--- 2018-02-16 manuscript_rearranged.tex
@@
 
-And in sociology \citet[p. 157]{spillane2017elephant},
-
- \begin{quote} ``the models include latent space positions that implicitly control for structural aspects of networks, such as reciprocity and centrality, allowing for estimation of covariate effects (Hoff, Raftery, and Handcock 2002).''
- \end{quote}
-
- and  \citet[p. 107]{berlusconi2017determinants}
-
- \begin{quote} ``Higher order dependencies can be taken into account by assigning each country a position in a ?social space,?
-based on its (unobserved) characteristics and the presence of other ties.
-Conditional independence between observations can then subsequently be assumed, given the countries?
-latent position within the social space (Hoff et al., 2002).''
- \end{quote}
-
 Though this list of quotations is by no means complete, the last
@@
 
-The latent space framework offers an attractive general purpose
-approach to adjusting for confounding structure in network models, as,
-unlike the major alternative framework for network modeling (ERGM),
-using the latent space model does not require the researcher to
-specify a set of network dependencies for which to control.  However,
-the latent variables in the LSM are free parameters that will be
-inferred to explain variation that is not explained by the observed
-covariates. As such, the LSM framework is not designed to encourage
-latent network structure to displace observed covariate effects when
-the latent structure includes confounders. As we note above, recent
-work on latent space approaches for network modeling describes exactly
-this modeling characteristic---that the latent variables are used to
-explain variation that cannot be explained by observed covariates. In
-a simulation study that follows, we examine whether the LSM can be
-used to adjust for confounding in a simple and ideal scenario..
+The latent space framework offers an attractive general purpose approach to adjusting for confounding structure in network models, as, unlike the major alternative framework for network modeling (ERGM), using the latent space model does not require the researcher to specify a set of network dependencies for which to control.
+However, the latent variables in the LSM are free parameters that will be inferred to explain variation that is not explained by the observed covariates.
+As such, the LSM framework is not designed to encourage latent network structure to displace observed covariate effects when the latent structure includes confounders.
+As we note above, recent work on latent space approaches for network modeling describes exactly this modeling characteristic---that the latent variables are used to explain variation that cannot be explained by observed covariates.
+In a simulation study that follows, we examine whether the LSM can be used to adjust for confounding in a simple and ideal scenario..
 
@@
 
-For simplicity and computational efficiency we consider a
-unidimensional latent space. We expect our results to generalize to
-higher dimensions and other variants of the LSM since each variant
-involves the use of unconstrained latent variables to represent
-unmeasured network structure. To generate an observed covariate that
-exhibits a controllable degree of colinearity with the latent network
-structure we follow a three-step process. First, we simulate
-unidimensional positions for each node, drawing from a normal
-distribution, and calculate the Euclidean distance $\mathbf{d}$
-between each pair of positions. Second, given a target covariance
-matrix ($\Sigma$) among the covariates and distances, $\langle
-\mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector
-and covariance, assuming that $\mathbf{x}$ has a normal distribution
-given $\mathbf{d}$ (see \cite[pp. 116--117]{eaton1983} for the
-conditional normal derivation). Third, we simulate $x$ as a normal
-random variable with the respective conditional means and
-covariance. Finally we standardize $x$ to have zero mean and unit
-variance. To generate the covariance matrix $\Sigma$, which controls
-the dependence between the omitted network structure and the observed
-covariate $\mathbf{x}$ we utilize the C-vine method of
-\cite{lewandowski2009generating}.
++For simplicity and computational efficiency we consider a unidimensional latent space, that is, we create a single unmeasured variable that can be represented using the Euclidean parameterization of the LSM.
+We expect our results to generalize to higher dimensions and other variants of the LSM since each variant involves the use of unconstrained latent variables to represent unmeasured network structure.
+To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a four-step process.
+First, we simulate unidimensional positions for each node, drawing from a standard normal distribution, and calculate the Euclidean distance $\mathbf{d}$ between each pair of positions.
+Second, we generate a target covariance matrix ($\Sigma$) using the $c$-vine method of \cite{lewandowski2009generating} which generates a uniform draw from the space of positive-definite matrices.
+The $c$-vine method makes use of a parameter $\eta$, which when equal to $1$ generates draws from uniformly from the space of positive-definite matrices, making the correlation between the latent positions and the measured covariate 0 in expectation.
+As $\eta$ increases above $1$ draws of $\Sigma$ have increasingly weak correlation\footnote{We show the distribution of correlations under each value of $\eta$ used in our simulation in Figure \ref{fig:vine}.}.
+Third, given the target covariance matrix ($\Sigma$) among the covariates and distances, $\langle \mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector and covariance, assuming that $\mathbf{x}$ has a normal distribution given $\mathbf{d}$ (see \cite[pp.
+116--117]{eaton1983} for the conditional normal derivation).
++Fourth, we simulate $x$ as a normal random variable with the respective conditional means and covariance.
+Finally we standardize $x$ to have zero mean and unit variance to ensure that the scale of estimation errors is comparable across simulations.
 
-We consider three exponential family distributions from which the
-adjacency matrix entries are drawn: Gaussian, Bernoilli, and
-Poisson. We also vary the size of the network under study, considering
-networks with $n = 25, 50,$ and $100$ nodes. To vary the degree of
-confounding attributable to the latent positions, we consider three
-values of the collinearity parameter $\eta$, $1$, $100$, and
-$1,000,000$, where $1$ corresponds to a standard uniform distribution
-of the correlation (i.e., moderate collinearity) between the observed
-covariate and the latent distances, and $1,000,000$ to independence
-between the observed covariate and the latent distances. See Figure
-\ref{fig:vine} for the distribution of the absolute value of the
-correlation generated at each value of $\eta$.
+We consider three exponential family distributions from which the adjacency matrix entries are drawn: Gaussian, Bernoilli, and Poisson.
+We also vary the size of the network under study, considering networks with $n = 25, 50,$ and $100$ nodes.
+To vary the degree of confounding attributable to the latent positions, we consider three values of the collinearity parameter $\eta$: $1$, $100$, and $1,000,000$.
+Here  where $\eta = 1$ gives a matrix where the correlation is in expectation $.5$, $\eta = 100$ results in moderate collinearity between the observed covariate and the latent distances, and $\eta = 1,000,000$ corresponds to independence between the observed covariate and the latent distances.
+See Figure \ref{fig:vine} for the distribution of the absolute value of the correlation generated at each value of $\eta$.
 
@@
 square prediction error of the model on new edges drawn condtional on
-$\mathbf{x}$ and $\mathbf{d}$, the bias of the estimated coefficient
-for the measured covariate, and the Type-1 and 2 error
-rates. Inference regarding Credible intervals for $\beta$ are defined
-by using the region of highest posterior density which covers $95\%$
-of the marginal posterior distribution
-\cite{turkkan1993computation}. We consider $\beta$ to be statistically
-significant at the 0.05 level if the 95\% credible interval excludes
-0.
+$\mathbf{x}$ and $\mathbf{d}$: $\frac{1}{n}\sum_{i=1}^{n}(y_{i} - \hat{y}_i)^2$, the bias of the estimated coefficient
+for the measured covariate $\frac{1}{n} \sum_{i=1}^n (\beta - \hat{beta})$, and the Type-1 and 2 error
+rates.
+These error rates are estimated by computing coverage: $\frac{1}{m} \sum_{i=1}^m\mathbb{I}(\beta \in \text{HPD}(\hat{\beta}_i))$, where $\text{HPD}(\hat{beta})$ is the region of highest posterior density which covers $95\%$ of the marginal posterior distribution and $m$ is the number of samples from said posterior \cite{turkkan1993computation}.
+When in the simulation $\beta = 0$, $1 - P(\text{coverage})$ gives the type-1 error rate, and $P(\text{coverage})$ givesthe type-2 error rate when $\beta = 1$.
+Hence we consider $\beta$ to be statistically
+significant at the 0.05 level if the 95\% credible interval excludes 0.
 

### 2018-03-12 manuscript_rearranged.tex
revising simulation description

@@
 confounders, and use it for this purpose. For example
-\citet{ward2007disputes}, in a high profile example, argue that the
+\citet{ward2007disputes} argue that the
 the LSM improves inference about the effects of democracy,
@@
 
-Likewise, similar claims are made in \citet{dorff2013networks,
-  cao2013democracies, ward2013gravity, kirkland2012multimember}, and
-\citet{cao2016transnational}.
+This example is representative of justifications for using the LSM that we see in other work in political science and international relations.
 
@@
 
-The above examples are drawn from political science, but the results
-of our analysis have implications in other fields. We see a similar
-logic for using the LSM articulated in recent work in ecology by
-\citet[p. 989]{nomano2015unrelated}.
+\noindent We also see examples in sociology,
+
+ \begin{quote} ``the models include latent space positions that implicitly control for structural aspects of networks, such as reciprocity and centrality, allowing for estimation of covariate effects (Hoff, Raftery, and Handcock 2002).'' \citep[p.
+157]{spillane2017elephant}
+ \end{quote}
+
+ \begin{quote} ``Higher order dependencies can be taken into account by assigning each country a position in a ?social space,?
+based on its (unobserved) characteristics and the presence of other ties.
+Conditional independence between observations can then subsequently be assumed, given the countries?
+latent position within the social space (Hoff et al., 2002).'' \citep[p.
+107]{berlusconi2017determinants}
+ \end{quote}
+
+\noindent in ecology,
 
@@
   while accounting for the variability deriving from the
-  transitivity.''
+  transitivity.'' \citep[p. 989]{nomano2015unrelated}
 \end{quote}
-
-And in sociology \citet[p. 157]{spillane2017elephant},
-
- \begin{quote} ``the models include latent space positions that implicitly control for structural aspects of networks, such as reciprocity and centrality, allowing for estimation of covariate effects (Hoff, Raftery, and Handcock 2002).''
- \end{quote}
-
- and  \citet[p. 107]{berlusconi2017determinants}
-
- \begin{quote} ``Higher order dependencies can be taken into account by assigning each country a position in a ?social space,?
-based on its (unobserved) characteristics and the presence of other ties.
-Conditional independence between observations can then subsequently be assumed, given the countries?
-latent position within the social space (Hoff et al., 2002).''
- \end{quote}
-
-Though this list of quotations is by no means complete, the last
-example to which we point comes from the business literature, and uses
-the bilinear form of the latent space model
-\citep[p. 7]{dass2011impact}.
+\noindent and in the business literature.
 
@@
   ($x_{tor,i}$ ) and trusted ($x_{ted,j}$ ) characteristics to be
-  investigated along with the dyadic covariates ($x_{d,i,j}$).''
+  investigated along with the dyadic covariates ($x_{d,i,j}$).'' \citep[p. 7]{dass2011impact}
 \end{quote}
@@
 However, the latent variables in the LSM are free parameters that will be inferred to explain variation that is not explained by the observed covariates.
-As such, the LSM framework is not designed to encourage latent network structure to displace observed covariate effects when the latent structure includes confounders.
-As we note above, recent work on latent space approaches for network modeling describes exactly this modeling characteristic---that the latent variables are used to explain variation that cannot be explained by observed covariates.
 In a simulation study that follows, we examine whether the LSM can be used to adjust for confounding in a simple and ideal scenario..
@@
 how the level of confounding affects the performance of the LSM
-relative to the GLM. We evaluate performance in terms of bias and
-inferential error. Since the LSM has also been used for predictive
-model selection, we also evaluate predictive performance
-\citep{ward2013gravity, fletcher2011social, fletcher2013network,
+relative to the generalized linear model (GLM). We evaluate performance in terms of bias and
+inferential error. Since the LSM has also been used for prediction,  we also evaluate predictive performance
+\citep[e.g., ]{ward2013gravity, fletcher2011social, fletcher2013network,
   chiu2011unifying}.
@@
 
-+For simplicity and computational efficiency we consider a unidimensional latent space, that is, we create a single unmeasured variable that can be represented using the Euclidean parameterization of the LSM.
-We expect our results to generalize to higher dimensions and other variants of the LSM since each variant involves the use of unconstrained latent variables to represent unmeasured network structure.
-To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a four-step process.
-First, we simulate unidimensional positions for each node, drawing from a standard normal distribution, and calculate the Euclidean distance $\mathbf{d}$ between each pair of positions.
+For simplicity and computational efficiency we consider a one-dimensional latent space.
+That is, we create a single unmeasured variable that can be represented using the Euclidean distance between two nodes in a one-dimensional space.
+We expect our results to generalize to higher dimensions and other variants of the LSM since the tendency for latent variables to be used in explaining structure that cannot be attributed covariates is universal with respect to both the dimension of the latent space and the form of the LSM.
+[HERE] To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a four-step process.
+First, we simulate one-dimensional positions for each node, drawing from a standard normal distribution, and calculate the Euclidean distance (i.e., absolute distance) $\mathbf{d}$ between each pair of positions.
 Second, we generate a target covariance matrix ($\Sigma$) using the $c$-vine method of \cite{lewandowski2009generating} which generates a uniform draw from the space of positive-definite matrices.

### 2018-03-12 manuscript_rearranged.tex
revising simulation

@@
 For simplicity and computational efficiency we consider a one-dimensional latent space.
-That is, we create a single unmeasured variable that can be represented using the Euclidean distance between two nodes in a one-dimensional space.
+That is, we create a variable that can be represented using the Euclidean distance between two nodes in a one-dimensional space.
 We expect our results to generalize to higher dimensions and other variants of the LSM since the tendency for latent variables to be used in explaining structure that cannot be attributed covariates is universal with respect to both the dimension of the latent space and the form of the LSM.
-[HERE] To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a four-step process.
+To generate an observed covariate that exhibits a controllable degree of correlation with the latent network structure we follow a four-step process.
 First, we simulate one-dimensional positions for each node, drawing from a standard normal distribution, and calculate the Euclidean distance (i.e., absolute distance) $\mathbf{d}$ between each pair of positions.
-Second, we generate a target covariance matrix ($\Sigma$) using the $c$-vine method of \cite{lewandowski2009generating} which generates a uniform draw from the space of positive-definite matrices.
+Second, we generate a target covariance matrix ($\Sigma$) using the $c$-vine method of \cite{lewandowski2009generating} [UNPACK FROM https://www.rdocumentation.org/packages/clusterGeneration/versions/1.3.4/topics/genPositiveDefMat] which generates a uniform draw from the space of positive-definite matrices.
 The $c$-vine method makes use of a parameter $\eta$, which when equal to $1$ generates draws from uniformly from the space of positive-definite matrices, making the correlation between the latent positions and the measured covariate 0 in expectation.

### 2018-04-23 manuscript_rearranged.tex
clarified explanation of covariate generation

@@
 \title{\vspace{-2cm} Inference on the Effects of Observed Features
-\\ in Latent Space Models for Networks }
+\\ in Latent Variable Models for Networks }
 
@@
 To generate an observed covariate that exhibits a controllable degree of correlation with the latent network structure we follow a four-step process.
-First, we simulate one-dimensional positions for each node, drawing from a standard normal distribution, and calculate the Euclidean distance (i.e., absolute distance) $\mathbf{d}$ between each pair of positions.
-Second, we generate a target covariance matrix ($\Sigma$) using the $c$-vine method of \cite{lewandowski2009generating} [UNPACK FROM https://www.rdocumentation.org/packages/clusterGeneration/versions/1.3.4/topics/genPositiveDefMat] which generates a uniform draw from the space of positive-definite matrices.
-The $c$-vine method makes use of a parameter $\eta$, which when equal to $1$ generates draws from uniformly from the space of positive-definite matrices, making the correlation between the latent positions and the measured covariate 0 in expectation.
-As $\eta$ increases above $1$ draws of $\Sigma$ have increasingly weak correlation\footnote{We show the distribution of correlations under each value of $\eta$ used in our simulation in Figure \ref{fig:vine}.}.
-Third, given the target covariance matrix ($\Sigma$) among the covariates and distances, $\langle \mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector and covariance, assuming that $\mathbf{x}$ has a normal distribution given $\mathbf{d}$ (see \cite[pp.
-116--117]{eaton1983} for the conditional normal derivation).
-+Fourth, we simulate $x$ as a normal random variable with the respective conditional means and covariance.
-Finally we standardize $x$ to have zero mean and unit variance to ensure that the scale of estimation errors is comparable across simulations.
+The steps are as follows.
+\begin{enumerate}
+\item  Simulate one-dimensional positions for each node (i.e., a scalar value for each node), drawing from a standard normal distribution, and calculate the Euclidean distance (i.e., absolute distance) $\mathbf{d}$ between each pair of positions.
+We then normalize $\mathbf{d}$ to assure it has unit variance by $\mathbf{d} := \mathbf{d}/ \text{sd}(\mathbf{d}),$ where sd($\cdot$) is the empirical standard deviation.
+
+ \item Generate a correlation coefficient $(\rho)$ on the interval $(-1,1)$ as $\rho = 1-2\times u$, where $u \sim  \text{Beta}(\eta,\eta)$.
+$\eta$ is used to control the magnitude of the correlation coefficient.
+The smaller the value of $\eta$, the larger the absolute magnitude of the correlation coefficient, in expectation.
+This relationship is illustrated in Figure \ref{fig:vine}.
+When $\eta = 1$, the correlations generated exhibit the expected uniform distribution.
+When $\eta = 1,000,000$, each correlation coefficient is approximately 0.
+\footnote{This methodology generalizes to higher dimensional correlation matrices as the `c-vine' method developed by \cite{lewandowski2009generating}, and implemented in the \R package \texttt{clusterGeneration} \citep{clusterGeneration}, which we use in our simulations.}
+
+ \item Simulate $\mathbf{x} = \frac{\rho}{\sqrt{1-\rho^2}}\mathbf{d}+\mathbf{w},$ where $\mathbf{w} \sim N(0,1)$.
+This assures that the correlation between $\mathbf{x}$ and $\mathbf{d}$ is $\rho$.\footnote{This could be generalized to more than one covariate using the conditional normal distribution derived by \cite[pp.
+116--117]{eaton1983}.}
+
+ \item Finally we standardize $\mathbf{x} := \mathbf{x}/ \text{sd}(\mathbf{x})$ to have unit variance to ensure that the variance of $\mathbf{x}$ relative to $\mathbf{d}$ is unrelated to $\rho$.
+This is to assure that $\rho$ does not affect the efficiency with which the effect of $\mathbf{x}$ can be estimated through some property unrelated to the relationship between $\mathbf{x}$ and $\mathbf{d}$.
+
+\end{enumerate}
 
@@
 \includegraphics[width=\textwidth]{figures/max_r_vine.png}
-\caption{The distribution of the absolute value of the correlation between the observed covariate and the latent distances. \label{fig:vine}}
+\caption{The distribution of the absolute value of the correlation between the observed covariate and the latent distances.
+The value in the plot title corresponds to the value of $\eta$.
+\label{fig:vine}}
 \end{figure}

### 2018-04-23 src/cvine_simplification_test.R

@@ -1,0 +1,47 @@
+# this script tests whether a single draw from a beta distribution
+# is equivalent to the c-vine method for generating correlations in 2 dimensions
+require(clusterGeneration)
+
+# dimension is always 2
+dim = 2
+
+# change eta to re-run test
+eta = 2
+
+# required for replicating genPositiveDefMat()
+rangeVar = c(1, 10)
+
+# assure same seed for both routines
+set.seed(123)
+
+## begin unpacking of 'c-vine' based generation with dim = 2 ##
+
+# generate correlation from beta
+rho <- 2 * rbeta(1, eta, eta) - 1
+
+# reshape to a correlation matrix
+rr <- matrix(c(1, rho, rho, 1), 2, 2)
+
+# create random matrix by which to scale the covariance 
+sigma2 <- runif(dim, min = rangeVar[1], max = rangeVar[2])
+
+# extract diagonal, to be used as standard deviations
+dd <- diag(sqrt(sigma2))
+
+# scale correlation to get covariance
+Sigma <- dd %*% rr %*% dd
+
+# convert back to correlation matrix
+cor.unpacked <- cov2cor(Sigma)
+
+## replicating with clusterGeneration ##
+set.seed(123)
+Sigma.gpdm <- genPositiveDefMat(d=dim,"c-vine",eta=eta)$Sigma
+cor.gpdm <- cov2cor(Sigma.gpdm)
+
+## test equivalence ##
+all((c(round(cor.unpacked,10))==c(round(rr,10))) & (c(round(cor.gpdm,10))==c(round(rr,10))))
+
+
+
+

### 2018-04-23 manuscript_rearranged.tex
revising description

@@
 
+Recent work on the LSM has explicitly addressed the potential problem
+of Type 2 error. The AMEN framework for latent variable modeling of
+networks is explicitly designed to model structure in the
+residuals---structure that is leftover after accounting for the
+observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
+latent factors represent patterns in the network that, ``aren't
+explained by the known regressors.''
+\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
+designed such that the multiplicative effects, ``capture higher-order
+dependence patterns that are left over [in the stochastic linear
+  predictor] after accounting for any known covariate information.''
+The approach incorporated in AMEN is effective at avoiding correlation
+between the observed covariates and latent factors, which would result
+in Type 2 error. However, there have been no methodological
+innovations to avoid Type 1 error in latent variable modeling with
+networks. Through a simulation study we find that the LSM, as
+implemented in both the R \citep{R} packages \texttt{latentnet}
+\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
+terms of Type 1 error when the unmeasured network structure confounds
+the relationship between the observed covariate and the network.
+
 \section{Simulation Study}
@@
 
- \item Simulate $\mathbf{x} = \frac{\rho}{\sqrt{1-\rho^2}}\mathbf{d}+\mathbf{w},$ where $\mathbf{w} \sim N(0,1)$.
+ \item Simulate the observed covariate $\mathbf{x} = \frac{\rho}{\sqrt{1-\rho^2}}\mathbf{d}+\mathbf{w},$ where $\mathbf{w} \sim N(0,1)$.
 This assures that the correlation between $\mathbf{x}$ and $\mathbf{d}$ is $\rho$.\footnote{This could be generalized to more than one covariate using the conditional normal distribution derived by \cite[pp.
@@
 
- \item Finally we standardize $\mathbf{x} := \mathbf{x}/ \text{sd}(\mathbf{x})$ to have unit variance to ensure that the variance of $\mathbf{x}$ relative to $\mathbf{d}$ is unrelated to $\rho$.
+ \item Finally, we standardize $\mathbf{x} := \mathbf{x}/ \text{sd}(\mathbf{x})$ to have unit variance to ensure that the variance of $\mathbf{x}$ relative to $\mathbf{d}$ is unrelated to $\rho$.
 This is to assure that $\rho$ does not affect the efficiency with which the effect of $\mathbf{x}$ can be estimated through some property unrelated to the relationship between $\mathbf{x}$ and $\mathbf{d}$.
@@
 We consider the LSM with several different priors on the coefficient
-for the observed covariate and the latent space. We set the prior
+for the observed covariate ($\mathbf{x}$) and the latent space. We set the prior
 variance of $\beta$ to be either $1$ or $10$ and use a diffuse normal
@@
 \texttt{latentnet} package in \texttt{R} \citep{latentnet}. For each
-iteration of the simulation, we run 10,000 burn in iterations,
+iteration of the simulation, we run 10,000 burn in iterations of the Metropolis-Hastings sampler in \texttt{latentnet},
 followed by 1,000,000 iterations of the sampler. Every 100th post burn
@@
 otherwise the number of iterations is doubled. If the convergence
-criterion is still not satisfied, then the aforementioned step in the
-simulation is flagged for review. However, this was not necessary in
-any cases. At each point in the simulation's parameter space, we
+criterion were still not satisfied, then the aforementioned step would be flagged for review. However, this was not necessary in
+any cases---all converged within 3,000,000 iterations. At each point in the simulation's parameter space, we
 execute 1,000 Monte Carlo iterations. \footnote{One of the primary
@@
 scale coefficient in logistic regression (See the derivation of the
-bias under an independent but omitted covariate in
+bias under an omitted orthogonal covariate in
 \cite{mood2010logistic}).
@@
 significant at the 0.05 level if the 95\% credible interval excludes 0.
-
-Recent work on the LSM has explicitly addressed the potential problem
-of Type 2 error. The AMEN framework for latent variable modeling of
-networks is explicitly designed to model structure in the
-residuals---structure that is leftover after accounting for the
-observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
-latent factors represent patterns in the network that, ``aren't
-explained by the known regressors.''
-\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
-designed such that the multiplicative effects, ``capture higher-order
-dependence patterns that are left over [in the stochastic linear
-  predictor] after accounting for any known covariate information.''
-The approach incorporated in AMEN is effective at avoiding correlation
-between the observed covariates and latent factors, which would result
-in Type 2 error. However, there have been no methodological
-innovations to avoid Type 1 error in latent variable modeling with
-networks. Through a simulation study we find that the LSM, as
-implemented in both the R \citep{R} packages \texttt{latentnet}
-\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
-terms of Type 1 error when the unmeasured network structure confounds
-the relationship between the observed covariate and the network.
-
-We also repeat this simulation study using \texttt{amen}
-\citep{hoff2015dyadic, minhas2016inferential}. We omit the Poisson
-family, and, due to the computational cost of the simulation, we
-consider only networks with 25 nodes or 100 nodes. Rather than
-constructing an omitted variable as described above, we draw from a
-standard normal distribution $\mathbf{u}$, and, because we are
-considering symmetric networks, take the outer product of $\mathbf{u}$
-with itself to construct the latent factor. We take 100,000 MCMC
-samples after 500 burn-in iterations, and we assess convergence using
-the diagnostic of \cite{raftery1992practical}.
 
@@
 \ref{fig:estimation_nls} we can see that when there is no omitted
-network structure, the LSM exhibits bias greater than that of the GLM.
+network structure, the LSM exhibits bias greater than that of the GLM.\footnote{ We also repeat this simulation study using \texttt{amen}
+\citep{hoff2015dyadic, minhas2016inferential,amen},  which is a software implementation of a much more recently developed algorithm for latent variable modeling of networks.
+The results of the simulations using \texttt{amen} , which are in agreement with the results with \texttt{latentnet}, are in the appendix.}
 
@@
 Below we replicate the simulation study results figures using the results from our simulation with the recently developed \texttt{amen} package.
+The latent variable is constructed differently in \texttt{amen} than in the LSM.
+Rather than
+taking the Euclidean distance between two nodes' latent variables, in the simulation with \texttt{amen} we use the outer product between two nodes' latent variables (just the scalar product in our case, since the latent variable has one dimension).
+Since inference with \texttt{amen} tends to converge much faster than with the LSM, we take 100,000 MCMC
+samples after 500 burn-in iterations, and we assess convergence using
+the diagnostic of \cite{raftery1992practical}.
 Note that these simulations exclude (1) the Poisson family, which is not available in the \texttt{amen} package, and (2) the 50-node networks, which we left out in the interest of lessening the computational burden of the simulation.

### 2018-04-24 manuscript_rearranged.tex
continuing to revise

@@
 \noindent Due to the complex interdependence found in network data,
-networks scholars draw upon an increasingly sophisticated
+social networks scholars draw upon an increasingly sophisticated
 toolkit for building models of networks. The Latent space model (LSM) for
@@
 confounding structure.
-latent network structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity.
+Latent network structure includes unmeasured actor attributes that can be represented spatially, as well as network dependencies such as reciprocity and transitivity.
+Adjusting for unmeasured confounding structure or variables goes beyond the abilities of conventional latent variable models (e.g., random effects models), which leads us to scrutinize whether this can be accomplished with the LSM.
 There has been little research that considers the LSM's
@@
 We
-conclude that the LSM is most appropriate for exploratory or predictive modeling, unless it can be assumed that all potential confounders have been measured.
+conclude that the LSM is most appropriate for exploratory or predictive modeling, unless it can be assumed---as is common in random effects modeling---that all potential confounders have been measured and modeled through fixed effects.
 %\footnote{This work was supported by National Science Foundation
@@
 Statistical models in which the outcome data (i.e., the dependent variable) consists of network data have proliferated, and grown increasingly
-sophisticated, over the past decade.
+sophisticated, over the past two decades.
 The defining feature of network data that renders conventional statistical models---namely, regression models---inappropriate, is that ties in a network exhibit complex forms of interdependence such as reciprocity, transitivity, and
@@
 bias can affect results in conventional regression models
-\citep{ward2007disputes,kinne2014,cranmerisq,hays2010spatial}.  We refer to network dependencies that cannot be modeled with observed covariates as {\em latent network structure}. A
+\citep{block2017forms,cranmerisq}.  We refer to network dependencies that cannot be modeled with observed covariates as {\em latent network structure}. A
 number of statistical modeling frameworks have been proposed to
 account for latent network structure in network data. These include, but are not limited to, the exponential random
-graph model (ERGM) {\bf [UPDATE STRING CITES TO LOOK OUTSIDE POLISCI]} \citep[e.g.,
-][]{lazer2010,cranmer2011pa,desmarais2012psj}, the latent space model
+graph model (ERGM) \citep[e.g.,
+][]{cranmer2011pa,krivitsky2017exponential}, the latent space model
 (LSM) \citep[e.g.,
-][]{ward2007disputes,ward2007persistent,kirkland2012multimember}, and
+][]{ward2007persistent,kirkland2012multimember,berlusconi2017determinants}, and
 the stochastic actor oriented model (SAOM) \citep[e.g.,
-][]{berardo2010ajps,kinne2014}.
+][]{snijders2001statistical,berardo2010ajps,kinne2014}.
 
@@
 \citep{cranmer2011pa}. The LSM takes a different approach, which
-involves the incorporation of latent variables to model latent network structure. The LSM has an advantage over ERGM and SAOM in that
+involves the incorporation of latent variables to model latent network structure. Unlike with ERGM and SAOM, with the LSM
 researchers need not develop a precise set of hypothesized dependencies in
-order to model latent network structure. However, this advantage hinges upon the capacity for the
-LSM to differentiate the effects of latent network structure 
-from those of the observed covariates.
+order to model latent network structure. However, it is not clear from the literature whether the
+LSM is capable of differentiating the effects of latent network structure (i.e., the random effects)
+from those of the observed covariates (i.e., the fixed effects).
+If the LSM can be thought of as a conventional mixed effects model, in which it must be assumed that the random effects are uncorrelated with the observed covariates, then the LSM would not be able to adjust for confounding network structure (see, e.g., \citep{gardiner2009fixed}).
 
@@
 the unobserved structure that can be correlated with the observed
-variable may be attributed to the observed variable. Under this
+variable may be attributed to the observed variable---classic confounder bias. Under this
 condition the latent space parameters are used to model sources of
@@
 and result in a high Type 2 error rate.
+
+Recent work on the LSM has explicitly addressed the potential problem
+of Type 2 error. The AMEN framework for latent variable modeling of
+networks is explicitly designed to model structure in the
+residuals---structure that is leftover after accounting for the
+observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
+latent factors represent patterns in the network that, ``aren't
+explained by the known regressors.''
+\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
+designed such that the multiplicative effects, ``capture higher-order
+dependence patterns that are left over [in the stochastic linear
+  predictor] after accounting for any known covariate information.''
+The approach incorporated in AMEN is effective at avoiding correlation
+between the observed covariates and latent factors, which would result
+in Type 2 error. However, there have been no methodological
+innovations to avoid Type 1 error in latent variable modeling with
+networks.
+
+The purpose of the current study is to evaluate the performance of the LSM in making inferences regarding the effects of observed covariates in the presence of different forms of latent network structure.
+in the next section we provide a brief overview of the LSM and provide evidence that it has seen widespread use for the purpose of adjusting for latent network structure that could potentiall confound the effects of observed covariates.
+We then present a simulation study in which we evaluate the performance of the LSM.
+We find that the LSM, as
+implemented in the \R \citep{R} package \texttt{latentnet}
+\citep{latentnet} and AMEN, as implemented in the \R package \texttt{amen} \citep{amen}, performs poorly in
+terms of Type 1 error when the unmeasured network structure confounds
+the relationship between the observed covariate and the network.
 
@@
 
-This ``Euclidean'' LSM is the original form proposed by
-\citet{hoff2002latent}, and, as far as we can tell, the most commonly
+This ``Euclidean'' (or latent distance) LSM is one of the the original forms proposed by
+\citet{hoff2002latent}, and is, as far as we can tell, the most commonly
 used specification in the literature.
+Several methodological extensions have been introduced that either build directly upon the LSM or offer alternative specifications of the latent variable structure \citep[e.g., ][]{hoff2009multiplicative,westveld2011mixed,handcock2007model,hoff2008modeling,hoff2015dyadic}.
+The computational intensiveness of working with these models precludes us from evaluating each separate version, but we see no reason why our simulation results would not generalize to the broad class of latent variable models for networks.
 
@@
 
-The LSM has seen use in a variety of fields in which network data is
-common, particularly the social sciences. The apparent appeal of the
-LSM appears to be driven primarily by the LSM's usefulness in modeling
-transitivity (i.e., clustering) and homophily, which are ubiquitous in
-social networks. In political science the LSM and variations on the
-form developed in \cite{hoff2002latent} have been used to estimate the
-effect of democracy on the probability of a militarized interstate
-dispute \citep{ward2007disputes}, the amount of portfolio investment
-between states \citep{cao2013democracies}, and the effect of
-multimember districts on the probability of collaboration between
-state legislators in the United States
-\citep{kirkland2012multimember}. Variations of the LSM developed for
-networks measured over time have been applied to the study of
-international trade, wherein the effects of various features of
-trading partners are estimated \citep{ward2013gravity}. In ecology the
-LSM has been used to study the sociality of elephants
-\citep{vance2009social} and orcas \citep{fearnbach2014spatial}, birds
-\citep{nomano2015unrelated}, to discover ecological communities
-\citep{fletcher2011social, fletcher2013network}, and to study food
-webs \citep{chiu2011unifying}. In epidemeology it has been used to
-identify clusters of infected persons for later isolation
-\citep{zhang2015cluster} and to study patterns of interaction amongst
-physicians \citep{paul2014results}. In marketing and business research
-it has been used to study inter-group trust \citep{dass2011impact},
-optimal bundling and pricing of goods and brands for retailers
-\citep{dass2012assessing}. It has been used to describe topic-specific
-patterns of interaction in e-mail communication networks
-\citep{krafft2012topic}.  It has been used to estimate the effects of
-education policy interventions on the structure of friendship networks
-among students \citep{sweet2011modeling}. Lastly, in neuroscience it
-has been proposed as a method for modelling fMRI data
-\citep{simpson2013analyzing}.
+%The LSM has seen use in a variety of fields in which network data is
+%common, particularly the social sciences. The apparent appeal of the
+%LSM appears to be driven primarily by the LSM's usefulness in modeling
+%transitivity (i.e., clustering) and homophily, which are ubiquitous in
+%social networks. In political science the LSM and variations on the
+%form developed in \cite{hoff2002latent} have been used to estimate the
+%effect of democracy on the probability of a militarized interstate
+%dispute \citep{ward2007disputes}, the amount of portfolio investment
+%between states \citep{cao2013democracies}, and the effect of
+%multimember districts on the probability of collaboration between
+%state legislators in the United States
+%\citep{kirkland2012multimember}. Variations of the LSM developed for
+%networks measured over time have been applied to the study of
+%international trade, wherein the effects of various features of
+%trading partners are estimated \citep{ward2013gravity}. In ecology the
+%LSM has been used to study the sociality of elephants
+%\citep{vance2009social} and orcas \citep{fearnbach2014spatial}, birds
+%\citep{nomano2015unrelated}, to discover ecological communities
+%\citep{fletcher2011social, fletcher2013network}, and to study food
+%webs \citep{chiu2011unifying}. In epidemeology it has been used to
+%identify clusters of infected persons for later isolation
+%\citep{zhang2015cluster} and to study patterns of interaction amongst
+%physicians \citep{paul2014results}. In marketing and business research
+%it has been used to study inter-group trust \citep{dass2011impact},
+%optimal bundling and pricing of goods and brands for retailers
+%\citep{dass2012assessing}. It has been used to describe topic-specific
+%patterns of interaction in e-mail communication networks
+%\citep{krafft2012topic}.  It has been used to estimate the effects of
+%education policy interventions on the structure of friendship networks
+%among students \citep{sweet2011modeling}. Lastly, in neuroscience it
+%has been proposed as a method for modelling fMRI data
+%\citep{simpson2013analyzing}.
 
@@
 
-Although the LSM has often been used as an exploratory or predictive
+Although the LSM has been used as an exploratory or predictive
 model, it has been applied in many cases for explanatory causal
@@
 \citet{ward2007disputes} argue that the
-the LSM improves inference about the effects of democracy,
+the version of the LSM that they use improves inference about the effects of democracy,
 international trade, and participation in international organizations
@@
 
-The latent space framework offers an attractive general purpose approach to adjusting for confounding structure in network models, as, unlike the major alternative framework for network modeling (ERGM), using the latent space model does not require the researcher to specify a set of network dependencies for which to control.
-However, the latent variables in the LSM are free parameters that will be inferred to explain variation that is not explained by the observed covariates.
-In a simulation study that follows, we examine whether the LSM can be used to adjust for confounding in a simple and ideal scenario..
-
-Recent work on the LSM has explicitly addressed the potential problem
-of Type 2 error. The AMEN framework for latent variable modeling of
-networks is explicitly designed to model structure in the
-residuals---structure that is leftover after accounting for the
-observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
-latent factors represent patterns in the network that, ``aren't
-explained by the known regressors.''
-\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
-designed such that the multiplicative effects, ``capture higher-order
-dependence patterns that are left over [in the stochastic linear
-  predictor] after accounting for any known covariate information.''
-The approach incorporated in AMEN is effective at avoiding correlation
-between the observed covariates and latent factors, which would result
-in Type 2 error. However, there have been no methodological
-innovations to avoid Type 1 error in latent variable modeling with
-networks. Through a simulation study we find that the LSM, as
-implemented in both the R \citep{R} packages \texttt{latentnet}
-\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
-terms of Type 1 error when the unmeasured network structure confounds
-the relationship between the observed covariate and the network.
+The latent space framework offers an attractive general purpose approach to adjusting for confounding structure in network models, as, unlike the major alternative frameworks for network modeling (ERGM, SAOM), using the latent space model does not require the researcher to specify a set of network dependencies for which to adjust.
+However, we hypothesize that, as with random effects in mixed effect models, the latent variables in the LSM are free parameters that will be inferred to explain variation that is not explained by the observed covariates.
 
